---
title: "A worked example: structured body-size model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A worked example: structured body-size model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_drmTMB <- requireNamespace("drmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_drmTMB
)
```

The "Get started" vignette walks the surface on a deliberately small fit
(`n = 80`, one continuous predictor, one random intercept). This article
exercises more of the renderer machinery on a model an ecologist would
actually run: a Gaussian location-scale fit with a continuous predictor, a
log-transformed predictor, a two-level factor contrast, and a random
intercept on the mean submodel.

The goal is not to teach you `drmTMB`; it is to show what each
`symbolize()` surface returns when the inputs are richer.

## 1. Setup and fit

We simulate a body-size dataset with `temperature` (continuous),
`food` (continuous, lognormal so `log(food)` is the natural scale), `sex`
(two-level factor), and `site` (eight-level grouping). The fitted model
has the mean depending on all four predictors and the residual standard
deviation depending on `temperature` and `sex`.

```{r fit}
library(symbolizer)
library(drmTMB)

set.seed(20260523)
n           <- 200
temperature <- runif(n, 8, 30)
food        <- rlnorm(n, meanlog = 1.5, sdlog = 0.6)
sex         <- factor(sample(c("female", "male"), n, replace = TRUE))
site        <- factor(sample(letters[1:8], n, replace = TRUE))

site_re <- rnorm(8, sd = 1.5)
mu      <- 5 + 0.5 * temperature + 2 * log(food) +
             3 * (sex == "male") + site_re[site]
sigma   <- exp(0.3 + 0.05 * temperature + 0.1 * (sex == "male"))

dat <- data.frame(
  body_mass   = rnorm(n, mu, sigma),
  temperature = temperature,
  food        = food,
  sex         = sex,
  site        = site
)

fit <- drmTMB(
  drm_formula(
    body_mass ~ temperature + log(food) + sex + (1 | site),
    sigma ~ temperature + sex
  ),
  family = gaussian(),
  data   = dat
)
```

`symbolize()` is the single entry point. We pass biologically meaningful
symbols (`W_i` for body mass, `T_i` for temperature, `F_i` for food, `S_i`
for sex) and units; the rendered equations carry these instead of the raw
R names.

```{r symbolize}
sym <- symbolize(
  fit,
  symbols = c(body_mass = "W_i", temperature = "T_i",
              food      = "F_i", sex         = "S_i"),
  units   = c(body_mass = "g",   temperature = "C",
              food      = "g/day"),
  context = "structured body-size location-scale model"
)
sym
```

**Takeaway.** One call to `symbolize()` produces the structured object;
every later section is a renderer pulling rows from it.

## 2. The mean structure

`equations()` returns one row per renderable block (the conditional
distribution, each linear predictor, and any random-effect distribution).
Both notations always live in their own columns; `notation = "both"`
controls only how the result prints.

```{r equations}
equations(sym, notation = "both")
```

Reading the mu row line by line:

- `\log(F_i)` shows up as the transformed predictor. The bridge
  recognises `log(food)` as a transformation role and renders the
  transformation in math, not as a plain symbol.
- The sex contrast renders as `\beta_{3} \, [sex = \mathrm{male}]`. The
  bracket notation is the contrast convention: the coefficient applies
  whenever `sex = male`, and the female baseline is folded into the
  intercept.
- The random-intercept term `u_{site(i)}` is appended to the mu predictor,
  flagged by its subscript that it depends on which `site` observation
  `i` belongs to.

Spliced into a math-aware renderer (the pkgdown site emits MathML through
pandoc; Quarto and most LaTeX renderers use KaTeX or MathJax), the
two-notation block looks like this:

```{r equations-latex, results = 'asis'}
cat("$$", as_latex(sym, notation = "both"), "$$", sep = "\n")
```

**Takeaway.** The mu line carries four predictors with different roles
(intercept, slope, transformation, factor contrast) plus a random
intercept, all rendered from the same `symbolize()` call.

## 3. The scale structure

The sigma submodel does its own work. Pulling just the `sigma`
linear-predictor row:

```{r sigma-row}
eq <- equations(sym, notation = "both")
eq[eq$submodel == "sigma" & eq$kind == "linear_predictor", ]
```

Two features deserve attention.

First, the sigma submodel has its own predictors: residual variability
changes with `temperature` and with `sex`. This is the location-scale
part — the model lets the spread depend on covariates, not just the
mean.

Second, the link on sigma is `log`, which `drmTMB` locks in. The
biological reading is multiplicative: a unit change in `temperature`
multiplies the residual standard deviation by `\exp(\gamma_{1})`, and the
male contrast multiplies it by `\exp(\gamma_{2})` relative to females.

**Takeaway.** The residual SD is not constant; it varies on the
log-scale with the same kind of linear-predictor grammar as the mean.

## 4. What the symbols mean

`symbol_table()` lists every symbol used in the equations, together with
its abstract dimension (the shape rule) and its concrete dimension (the
actual sizes from this fit). With `n = 200` observations, four mu
coefficients (intercept + `temperature` + `log(food)` + sex contrast),
and three sigma coefficients (intercept + `temperature` + sex contrast),
the concrete columns now show real numbers, not just placeholders.

```{r symbol-table}
symbol_table(sym)
```

Notice the design matrix rows: `\mathbf{X}` is `\mathbb{R}^{n \times
p_\mu}` abstractly, `\mathbb{R}^{200 \times 4}` concretely; `\mathbf{Z}`
is the sigma submodel's design matrix at `\mathbb{R}^{200 \times 3}`.
The random-intercept row carries `\mathbb{R}^{8}` because `site` has
eight levels.

**Takeaway.** The symbol table is the page a reader uses to look up what
each glyph in the equations actually means and how big it is.

## 5. What is assumed

`assumption_table()` separates **stated** assumptions (those baked into
the formula: distribution, link, linear predictor), **implied**
assumptions (those that follow from the parameterisation: positivity of
`\sigma_i`, conditional independence), and **not_checked** assumptions
(those still on the user's plate: missing-at-random).

```{r assumption-table}
assumption_table(sym)
```

Because this fit carries a random intercept, the table swaps in the
RE-conditional independence row automatically: observations are
conditionally independent **given the predictors and the random
effects**. We can verify it directly from `sym$assumptions`:

```{r assumption-re-swap}
sym$assumptions$assumption
```

The plain `independence` row is absent; `independence_given_random_effects`
appears in its place. A fixed-effects-only fit would show the reverse.

**Takeaway.** Stated, implied, not_checked is the audit lens: the model
guarantees the first two; the third is your responsibility.

## 6. Reading each coefficient

`parameter_interpretation()` reads each fixed-effect estimate on every
scale that makes sense for its submodel: a link-scale reading, a
natural-scale reading, a variance-scale reading (sigma only), and a
biological reading.

```{r parameter-interpretation}
parameter_interpretation(sym)
```

For this fit, the most informative comparison is the `temperature` slope
on each submodel.

On mu, `\beta_{1}` is additive on the response: a unit change in
`temperature` shifts the expected body mass by the coefficient value,
because the mu link is identity.

On sigma, `\gamma_{1}` is additive on the log scale, which means a unit
change in `temperature` *multiplies* the residual SD by
`\exp(\gamma_{1})`. The biological reading expresses this directly: a
unit change in temperature multiplies the unexplained variability of
`body_mass` by `\exp(\gamma_{1})`. Two coefficients on the same
predictor, two genuinely different scales.

The `log(food)` row deserves a closer look: a transformed predictor
takes the `transformation` role and gets its own templated readings on
the link, natural, and biological scales. The natural-scale reading
reports the slope per unit of `log(food)`, not per raw unit of `food` —
the mathematics in `equations()` and the English in this table now
agree about which scale the coefficient lives on.

```{r transformation-row}
sym$interpretation[sym$interpretation$term_label == "log(food)", , drop = FALSE]
```

**Takeaway.** v0.1 ships interpretation templates for `intercept`,
`slope`, `factor_contrast`, and `transformation` roles on both mu and
sigma. Interactions remain on the roadmap.

## 7. R syntax to math, both directions

`formula_bridge()` is the educator-facing translation table from R
formula syntax to mathematics. Two rows, both notations.

```{r formula-bridge}
formula_bridge(sym)
```

For the mu submodel, the R syntax
`body_mass ~ temperature + log(food) + sex + (1 | site)` maps to the
indexed equation in the `mathematics` column and to
`\boldsymbol{\mu} = \mathbf{X} \boldsymbol{\beta} + \mathbf{u}` in the
`mathematics_matrix` column. The random-intercept piece appears as
`+ \mathbf{u}` in the matrix form and as `+ u_{site(i)}` in the indexed
form. A reader can compare the two columns and learn the translation
without doing it by hand.

**Takeaway.** R syntax on the left, math on the right, both notations
side by side.

## 8. The notation bridge for educators

`notation_bridge()` is the deeper teaching surface: every symbol *and*
every equation pairs its index form with its matrix form and tags both
shapes.

```{r notation-bridge}
notation_bridge(sym)
```

The `dimension` column carries the shape rule
(`\mathbb{R}^n`, `\mathbb{R}^{n \times p_\mu}`). The
`dimension_concrete` column plugs in the actual numbers from this fit
(`\mathbb{R}^{200}`, `\mathbb{R}^{200 \times 4}`). A reader learning
matrix notation can stare at this table and see, in one place, that the
`\boldsymbol{\mu}` vector has 200 entries, the `\boldsymbol{\beta}`
vector has four, and `\mathbf{X}` is the 200 by 4 matrix that connects
them.

**Takeaway.** Abstract shape rules on one side, concrete numbers from
this fit on the other. That is the bridge.

## 9. Roadmap and capabilities

`symbolizer_capabilities()` is the single source of truth for what
`symbolize()` knows how to render. Each row carries one of five status
words: **Stable**, **First slice**, **Opt-in control**, **Planned or
reserved**, or **Unsupported or blocked**.

```{r capabilities}
symbolizer_capabilities()
```

For the model fit in this vignette, the relevant rows are:

- `drmTMB / gaussian / mu` — **Stable**
- `drmTMB / gaussian / sigma` — **Stable**
- `drmTMB / gaussian / random_effects` — **First slice** (intercept-only
  `(1 | group)`; other RE structures still error)

Everything else in the table is **Planned or reserved**. If
`symbolize()` errors with a capability message, the registry tells you
which version is scheduled to lift the restriction.

**Takeaway.** Read the status word before designing around a tuple. The
registry, not the prose, decides what `symbolize()` accepts.
