---
title: "Get started with symbolizer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with symbolizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_drmTMB <- requireNamespace("drmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_drmTMB
)
```

## Why structured symbolic models

[`equatiomatic`](https://datalorax.github.io/equatiomatic/) renders a fitted
model as a LaTeX string. That is enough when the only goal is a publishable
equation. `symbolizer` returns a richer object: a *structured symbolic model*
that other renderers can read. From the same object you can pull the equation,
the symbol dictionary, the assumption table, the formula-to-math bridge, and
the per-coefficient interpretation on link, natural, variance, and biological
scales.

The package is educator-first. It exposes equations in both index (scalar) and
matrix notation side by side, and it labels every assumption as *stated*,
*implied*, or *not checked* so a reader can see what the fit guarantees and
what it does not.

**Takeaway.** The product is the `symbolized_model` object; everything else is
a renderer of that object.

## A short glossary

If you don't read formula-grammar terminology fluently, here's the
quickest possible map of what each word means in a biology context.

```{r glossary, echo = FALSE}
glossary <- data.frame(
  word = c(
    "response", "predictor", "factor", "submodel", "linear predictor",
    "design matrix", "coefficient", "link function", "random intercept",
    "latent variable", "loading"
  ),
  meaning = c(
    "the outcome you measured (e.g., body mass, abundance).",
    "a variable you think influences the response.",
    "a categorical predictor with named levels (e.g., sex with \"female\" and \"male\").",
    "one piece of the formula. A location-scale model has two submodels: one for the mean (mu), one for the residual SD (sigma).",
    "the sum that determines a parameter for an observation: intercept + slope x predictor + ... Sometimes a link function (e.g., log) is applied to make it work on its natural scale.",
    "the table the computer multiplies coefficients by to get fitted values. Each row is one observation; each column is one term (intercept, slope, factor contrast, ...).",
    "a single number the model estimates: an intercept, a slope, or a factor contrast.",
    "the transformation between a parameter's natural scale (e.g., probability) and the scale the linear predictor works on (e.g., logit). For `sigma` in a location-scale model, the link is log, so the linear predictor models log(sigma).",
    "a per-group offset that lets each group have its own baseline, while assuming those offsets follow a shared distribution.",
    "an unmeasured axis the model uses to explain shared variation in multiple traits or species. (Used in gllvmTMB.)",
    "how strongly a single trait or species responds to a latent axis."
  ),
  stringsAsFactors = FALSE
)
knitr::kable(glossary, col.names = c("word", "meaning"))
```

### How a factor becomes 0/1 columns

When you put a factor like `sex` (with levels `female`, `male`) in a formula,
R automatically encodes the non-reference levels as **dummy variables**: 0/1
columns the model can multiply by coefficients.

| Observation | sex      | sexmale (dummy column) |
| ----------- | -------- | ---------------------- |
| 1           | female   | 0                      |
| 2           | male     | 1                      |
| 3           | female   | 0                      |

For a `k`-level factor, R makes `k-1` dummy columns. The first level
(alphabetical by default) is the **reference**: it doesn't get a column of its
own and lives inside the intercept. `symbol_table()` marks the reference
explicitly -- look for the `[reference]` tag in the description.

The `sexmale` coefficient is the *difference* between male and female means at
the reference values of any other predictors, not the male mean itself.

**Takeaway.** A k-level factor produces k-1 dummy columns. The reference level
lives inside the intercept; each contrast coefficient is a *difference* from
the reference.

**Takeaway.** If a word above is unfamiliar, treat the next section as a
hands-on definition by example.

## A first `symbolize()` call

The canonical v0.1 example is a Gaussian location-scale model fit with
[`drmTMB`](https://itchyshin.github.io/drmTMB/): one submodel for the mean
(`mu`), one for the residual standard deviation (`sigma`).

```{r fit}
library(symbolizer)
library(drmTMB)

set.seed(1)
n <- 80
temperature <- runif(n, 10, 25)
dat <- data.frame(
  body_mass   = rnorm(n, 30 + 0.4 * temperature, exp(0.5 + 0.1 * temperature)),
  temperature = temperature
)

fit <- drmTMB(
  drm_formula(body_mass ~ temperature, sigma ~ temperature),
  family = gaussian(),
  data   = dat
)
```

`symbolize()` is the single entry point. Pass user-facing symbols and units so
the rendered equations carry biological meaning rather than R variable names:

```{r symbolize}
sym <- symbolize(
  fit,
  symbols = c(body_mass = "W_i", temperature = "T_i"),
  units   = c(body_mass = "g",   temperature = "C"),
  context = "avian body-size location-scale model"
)
```

**Takeaway.** One call gives you the structured object; the next sections
unpack what it carries.

## Equations in both notations

`equations()` returns one row per renderable block (conditional distribution,
each linear predictor, and any random-effect distribution). It always carries
both notations as columns; the `notation` argument only controls how the
result prints.

```{r equations}
equations(sym)
```

`as_latex()` produces the LaTeX string ready to splice into a document. With
`notation = "both"` the index and matrix forms are stacked in two `aligned`
blocks. Pasting the output into a math-aware renderer (the pkgdown site
emits MathML through pandoc; Quarto and most LaTeX renderers use KaTeX or
MathJax) shows them as proper math:

```{r as-latex-print}
cat(as_latex(sym, notation = "both"))
```

```{r as-latex-math, results = 'asis'}
cat("$$", as_latex(sym, notation = "both"), "$$", sep = "\n")
```

The matrix form follows a simple convention: bold lowercase is a vector
($\boldsymbol{\mu}$, $\boldsymbol{\beta}$), bold uppercase is a matrix
($\mathbf{X}$, $\mathbf{Z}$), and Greek vectors keep their Greek glyph.

`notation_bridge()` is the educator-facing table that pairs every piece of the
model with its counterpart in the other notation and tags the shape:

```{r notation-bridge}
notation_bridge(sym)
```

Read it as follows: the `dimension` column is the shape rule
(`\mathbb{R}^n`, `\mathbb{R}^{n \times p_\mu}`), and the `dimension_concrete`
column plugs in the actual sizes from this fit (`\mathbb{R}^{80}`,
`\mathbb{R}^{80 \times 2}`).

**Takeaway.** Both notations always coexist; the bridge teaches the reader to
move between them without translating by hand.

## Three views of your model

The dual-notation equations cover *what shape* the model has. To see *what
numbers* actually flow through that shape, call `as_html_three_views(sym)`:
a single self-contained HTML widget with three tabs over the same fit.

```{r three-views, results = 'asis'}
as_html_three_views(sym, head = 5, tail = 2)
```

Tab 1 (Equation) is the structural contract -- $\mathbf{y}$, $\boldsymbol{\mu}$,
$\boldsymbol{\beta}$, $\mathbf{X}$ -- with no indices. Tab 2 (Index) drops to
per-observation form: $y_i$, $\mu_i$, $\beta_0$, $\beta_1$, $T_i$. Tab 3 (Matrix
with data) actually stacks the numeric arrays: the response column, the rows
of the design matrix, the coefficient vector, fitted $\hat{\boldsymbol{\mu}}$
and $\hat{\boldsymbol{\sigma}}$. The accessor that returns those numeric arrays
without the HTML wrapping is `expand(sym)`.

**Takeaway.** Three views, one fit. A biologist can flip between *the shape of
the model*, *what happens per observation*, and *the data the computer is
actually multiplying*.

## Symbols, assumptions, formula bridge

`symbol_table()` lists every symbol that appears in the equations together
with its role, units, dimension, and description. Dimensions appear in both
abstract and concrete forms.

```{r symbol-table}
symbol_table(sym)
```

`assumption_table()` distinguishes three statuses: **stated** assumptions are
in the formula (distribution, link, linear predictor); **implied** assumptions
follow from the parameterisation (positivity of $\sigma$, conditional
independence); **not_checked** assumptions are still the user's responsibility
(missing-at-random).

```{r assumption-table}
assumption_table(sym)
```

**Note.** The status column shows whether the assumption is *explicit*
(written in the formula), *follows from the formula* (implied by the link or
parameterization), or *your responsibility* (something `symbolizer` cannot
check from the fit, e.g. missing-at-random).

`formula_bridge()` translates R syntax to mathematics. Each submodel has its
R formula on the left, a plain-English meaning, and the corresponding math in
both notations.

```{r formula-bridge}
formula_bridge(sym)
```

**Takeaway.** What the model says, what it assumes, and how its R syntax maps
to math live in three separate tables — so reviewers can audit each one
independently.

## Parameter interpretation

`parameter_interpretation()` reads each fixed-effect estimate on multiple
scales. By default (`scale = "all"`) it shows link, natural, variance (for
$\sigma$ coefficients), and biological readings together:

```{r parameter-interpretation}
parameter_interpretation(sym)
```

The `scale` argument filters to a single reading. The `"biological"` scale is
the one that goes in a results paragraph:

```{r parameter-interpretation-biological}
parameter_interpretation(sym, scale = "biological")
```

**Takeaway.** Every coefficient has an interpretation on every scale that
makes sense for its submodel; the package writes them so you do not have to.

## Random intercepts (first slice)

v0.1 supports Gaussian random intercepts of the form `(1 | group)` as a *first
slice*. The same `symbolize()` call works; the equation set picks up the
random-intercept term and a new distribution row, and the structured object
gains a `random_effects` and `variance_components` tibble.

```{r re-fit}
set.seed(2)
n        <- 80
n_groups <- 6
group    <- factor(rep(letters[seq_len(n_groups)], length.out = n))
temperature <- runif(n, 10, 25)
re <- rnorm(n_groups, sd = 1)
dat_re <- data.frame(
  body_mass = rnorm(
    n,
    30 + 0.4 * temperature + re[as.integer(group)],
    exp(0.5 + 0.05 * temperature)
  ),
  temperature = temperature,
  group       = group
)

fit_re <- drmTMB(
  drm_formula(body_mass ~ temperature + (1 | group), sigma ~ temperature),
  family = gaussian(),
  data   = dat_re
)

sym_re <- symbolize(
  fit_re,
  symbols = c(body_mass = "W_i", temperature = "T_i"),
  units   = c(body_mass = "g",   temperature = "C")
)
```

The mu linear predictor now carries a $+ u_{group(i)}$ term, and a new
distributional row appears for the random intercepts:

```{r re-equations}
equations(sym_re)
```

```{r re-latex, results = 'asis'}
cat("$$", as_latex(sym_re, notation = "both"), "$$", sep = "\n")
```

The grouping structure is exposed as tidy tibbles, ready to feed into reports
or other renderers:

```{r re-tables}
sym_re$random_effects
sym_re$variance_components
```

**Takeaway.** The same object carries everything a reader needs to discuss
random structure — what is grouped, how many levels, and which
between-group SD (`sd(group)`) goes with which submodel.

## What's supported, what's planned

`symbolizer_capabilities()` is the registry that gates `symbolize()`. Each row
carries one of five status words: **Stable**, **First slice**, **Opt-in
control**, **Planned or reserved**, or **Unsupported**. Today (v0.15)
`symbolize()` reads ten package families: `drmTMB`, `gllvmTMB`, `glmmTMB`,
`brms`, `lme4` (`lmer` + `glmer`), `MCMCglmm` (including animal models),
`sdmTMB` (spatial + spatiotemporal fields), `stats::lm` / `stats::glm`,
`metafor` (`rma.uni` + `rma.mv`, including location-scale meta-regression),
and `mgcv::gam` / `mgcv::bam` (with `gamm` / `gamm4` via the `$gam` slot).
The Gaussian location-scale path on `drmTMB` shown above is the most
mature; other classes are mostly First slice.

```{r capabilities}
symbolizer_capabilities()
```

The [Roadmap article](symbolizer-roadmap.html) lays out the full capability
matrix and what's planned next. If `symbolize()` errors with a capability
message, the registry tells you which version is scheduled to lift it.

**Takeaway.** The registry is the single source of truth for "what works".
Read the status word before designing around an unsupported tuple.
