---
title: "Multi-node structural equation models: piecewiseSEM and drmSEM"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Multi-node structural equation models: piecewiseSEM and drmSEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The live path fits a `piecewiseSEM::psem()` on the bundled `keeley`
# dataset, so it runs wherever piecewiseSEM is installed. The drmSEM
# section is shown as reference code only (eval = FALSE) -- it needs the
# off-CRAN drmSEM + a compiled drmTMB fit -- so the vignette stays
# build-safe, matching the sibling cross-package articles.
has_psem <- requireNamespace("piecewiseSEM", quietly = TRUE) &&
            requireNamespace("symbolizer", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_psem
)
library(symbolizer)
```

> *A piecewise structural equation model is a **system** of regressions —
> one fitted model per response, wired together into a hypothesized causal
> graph. This article shows how `symbolize()` reads the whole system: it
> walks each node, symbolizes it with the node's own class method, and
> hands back a collator whose equations, assumption table, and HTML widget
> are stacked node by node. Two flavours are covered: mean-only piecewise
> SEMs from `piecewiseSEM`, and distributional (location-scale — modelling
> spread or skew, not just the mean) piecewise SEMs from `drmSEM`.*

`symbolize()` reports the **structure** of a model system — its
responses, predictors, links, and distributional assumptions — not the
fitted coefficients, significance, or goodness-of-fit. The directed
paths below are the analyst's *assumed* structure; whether the data
support them is a separate question (see Shipley's d-separation test in
`piecewiseSEM`).

## A SEM is a list of regressions

Where a single GLM has one response, a piecewise SEM has several — each
its own fitted model, each a **node** in a directed graph. A variable
that is a predictor in one node is a response in another.

We use the `keeley` data bundled with `piecewiseSEM`: 90 plots of
California shrubland surveyed after wildfire (Grace & Keeley 2006). Five
of its columns appear below — stand **age**, **fire severity**
(`firesev`), an **abiotic** favourability index (`abiotic`), plant
**cover**, and species **richness** (`rich`). A simplified slice of the
Grace & Keeley causal backbone posits four directed paths plus one
correlated-error arc:

- `firesev ~ age` — the model treats fire severity as a response to
  stand age (older stands carry more accumulated fuel),
- `cover ~ firesev` — plant cover as a response to fire severity,
- `rich ~ abiotic + firesev` — richness as a response to abiotic
  favourability and fire severity,
- `rich %~~% cover` — richness and cover are allowed a residual
  covariance (some shared unmeasured cause beyond their common
  dependence on fire), declared as a bidirected arc, **not** a
  regression. *(This arc is added here to illustrate the feature; the
  age→firesev→cover/rich paths are the simplified published backbone.)*

`piecewiseSEM::psem()` glues those fitted nodes into one object.
`symbolize()` reads it as a unit.

```{r psem-fit, message = FALSE}
library(piecewiseSEM)
data(keeley)

sem <- psem(
  lm(firesev ~ age,              data = keeley),
  lm(cover   ~ firesev,          data = keeley),
  lm(rich    ~ abiotic + firesev, data = keeley),
  rich %~~% cover,
  data = keeley
)

sym <- symbolize(
  sem,
  symbols = c(firesev = "F_i", cover = "C_i", rich = "R_i",
              age = "A_i", abiotic = "B_i")
)
sym
```

`symbolize()` returns a `symbolized_psem` — a collator holding one
`symbolized_model` per node (`sym$parts`), the node names in declared
order (`sym$node_names`), and any `%~~%` arcs (`sym$cov_arcs`). Every
node was dispatched on its own class:

```{r psem-parts}
sym$node_names
vapply(sym$parts, function(p) p$model$class, character(1))
```

## The equations, node by node

`equations()` row-binds every node's equation block with a leading
`node` column, so you can see the whole system at once:

```{r psem-equations}
equations(sym)
```

`as_latex()` concatenates the per-node LaTeX under `## Node:` headers,
ready to splice into a manuscript:

```{r psem-latex}
cat(as_latex(sym))
```

## The assumptions — including the covariance arc

`assumption_table()` row-binds each node's assumptions, again with a
leading `node` column. The `%~~%` arc surfaces here as a
`Residual cov(...)` row — and **only** here. A bidirected arc (the
`%~~%` arc) is a covariance, not a regression, so it never appears as an
equation. It records only that two residuals *covary*; it neither names
the source nor implies one response affects the other:

```{r psem-assumptions}
assumption_table(sym)
```

## Three views of the whole SEM

`as_html_three_views()` renders the interactive three-views widget for
**every node**, stacked under node headers: the per-observation *Index*
form, the *Matrix* form, and the matrix form populated with the node's
own real data (see `vignette("symbolizer")` for what each view shows).

```{r psem-widget, results = "asis", message = FALSE}
as_html_three_views(sym, id = "psem")
```

## The distributional flavour: `drmSEM`

A `piecewiseSEM` node is mean-only — one linear predictor per response.
It can model how the **mean** of cover responds to fire severity, but not
how the **spread** of cover changes. When a path needs a
**distributional** structure (a scale submodel, a zero-inflation
component, a residual correlation between two responses), the node is a
`drmTMB` location-scale fit and the SEM is built with
[`drmSEM::drm_sem()`](https://itchyshin.github.io/drmSEM/). `symbolize()`
reads it through the `drmSEM`-side `symbolize.drm_sem` method, which
delegates each node to `symbolize.drmTMB`.

The code below is **not run here** (it needs the `drmSEM` and `drmTMB`
packages); it continues the same `keeley` story. Suppose we wanted to
ask whether the residual *spread* of cover widens at high fire severity —
plausible when severe burns produce divergent recovery — which the
mean-only `lm()` node cannot express:

```{r drmsem-ref, eval = FALSE}
library(drmSEM)
library(drmTMB)

sem <- drm_sem(
  # The `cover` node gains a SCALE submodel: `sigma ~ firesev` lets the
  # variability of cover -- not just its mean -- depend on fire severity.
  cover   = drm_node(drm_formula(cover ~ firesev, sigma ~ firesev),
                     family = gaussian()),
  firesev = drm_node(drm_formula(firesev ~ age),
                     family = gaussian()),
  data = keeley
)

sym <- symbolize(sem)     # dispatches to drmSEM::symbolize.drm_sem
sym                       # <symbolized_drm_sem>: 2 nodes ...
equations(sym)            # node column + per-node mu AND sigma rows
assumption_table(sym)     # per-node assumptions, distributional
as_html_three_views(sym)  # the same stacked widget, location-scale nodes
```

Everything downstream — the equations, the assumption table, the HTML
widget — works identically on the `drm_sem`, because both collators share
one parent class. See the
[drmSEM site](https://itchyshin.github.io/drmSEM/) for runnable
distributional-SEM examples.

## Why the same methods work on both

Every renderer you just saw — `equations()`, `as_latex()`,
`assumption_table()`, `as_html_three_views()` — works identically on a
`drm_sem`. Here is the mechanism. `symbolize.psem` tags its output
`symbolized_psem`, which inherits from `symbolized_model_set`; `drmSEM`'s
`symbolize.drm_sem` tags its output `symbolized_drm_sem`, which inherits
from the same parent. Both store their nodes in a **named** `parts` list.
So the rendering generics dispatch on the shared parent and read
`names(x$parts)` — one implementation, two collators. The capability
registry records which package owns each method:

```{r registry}
caps <- symbolizer_capabilities()
caps[caps$class %in% c("piecewiseSEM", "drm_sem"),
     c("class", "status", "since", "lives_in", "notes")]
```

| | `piecewiseSEM` (`psem`) | `drmSEM` (`drm_sem`) |
|---|---|---|
| Per-node model | `lm` / `glm` / `lmer` / `glmer` / `glmmTMB` | `drmTMB` location-scale |
| Structure per node | mean only (one linear predictor) | distributional (`mu`, `sigma`, `nu`, `zi`, `hu`, `rho12`) |
| Method lives in | `symbolizer` (bundled) | `drmSEM` |
| Collator class | `symbolized_psem` → `symbolized_model_set` | `symbolized_drm_sem` → `symbolized_model_set` |
| `%~~%` / residual covariance | `Residual cov(...)` assumption row | per-node `rho12` distributional component |

Reach for `piecewiseSEM` when each response is adequately summarised by
its mean; reach for `drmSEM` when a path's **variance, skew, or
zero-inflation** is itself part of the story.

## References

Grace, J. B., & Keeley, J. E. (2006). A structural equation model
analysis of postfire plant diversity in California shrublands.
*Ecological Applications*, 16(2), 503–514.
<https://doi.org/10.1890/1051-0761(2006)016%5B0503:ASEMAO%5D2.0.CO;2>

Lefcheck, J. S. (2016). piecewiseSEM: Piecewise structural equation
modelling in R for ecology, evolution, and systematics. *Methods in
Ecology and Evolution*, 7(5), 573–579.
<https://doi.org/10.1111/2041-210X.12512>

Shipley, B. (2009). Confirmatory path analysis in a generalized
multilevel context. *Ecology*, 90(2), 363–368.
<https://doi.org/10.1890/08-1034.1>
