---
title: "Latent variables in ecology: a gllvmTMB worked example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Latent variables in ecology: a gllvmTMB worked example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_pkgs <- requireNamespace("gllvmTMB", quietly = TRUE) &&
            requireNamespace("drmTMB",   quietly = TRUE) &&
            requireNamespace("lme4",     quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_pkgs
)
```

## 1. The biological question

A behavioural ecologist measures five traits — boldness in a novel arena,
exploration in a maze, aggression to a mirror, activity in the home cage,
and time-out-of-shelter — on each of forty fish, three times each. Two
themes keep coming up in the data. Bolder fish also tend to be more
exploratory. Active fish tend to spend more time out of shelter. There
is *also* a second pattern: when an individual fish is bolder than usual
on a given day, it tends to be more exploratory and more aggressive on
that *same* day. The first pattern is about stable differences between
individuals; the second is about coordinated within-individual
fluctuations.

The literature names them separately:

- *Behavioural syndromes* (Sih et al. 2004; Dingemanse & Dochtermann
  2013): the **between-individual** axes that organise trait values
  across the population — what makes individual A's boldness reliably
  predict individual A's exploration.
- *Integrated plasticity* (Dingemanse & Dochtermann 2013; Mathot &
  Dingemanse 2015): the **within-individual** axes that organise
  trait fluctuations across occasions for the *same* individual.

Both are factor-analytic objects in the same model. The output a
biologist wants is **two loading matrices** ($\boldsymbol{\Lambda}_B$
for between, $\boldsymbol{\Lambda}_W$ for within), the per-trait
**communality** at each tier, and the per-trait **repeatability**
$R_t = (\boldsymbol{\Sigma}_B)_{tt} / [(\boldsymbol{\Sigma}_B)_{tt} +
(\boldsymbol{\Sigma}_W)_{tt}]$.

**Takeaway.** The biological question splits into two layers:
*syndromes* (between) and *integrated plasticity* (within). One
generalised linear latent-variable model (GLLVM) writes both layers in
one fit.

## 2. The data

We simulate from a known truth so the biological reading at the end
can be checked. The truth puts traits 1-3 on a first between-individual
axis (the *boldness-exploration-aggression* syndrome), traits 4-5 on a
second between-individual axis (the *activity-shelter-use* syndrome),
and adds a single shared **within-individual** axis (a coordinated
state-shift that moves all five behaviours together across occasions).

```{r data}
library(symbolizer)
library(gllvmTMB)
library(lme4)

set.seed(20260523)
n_ind  <- 40L  # individuals (units)
n_sess <- 3L   # repeated sessions per individual
n_tr   <- 5L   # traits

# True BETWEEN-individual loading matrix Lambda_B (5 traits x 2 latents).
# Traits 1-3 load on axis 1 (the bold-explore-aggress syndrome); traits
# 4-5 load on axis 2 (the activity-shelter syndrome).
Lam_B <- matrix(c(1.0, 0.0,
                  0.7, 0.0,
                  0.6, 0.0,
                  0.0, 1.0,
                  0.0, 0.7),
                nrow = n_tr, ncol = 2L)
psi_B <- rep(0.4, n_tr)   # trait-specific between-individual uniqueness SD

# True WITHIN-individual loading vector Lambda_W (5 traits x 1 latent).
# A single state-shift axis: when an individual is "up" today, all five
# behaviours move together by trait-specific amounts. This is what
# integrated-plasticity literature calls a coordinated state.
Lam_W <- matrix(c(0.5, 0.3, 0.4, 0.3, 0.4), nrow = n_tr, ncol = 1L)
psi_W <- rep(0.2, n_tr)   # trait-specific within-individual uniqueness SD

# Per-individual between-latent scores z_B in R^2 (one per individual).
z_B <- matrix(rnorm(n_ind * 2L), nrow = n_ind, ncol = 2L)
psi_B_resid <- matrix(rnorm(n_ind * n_tr,
                            sd = rep(psi_B, each = n_ind)),
                      nrow = n_ind, ncol = n_tr)
# Between-individual trait mean (per individual, per trait).
mu_ind_trait <- z_B %*% t(Lam_B) + psi_B_resid

# Per-observation within-latent scores z_W in R^1 (one per (ind, sess)).
n_obs_per_ind <- n_sess
z_W <- matrix(rnorm(n_ind * n_sess), nrow = n_ind * n_sess, ncol = 1L)
psi_W_resid <- matrix(rnorm(n_ind * n_sess * n_tr,
                            sd = rep(psi_W, each = n_ind * n_sess)),
                      nrow = n_ind * n_sess, ncol = n_tr)
within_state <- z_W %*% t(Lam_W) + psi_W_resid    # (n_ind * n_sess) x n_tr

# Expand to long-format: one row per (individual, session, trait). Trait
# labels are short codes (t1..t5) for the math; their biological readings
# from §1 are:
#   t1 = boldness          t4 = activity
#   t2 = exploration       t5 = time-out-of-shelter
#   t3 = aggression
dat <- expand.grid(
  trait      = factor(paste0("t", seq_len(n_tr)),
                      levels = paste0("t", seq_len(n_tr))),
  session    = factor(seq_len(n_sess)),
  individual = factor(seq_len(n_ind))
)
# Look up the row index in `within_state` for each (individual, session).
obs_idx <- as.integer(dat$individual) * n_sess - n_sess +
           as.integer(dat$session)
dat$value <- mu_ind_trait[cbind(as.integer(dat$individual),
                                 as.integer(dat$trait))] +
              within_state[cbind(obs_idx,
                                  as.integer(dat$trait))]
# Stable (individual, session) observation id for the obs-level term.
dat$obs <- factor(paste(dat$individual, dat$session, sep = "_"))
head(dat, 6)
```

The DGP carries both layers: a stable between-individual structure
($\boldsymbol{\Lambda}_B, \boldsymbol{\Psi}_B$) and a coordinated
within-individual structure ($\boldsymbol{\Lambda}_W,
\boldsymbol{\Psi}_W$). When we fit Widget 2 below we will recover both.

**Takeaway.** Forty individuals, three sessions each, five traits —
long format, one row per (individual, session, trait). The simulated
truth has *both* a between-individual two-axis structure and a
within-individual one-axis structure.

## 3. The univariate root: one trait at a time

Before we model five traits jointly, recall the univariate mixed model
behavioural ecologists already know. For one trait $y$ measured on $N$
individuals at $J_i$ occasions:

$$
y_{ij} = \mu + u_i + e_{ij}, \qquad
u_i \sim \mathcal{N}(0, \sigma^2_u), \qquad
e_{ij} \sim \mathcal{N}(0, \sigma^2_e).
$$

The classical repeatability (Nakagawa & Schielzeth 2010) is

$$
R = \frac{\sigma^2_u}{\sigma^2_u + \sigma^2_e}.
$$

In `lme4` syntax, fit one trait at a time:

```{r univariate}
fit_uni <- lmer(value ~ 1 + (1 | individual),
                data = subset(dat, trait == "t1"))
vc <- as.data.frame(VarCorr(fit_uni))
vc[, c("grp", "vcov")]
sigma2_u <- vc$vcov[vc$grp == "individual"]
sigma2_e <- vc$vcov[vc$grp == "Residual"]
R_t1 <- sigma2_u / (sigma2_u + sigma2_e)
cat("Repeatability for t1:", round(R_t1, 3), "\n")
```

The random intercept `(1 | individual)` captures stable between-individual
variation; everything else lands in the residual. The repeatability $R$
is the between-individual share of total variance — computed
trait-by-trait, it's the univariate readout that the GLLVM upgrades to
two structured matrices.

**Takeaway.** Univariate $R = \sigma^2_u / (\sigma^2_u + \sigma^2_e)$ is
the trait-by-trait readout the multivariate model generalises. The
GLLVM upgrade replaces the two scalars with two $T \times T$ matrices
$\boldsymbol{\Sigma}_B, \boldsymbol{\Sigma}_W$ — and decomposes each
factor-analytically.

## 4. Why factor-analytic? The curse-of-dimensionality argument

A fully parameterised multivariate covariance for $T$ traits requires
$T(T+1)/2$ free parameters. For $T = 5$ traits at one tier that's 15
parameters; with both between and within tiers that doubles to 30. At
$T = 10$ traits per tier we'd need 110 covariance parameters before any
fixed effects enter. Animal-personality studies typically have
$N \approx 40$–$200$ individuals; the resulting likelihood is *flat* —
convergence is unstable and standard errors are wide enough that
interpretation becomes risky (McGillycuddy et al. 2025).

The **factor-analytic** answer is to decompose

$$
\boldsymbol{\Sigma}_B = \boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^{\!\top} + \boldsymbol{\Psi}_B,
$$

with $\boldsymbol{\Lambda}_B \in \mathbb{R}^{T \times d_B}$ a reduced-rank
loading matrix ($d_B \ll T$) and $\boldsymbol{\Psi}_B$ a diagonal of
trait-specific uniquenesses. The parameter count drops from $T(T+1)/2$ to

$$
T(d_B + 1) - d_B(d_B - 1)/2.
$$

For $T = 5, d_B = 2$: $15 \to 14$; for $T = 10, d_B = 2$: $55 \to 29$;
for $T = 10$ at both tiers with $d_B = d_W = 2$: $110 \to 58$. The
reduction is modest when $T$ is small and dramatic as $T$ grows.

The *interpretive* payoff is the same shape both directions. The shared
loadings $\boldsymbol{\Lambda}_B$ *name* the between-individual axes —
the syndromes. The uniquenesses $\boldsymbol{\Psi}_B$ name what the
shared axes do not explain — trait-specific between-individual variance.
At the within-individual tier, $\boldsymbol{\Lambda}_W$ names the
coordinated states; $\boldsymbol{\Psi}_W$ names the trait-specific
within-individual residual variance. **Both tiers are factor-analytic
in the same way**; that's the unifying move.

**Takeaway.** Reduced-rank decomposition
$\boldsymbol{\Sigma} = \boldsymbol{\Lambda}\boldsymbol{\Lambda}^{\!\top} + \boldsymbol{\Psi}$
stabilises the fit *and* surfaces the biology — and it applies to both
tiers without changing form.

## 5. The model in symbols

The same model can be written two ways. The **long form** matches the
way `gllvmTMB()` actually fits the data — one row per (individual,
occasion, trait) cell, with a scalar response. The **wide form**
matches the factor-analytic textbooks — one row per observation, with
a trait-vector response. The two panels below place them side by side;
both decompose each tier's covariance via
$\boldsymbol{\Sigma} = \boldsymbol{\Lambda}\boldsymbol{\Lambda}^{\!\top} + \boldsymbol{\Psi}$.

```{r model-long-wide, results = 'asis', echo = FALSE, eval = TRUE}
cat('<div style="display: flex; gap: 1rem; flex-wrap: wrap; margin: 1em 0;">

<div style="flex: 1; min-width: 320px;">

#### Long form

Each row of the data is one (individual, occasion, trait) cell. The
scalar response is $y_{ijt}$.

$$
y_{ijt} \\mid \\boldsymbol{\\mu},\\, \\boldsymbol{\\Lambda}_B,\\, \\mathbf{z}_{B,i},\\, \\boldsymbol{\\Lambda}_W,\\, \\mathbf{z}_{W,ij},\\, \\boldsymbol{\\Psi}
  \\sim \\mathcal{N}\\!\\left(\\mu_t + (\\boldsymbol{\\Lambda}_B \\mathbf{z}_{B,i})_t + (\\boldsymbol{\\Lambda}_W \\mathbf{z}_{W,ij})_t,\\;
   \\psi_{W,t}^2 + (\\boldsymbol{\\Lambda}_W \\boldsymbol{\\Lambda}_W^{\\!\\top})_{tt}\\right)
$$

Native to `gllvmTMB`\'s formula interface. Each row\'s contribution is a
scalar Gaussian.

</div>

<div style="flex: 1; min-width: 320px;">

#### Wide form

Each row is one observation; the response is a $T$-vector
$\\mathbf{Y}_{ij}$.

$$
\\mathbf{Y}_{ij} \\mid \\boldsymbol{\\mu},\\, \\boldsymbol{\\Lambda}_B,\\, \\mathbf{z}_{B,i},\\, \\boldsymbol{\\Sigma}_W
  \\sim \\mathcal{MN}\\!\\left(\\boldsymbol{\\mu} + \\boldsymbol{\\Lambda}_B\\,\\mathbf{z}_{B,i},\\; \\boldsymbol{\\Sigma}_W\\right),
\\quad \\boldsymbol{\\Sigma}_W = \\boldsymbol{\\Lambda}_W\\boldsymbol{\\Lambda}_W^{\\!\\top} + \\boldsymbol{\\Psi}_W
$$

Native to factor-analytic textbooks. The within-individual conditional
covariance $\\boldsymbol{\\Sigma}_W = \\boldsymbol{\\Lambda}_W\\boldsymbol{\\Lambda}_W^{\\!\\top} + \\boldsymbol{\\Psi}_W$
replaces the scalar $\\sigma_\\epsilon^2 \\mathbf{I}_T$ row-level
residual when the within-tier is structured.

</div>

</div>
')
```

*Bridge.* The two forms describe the same model. The wide form makes
the covariance decomposition explicit; the long form is what
`gllvmTMB()` actually fits.

Index $i$ ranges over individuals, $j$ over occasions, $t$ over traits,
$k = 1, \ldots, d_B$ over between-axes, $\ell = 1, \ldots, d_W$ over
within-axes. $\mu_t$ is the trait $t$ grand mean; $\lambda_{B,tk}$ is
the loading of trait $t$ on between-axis $k$; $\lambda_{W,t\ell}$ is
the loading of trait $t$ on within-axis $\ell$; $z_{B,ik}$ is individual
$i$'s position on between-axis $k$; $z_{W,ij\ell}$ is (individual $i$,
session $j$)'s position on within-axis $\ell$;
$\boldsymbol{\Psi}_B = \mathrm{diag}(\psi_{B,1}^2, \ldots, \psi_{B,T}^2)$
collects the trait-specific between-individual uniquenesses; similarly
$\boldsymbol{\Psi}_W$ for within-individual.

**Takeaway.** Long form is one scalar equation per (individual,
occasion, trait); wide form is one $T$-vector equation per observation.
Each tier has its own $\boldsymbol{\Lambda}\boldsymbol{\Lambda}^{\!\top} + \boldsymbol{\Psi}$
decomposition. Same model, two notations.

## 6. Widget 1 — Behavioural syndromes

The first fit decomposes the **between-individual** trait covariance
only. Each trait gets its own grand mean; the shared between-individual
structure is captured by a rank-$d_B$ loading matrix
$\boldsymbol{\Lambda}_B$ plus per-trait uniquenesses
$\boldsymbol{\Psi}_B$:

$$
\boldsymbol{\Sigma}_B = \boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^{\!\top} + \boldsymbol{\Psi}_B.
$$

The within-individual residual is the shared row-level Gaussian
$\sigma_\epsilon^2$ — a single scalar, unstructured.

```{r fit-syndromes}
fit_syndromes <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | individual, d = 2) +
          unique(0 + trait | individual),
  data     = dat,
  family   = gaussian(),
  trait    = "trait",
  unit     = "individual",
  silent   = TRUE
)
sym_syndromes <- symbolize(
  fit_syndromes,
  symbols = c(value = "y_{ij}", trait = "j", individual = "i"),
  context = "behavioural syndromes (between-individual only)"
)
```

### Three views — syndromes

```{r widget-syndromes, results = 'asis', echo = FALSE, message = FALSE}
as_html_three_views(sym_syndromes, id = "syndromes")
```

The widget's **Tab 3 — Equations with data** carries the new
implied-covariance block: a $5 \times 5$ numerical
$\boldsymbol{\Sigma}_B$ shown next to its decomposition
$\boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^{\!\top} +
\boldsymbol{\Psi}_B^{\,2}$. The arithmetic closes element-by-element
to within rounding.

**Takeaway.** Widget 1 is the canonical *behavioural syndromes* fit.
The two-axis $\boldsymbol{\Lambda}_B$ names the syndromes (which traits
move together between individuals); $\boldsymbol{\Psi}_B$ captures the
leftover per-trait between-individual variance.

## 7. Widget 2 — Adding integrated plasticity

The second fit adds the **within-individual** reduced-rank structure.
$\boldsymbol{\Lambda}_W$ captures shared within-individual axes (how
traits covary across occasions, within an individual);
$\boldsymbol{\Psi}_W$ captures per-trait within-individual uniquenesses.
Together:

$$
\boldsymbol{\Sigma}_W = \boldsymbol{\Lambda}_W \boldsymbol{\Lambda}_W^{\!\top} + \boldsymbol{\Psi}_W.
$$

When `unique(0 + trait | obs)` is added, `gllvmTMB` prints an info
message auto-suppressing $\sigma_\epsilon$:

> ℹ Auto-suppressing `sigma_eps`: `unique(0 + trait | obs)` is at the
> per-row level, so it already absorbs the observation residual.
> • Fixed at 0.00111 (~1/1000 of sd(y)) to keep the Gaussian density
> well-defined; the row-level residual variance is fully captured by
> `unique()`.

This is intended — the row-level residual variance is now structured
per-trait in $\boldsymbol{\Psi}_W$ rather than carried by a single
scalar $\sigma_\epsilon^2$. The package handles the constraint for you;
no `dispformula = ~0` ceremony required (that's the `glmmTMB`
equivalent, covered in §10).

```{r fit-two-tier, message = FALSE, warning = FALSE}
fit_two_tier <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | individual, d = 2) +
          unique(0 + trait | individual) +
          latent(0 + trait | obs, d = 1) +
          unique(0 + trait | obs),
  data     = dat,
  family   = gaussian(),
  trait    = "trait",
  unit     = "individual",
  unit_obs = "obs",
  cluster  = "session",
  silent   = TRUE
)
sym_two_tier <- symbolize(
  fit_two_tier,
  symbols = c(value = "y_{ij}", trait = "j", individual = "i"),
  context = "syndromes + integrated plasticity (two-tier)"
)
```

### Three views — two-tier

```{r widget-two-tier, results = 'asis', echo = FALSE, message = FALSE}
as_html_three_views(sym_two_tier, id = "two-tier")
```

What changes between Widget 1 and Widget 2:

- **Tab 1 (Index)**: the conditional mean gains
  $(\boldsymbol{\Lambda}_W \mathbf{z}_{W,ij})_t$. The conditional residual
  variance loses $\sigma_\epsilon^2$ and gains
  $\psi_{W,t}^2 + (\boldsymbol{\Lambda}_W \boldsymbol{\Lambda}_W^{\!\top})_{tt}$
  per trait.
- **Tab 2 (Matrix)**: the matrix-Normal carries both shared structures —
  $\boldsymbol{\eta} = \mathbf{1}_n \boldsymbol{\mu}^{\!\top} +
  \mathbf{Z}_B \boldsymbol{\Lambda}_B^{\!\top} +
  \mathbf{Z}_W \boldsymbol{\Lambda}_W^{\!\top}$, and the across-trait
  conditional covariance becomes $\boldsymbol{\Sigma}_W$ instead of
  $\sigma_\epsilon^2 \mathbf{I}_T$.
- **Tab 3 (Equations with data)**: gains a second implied-covariance
  block for $\boldsymbol{\Sigma}_W$, plus a per-trait repeatability row
  $R_t = (\boldsymbol{\Sigma}_B)_{tt} /
  [(\boldsymbol{\Sigma}_B)_{tt} + (\boldsymbol{\Sigma}_W)_{tt}]$.

**Takeaway.** Widget 2 carries the full two-tier model. The reader can
*see* the syndromes ($\boldsymbol{\Lambda}_B$, $\boldsymbol{\Sigma}_B$),
the integrated plasticity ($\boldsymbol{\Lambda}_W$,
$\boldsymbol{\Sigma}_W$), and the per-trait repeatability — all in one
widget, with arithmetic that closes.

## 8. Reading the latent axes biologically

Two payoffs once you have $\boldsymbol{\Lambda}_B, \boldsymbol{\Psi}_B,
\boldsymbol{\Lambda}_W, \boldsymbol{\Psi}_W$ in hand.

### 8.1 Communality and uniqueness at each tier

For trait $t$ at tier $g \in \{B, W\}$:

$$
c^2_{g,t} = \frac{\sum_{k} \lambda_{g,tk}^2}{(\boldsymbol{\Sigma}_g)_{tt}},
\qquad
\psi^*_{g,t} = \frac{(\boldsymbol{\Psi}_g)_{tt}}{(\boldsymbol{\Sigma}_g)_{tt}},
\qquad c^2_{g,t} + \psi^*_{g,t} = 1.
$$

$c^2_{g,t}$ is the share of trait $t$'s total variance at tier $g$ that
the shared axes explain; $\psi^*_{g,t}$ is what's idiosyncratic to that
trait at that tier.

```{r communality}
extract_communality(fit_two_tier)
```

Mapping back to §1's named behaviours: **t1 = boldness, t2 =
exploration, t3 = aggression, t4 = activity, t5 = time-out-of-shelter**.
So a high $c^2_B$ for *t1* (boldness) means the between-individual
component of boldness is well-explained by the shared two-axis
syndrome structure; a low $c^2_B$ for *t3* (aggression) would mean
aggression's between-individual variation is mostly trait-idiosyncratic
($\boldsymbol{\Psi}_B$), not part of the shared syndromes.

A trait with high $c^2_B$ and high $c^2_W$ is *integrated at both
tiers*: both its between-individual position and its within-individual
fluctuation share the syndrome structure. A trait with high $c^2_B$ but
low $c^2_W$ is *syndromatic but not plastically integrated* —
individuals' average levels align, but their occasion-to-occasion
fluctuations don't.

### 8.2 Repeatability and phenotypic-correlation decomposition

Per-trait repeatability already appears at the bottom of Widget 2's
Tab 3:

$$
R_t = \frac{(\boldsymbol{\Sigma}_B)_{tt}}{(\boldsymbol{\Sigma}_B)_{tt} + (\boldsymbol{\Sigma}_W)_{tt}}.
$$

The phenotypic correlation between traits $t$ and $m$ decomposes as

$$
r_{P,tm} = r_{B,tm}\sqrt{R_t R_m} + r_{W,tm}\sqrt{(1 - R_t)(1 - R_m)},
$$

so the phenotypic-level correlation is a *weighted mix* of the
between-individual and within-individual correlations, weighted by how
much of each trait's variance lives at each tier (Dingemanse &
Dochtermann 2013). Symbolizer's `extract_correlations(fit)` returns
all three matrices ($\boldsymbol{C}_B, \boldsymbol{C}_W,
\boldsymbol{C}_P$) so a reader can verify this empirically:

```{r correlations}
co <- extract_correlations(fit_two_tier)
co_B <- co[co$tier == "B", c("trait_i", "trait_j", "correlation")]
co_W <- co[co$tier == "W", c("trait_i", "trait_j", "correlation")]
head(co_B)
head(co_W)
```

**Takeaway.** $c^2$ and $\psi^*$ per tier are the trait-level integration
summaries; $R_t$ is the per-trait repeatability; phenotypic correlations
decompose as a weighted mix of between and within. All three readings
fall out of the same two fitted matrices.

## 9. Identifiability gotchas: rotation, sign, and the within-tier σε

The loading matrices $\boldsymbol{\Lambda}_B, \boldsymbol{\Lambda}_W$
are identified up to **per-column sign at any rank**, and
*additionally* up to **rotation when their rank exceeds 1**. So even
the rank-1 $\boldsymbol{\Lambda}_W$ in this fit is fixed only up to a
column reflection: flipping every entry in a column and the
corresponding latent scores leaves the likelihood unchanged. Two fits
that produce indistinguishable likelihoods can therefore have
entirely different-looking $\boldsymbol{\Lambda}_B$ — this is the
classical factor-analysis identifiability problem. The implied
covariance $\boldsymbol{\Sigma}_B = \boldsymbol{\Lambda}_B
\boldsymbol{\Lambda}_B^{\!\top} + \boldsymbol{\Psi}_B$ is invariant
under both rotation and sign change; $\boldsymbol{\Lambda}_B$ itself
is not.

Two practical conventions, both surfaced in
`assumption_table(sym)`:

- **Lower-triangular** $\boldsymbol{\Lambda}_B$ for *confirmatory* use —
  pin the upper triangle to zero so each column has a designated
  leading trait. This is the default identification `gllvmTMB` uses on
  fitting.
- **Varimax-rotated** $\boldsymbol{\Lambda}_B$ for *interpretation* —
  rotate the fitted loadings so each axis loads strongly on a small set
  of traits and near zero on the rest. Applied *after* fitting;
  preserves $\boldsymbol{\Sigma}_B$. The column labels may reorder under
  rotation; the **block structure** is what the reader interprets.

```{r loadings}
getLoadings(fit_two_tier, rotate = "varimax")
```

A second category of identifiability concerns the **row-level residual
variance**. When both `latent(0 + trait | obs, ...)` and
`unique(0 + trait | obs)` are present (Widget 2), `gllvmTMB`
auto-suppresses $\sigma_\epsilon^2$ — pinning it to ~10⁻³ to keep the
Gaussian density well-defined. This is *not* a defect; it's the
package's way of preventing a known collinearity between two
parametrisations of the same row-level variance. The within-tier
$\boldsymbol{\Psi}_W$ already captures per-trait within-individual
residual variance; a free $\sigma_\epsilon^2$ scalar on top would
double-count. Reading the auto-suppression message in the fit output
is the expected behaviour, not a warning to worry about.

A third category: **$\boldsymbol{\Psi}_B$ or $\boldsymbol{\Psi}_W$
entries near zero**. When a trait's between-individual (or
within-individual) variance is already fully explained by the shared
$\boldsymbol{\Lambda}\boldsymbol{\Lambda}^{\!\top}$ structure, the
optimiser pins that trait's uniqueness $\psi$ to the boundary (~10⁻⁴
or smaller). Widget 1's reported $\boldsymbol{\Psi}_B = (0.00004,
0.56, 0.48, 0.93, 0.41)$ for the simulated fit reads as: *trait 1
(boldness) is fully captured by the shared between-individual axes,
so its trait-specific uniqueness disappears.* This is **not** a fit
failure — it is the model telling you which traits sit fully on the
shared structure. The implied covariance
$\boldsymbol{\Sigma}_B = \boldsymbol{\Lambda}_B
\boldsymbol{\Lambda}_B^{\!\top} + \boldsymbol{\Psi}_B$ remains
well-defined; only the *decomposition* loses its uniqueness piece
for that trait.

**Takeaway.** Rotation and sign of $\boldsymbol{\Lambda}$ are *stated
assumptions*, not bugs. The implied covariances are invariant; the
loadings are not. Read the block structure, not the cell values. And
the within-tier $\sigma_\epsilon$ auto-suppression is by design.

## 10. The glmmTMB bridge

`gllvmTMB` is purpose-built for generalised linear *latent-variable*
models. `glmmTMB` is a general-purpose GLMM package that, since
McGillycuddy et al. (2025), can fit the same math via the `rr()` and
`diag()` keywords. The Nakagawa et al. (in prep) framework uses the
`glmmTMB` route; for readers coming from there, the two-tier fit
translates one-to-one:

```{r glmmtmb-equivalence, eval = FALSE}
# gllvmTMB (this article's path)
fit_two_tier <- gllvmTMB(value ~ 0 + trait +
  latent(0 + trait | individual, d = 2) + unique(0 + trait | individual) +
  latent(0 + trait | obs, d = 1) + unique(0 + trait | obs),
  data = dat, family = gaussian(),
  trait = "trait", unit = "individual",
  unit_obs = "obs", cluster = "session")

# glmmTMB (Nakagawa-paper path)
fit_two_tier_glmm <- glmmTMB::glmmTMB(value ~ 0 + trait +
  rr(0 + trait | individual, d = 2) + diag(0 + trait | individual) +
  rr(0 + trait | obs, d = 1) + diag(0 + trait | obs),
  data = dat, family = gaussian(),
  dispformula = ~0)
```

| `gllvmTMB` | `glmmTMB` | Math |
|---|---|---|
| latent(0 + trait \| g, d = k) | rr(0 + trait \| g, d = k) | $\boldsymbol{\Lambda}_g$ of rank $k$ |
| unique(0 + trait \| g) | diag(0 + trait \| g) | $\boldsymbol{\Psi}_g$ diagonal |
| auto-suppression of $\sigma_\epsilon$ on unique(. \| obs) | `dispformula = ~0` | both prevent double-counting $\boldsymbol{\Psi}_W$ vs $\sigma_\epsilon^2$ |

Both packages return the same fitted $\boldsymbol{\Lambda},
\boldsymbol{\Psi}, \boldsymbol{\Sigma}$ matrices to numerical
precision. The article path leads with `gllvmTMB` because it ships
GLLVM-specific accessors (`getLoadings()`, `extract_communality()`,
`extract_correlations()`) and a wide/long dual interface that
`glmmTMB` doesn't have.

**Takeaway.** Same math; different syntax. Use `gllvmTMB` when you want
GLLVM-specific accessors; use `glmmTMB` when your model also needs
zero-inflation or dispersion modelling and you're happy to reconstruct
$\boldsymbol{\Sigma}$ manually.

## 11. What's available now, what's next

The accessors in §8 — `getLoadings()`, `extract_communality()`,
`extract_correlations()` — are `gllvmTMB`-side. The symbolizer surface
wraps the Gaussian latent-variable case as a **First slice** at the
two-tier level: every piece walked above is shipped today.

**Shipped today (v0.21.6-redo):**

- `symbolize.gllvmTMB()` populates the widget-shape slots `Lambda_B`,
  `Psi_B`, `Sigma_B`, `Lambda_W`, `Psi_W`, `Sigma_W`, `Repeatability`
  for the Gaussian latent-variable family.
- `as_html_three_views()` Tab 3 emits a numerical implied-covariance
  block per tier plus a per-trait repeatability row when both tiers
  are present.
- The two-tier `latent(0 + trait | obs, d = d_W) + unique(0 + trait | obs)`
  syntax is fully tested. gllvmTMB's σ_ε auto-suppression is
  documented in `assumption_table(sym)`.
- `compare_symbolic(fit_d1, fit_d2)` diffs the structural specifications
  of $d_B = 1$ versus $d_B = 2$ (or any rank-pair). See
  `vignette("symbolizer-compare")`.

**Still planned:**

1. **Non-Gaussian gllvmTMB families** (count, Tweedie, ordinal) at the
   two-tier level — each adds a link to the linear predictor and
   reshapes the loadings interpretation.
2. **Uncertainty on $\boldsymbol{\Lambda}$.** `bootstrap_Sigma()` from
   `gllvmTMB` returns parametric-bootstrap draws of the loading matrix;
   a future slice will surface these so the `symbol_table()` rows for
   $\boldsymbol{\Lambda}$ and $\boldsymbol{\Sigma}$ carry confidence
   regions, not just point estimates.
3. **Rank selection.** A simple AIC + interpretability grid over
   $d_B, d_W \in \{0, 1, 2, 3\}$ is the standard practical move; a
   future slice will ship this as a helper.
4. **Reaction-norm / sex-specific extensions** (Appendices B.1 / B.2 of
   the Nakagawa et al. (in prep) framework) — fold these into the
   same $\boldsymbol{\Lambda} \boldsymbol{\Lambda}^{\!\top} +
   \boldsymbol{\Psi}$ structure on an augmented random-effect vector.

See `vignette("symbolizer-roadmap")` for the full capability matrix.

## References

- Dingemanse, N. J., & Dochtermann, N. A. (2013). Quantifying individual variation in behaviour: mixed-effect modelling approaches. *Journal of Animal Ecology* 82: 39-54.
- Mathot, K. J., & Dingemanse, N. A. (2015). Energetics and behaviour: unrequited needs and new directions. *Trends in Ecology & Evolution* 30: 199-206.
- McGillycuddy, M., et al. (2025). Reduced-rank covariance structures in `glmmTMB`. (Forthcoming methods paper.)
- Nakagawa, S., & Schielzeth, H. (2010). Repeatability for Gaussian and non-Gaussian data: a practical guide for biologists. *Biological Reviews* 85: 935-956.
- Nakagawa, S., Mathot, K., Dinnage, R., et al. (in prep). Quantifying between- and within-individual correlations and the degree of trait integration: leveraging latent variable modeling to study behavioural syndromes and other phenotypic integration.
- Pigliucci, M. (2003). Phenotypic integration: studying the ecology and evolution of complex phenotypes. *Ecology Letters* 6: 265-272.
- Sih, A., Bell, A., & Johnson, J. C. (2004). Behavioral syndromes: an ecological and evolutionary overview. *Trends in Ecology & Evolution* 19: 372-378.
