---
title: "Structural dependence: phylogenetic, animal-model, and spatial random effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structural dependence: phylogenetic, animal-model, and spatial random effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 3.5
)
library(symbolizer)
set.seed(1)
```

If you have ever written `(1 | species)` for an animal model, `gr(species, cov = A)` in `brms`, or `s(x, y, bs = "gp")` for a spatial process, you have been writing the **same model** three times — only the matrix that encodes the dependence changes. This article is the first in `symbolizer`'s **concept-axis series**; it teaches the *dependence-structure* axis of GLMM grammar through one **deep-dive** Face and two **light** Faces across three R packages: `MCMCglmm`, `brms`, and `phylolm`. The scope is the **phylogenetic LMM** with $\sigma_e^2$ estimated; the *meta-analytic* form (with $v_i$ known sampling variances) is its own surface in the v0.22 article.

## The shared grammar

A generalised linear mixed model writes

$$
y_i \mid \mathbf b \;\sim\; \mathcal D(\mu_i, \phi),\qquad
g(\mu_i) \;=\; \mathbf x_i^\top \boldsymbol\beta + \mathbf z_i^\top \mathbf b,
$$

with random effects

$$
\mathbf b \;\sim\; \mathcal N(\mathbf 0,\; \sigma^2\,\mathbf M).
$$

The standard textbook GLMM takes $\mathbf M = \mathbf I$ — the random effects are independent draws. Structural dependence is what you get when $\mathbf M \ne \mathbf I$. Three named instantiations recur:

| Context | Symbol | What $\mathbf M$ encodes |
|---|---|---|
| Phylogenetic | $\mathbf A$ | Shared evolutionary history (tree → correlation) |
| Pedigree (animal model) | $\mathbf A$ | Genetic relatedness (pedigree → correlation) |
| Spatial | $\boldsymbol\Omega$ | Geographic proximity (distance + kernel) |

In abstract / cross-package teaching contexts (this article's prose), we write $\mathbf M$ for the generic role. In a fitted model with a known context (e.g. `drmTMB::phylo()` calls), `symbolizer` chooses the domain-specific letter — $\mathbf A$ for phylogenetic / animal-model, $\boldsymbol\Omega$ for spatial. The symbol-dictionary description for $\boldsymbol\Omega$ explicitly disambiguates it from (a) Wishart precision matrices and (b) CAR / SAR spatial weights matrices.

## Where the matrix comes from

For phylogenetic dependence under a Brownian-motion model of trait evolution, $\mathbf A$ is the **phylogenetic correlation matrix**:

$$
A_{ij} \;=\; \frac{T_{ij}}{T},
$$

where $T_{ij}$ is the shared branch length between species $i$ and $j$ measured from the root, and $T$ is the tree height. For an ultrametric tree all tips are equidistant from the root and $A_{ii} = 1$ by construction. The matrix is symmetric positive-definite.

Two operational representations exist:

```{r build-A, message = FALSE}
suppressPackageStartupMessages({
  library(metadat); library(ape); library(metafor); library(MCMCglmm)
})

# Real comparative-biology data: Mizuno et al. (2026, Research Synthesis
# Methods; tutorial at https://ayumi-495.github.io/phylo_spatial_tutorial/)
# use Moura et al. (2021)'s assortative-mating dataset for the phylogenetic
# example. dat.moura2021 carries 1,828 Pearson correlations (size-size
# assortative mating) across 341 species, plus a phylogeny. We follow the
# Mizuno workflow: compute Fisher-z effect sizes via metafor::escalc,
# subsample 60 species (each with at least 2 effect sizes for stable
# aggregation), and aggregate to one mean Fisher-z per species so the
# example is a clean phylogenetic random-intercept model on a Gaussian
# trait. The full multilevel-meta-analytic structure (effect-size +
# study + species_phylo + species_nonphylo random effects) is the
# v0.22 article's topic.
data(dat.moura2021, package = "metadat")
moura_es <- escalc(measure = "ZCOR", ri = ri, ni = ni,
                   data = dat.moura2021$dat)

# Tree tip labels carry Open Tree of Life format ("Genus_species_ottNNN");
# strip the suffix and convert underscores to spaces to match dat$species
# ("Genus species"). Drop node labels (duplicated in the source tree;
# MCMCglmm::inverseA rejects them otherwise).
tree_raw <- dat.moura2021$tree
tree_raw$tip.label <- gsub("_", " ", sub("_ott[0-9]+$", "", tree_raw$tip.label))
tree_raw$node.label <- NULL

n_per_sp <- table(moura_es$species)
sp_pool  <- intersect(tree_raw$tip.label,
                      names(n_per_sp[n_per_sp >= 2L]))
set.seed(1)
sp_keep  <- sample(sp_pool, 60L)

# Aggregate to one mean Zr per species + variance-of-the-mean for V.
agg <- aggregate(cbind(yi, vi) ~ species,
                 data = moura_es[moura_es$species %in% sp_keep, ],
                 FUN  = function(v) c(mean = mean(v), n = length(v)))
dat <- data.frame(
  species = factor(as.character(agg$species), levels = sp_keep),
  Zr      = agg$yi[, "mean"],
  vi      = (agg$vi[, "mean"] * agg$yi[, "n"]) / agg$yi[, "n"]^2,
  n_eff   = agg$yi[, "n"]
)

tree   <- keep.tip(tree_raw, sp_keep)
# Ultrametricise the tree once, here, so every downstream phylogenetic
# fit is on the SAME ultrametric tree. Brownian-motion on tips-only A
# requires equal root-to-tip distances by construction (A_ii = 1 only
# when tips are equidistant from the root); metafor / MCMCglmm / brms /
# glmmTMB accept a non-ultrametric tree silently but the BM prior is
# then misspecified. drmTMB ENFORCES ultrametricity (it errors otherwise),
# which surfaces the assumption that the other packages let pass. The
# Moura tree comes from Open Tree of Life with mixed-source branch
# lengths, so we smooth it with ape::chronos() (penalised-likelihood
# molecular-clock estimate; lambda = 1 = smoothest plausible clock).
tree   <- ape::chronos(tree, lambda = 1, quiet = TRUE)
class(tree) <- "phylo"
A_tips <- vcv.phylo(tree, corr = TRUE)
# Reorder A_tips rows/cols to match `sp_keep` (a random sample). Without
# this, A_tips inherits the depth-first tree-tip order, so head 5x5
# clusters closely-related species (all passerines, all ~0.85+) and
# hides the structural-dependence "mix of relatedness" the widget is
# meant to demonstrate. Florence-audit 2026-05-26.
A_tips <- A_tips[sp_keep, sp_keep]
# Representation 2 -- all-nodes A-inverse (tips + internal nodes, sparse
# precision; Hadfield & Nakagawa 2010). With the tree now ultrametric,
# scale = TRUE works cleanly (normalises total tree depth to 1, the
# canonical BM-prior parameterisation).
A_all  <- inverseA(tree, nodes = "ALL", scale = TRUE)

cat("species k =", nrow(dat),
    "  Zr range = [", round(min(dat$Zr), 2), ",", round(max(dat$Zr), 2), "]\n")
cat("A off-diagonal quantiles (25/50/75/99%): ",
    paste(round(quantile(A_tips[upper.tri(A_tips)], c(.25,.5,.75,.99)), 3),
          collapse = " / "), "\n")
```

The two representations encode the **same** Brownian-motion prior. The tips-only version is the operational default for `metafor`, `glmmTMB::propto`, and `brms::gr(cov = A)`. The all-nodes version is what `MCMCglmm` uses internally (`ginverse = list(animal = Ainv)`) and what `drmTMB::phylo()` builds under the hood (`drmTMB` cites Hadfield & Nakagawa 2010 explicitly in its `?phylo` help page). The all-nodes form is *richer*: it gives access to ancestral-state BLUPs at internal nodes; the tips-only form marginalises those out. In a balanced setup the two agree on the variance estimate $\hat\sigma_p^2$ — a result we'll return to in §"Tips vs all-nodes".

## Three packages, one phylogenetic LMM

We fit the same model three different ways on the 60-species Moura-derived dataset. The model is

$$
\mathrm{Zr}_i \;=\; \beta_0 + u_{p_{k[i]}} + e_i,\qquad
u_p \sim \mathcal N(\mathbf 0,\; \sigma_p^2\,\mathbf A),\quad
e_i \sim \mathcal N(0,\; \sigma_e^2),
$$

where:

| Symbol | Meaning | Status |
|---|---|---|
| $\mathrm{Zr}_i$ | Fisher-$z$ transformed correlation for species $i$ — here treated as an ordinary continuous trait | observed |
| $\beta_0$ | global mean of $\mathrm{Zr}$ across species | **estimated** |
| $u_{p_{k[i]}}$ | phylogenetic random effect, indexed by species $k[i]$ for observation $i$ | **estimated** (latent) |
| $\sigma_p^2$ | phylogenetic variance — the scale of $\mathbf u_p$ | **estimated** |
| $\mathbf A$ | $k \times k$ tips-only phylogenetic correlation matrix derived from the tree under Brownian motion (or its all-nodes augmentation; see §"Tips-only vs all-nodes") | constructed from tree |
| $e_i$ | residual on the response scale | **estimated** |
| $\sigma_e^2$ | residual variance — what each package would call $\sigma^2_{\text{units}}$, $\sigma^2_{\text{res}}$, or the marginal $\sigma^2 (1 - \lambda)$ depending on the parameterisation | **estimated** |

Three notes that often trip up first readers:

1. **$\sigma_e^2$ is estimated, not the per-effect sampling variance $v_i$.** In meta-analysis we'd also feed in $v_i$ (a *known* sampling variance per study) via `metafor::rma.mv(V = vi, ...)`. That's the model class in the v0.22 article — same $\sigma_p^2 \mathbf A$ phylogenetic term, but with $\mathbf V = \operatorname{diag}(v_i)$ added on top of $\sigma_e^2 \mathbf I$ (Cinar et al. 2021, Eq. 1–10).
2. The Moura data was *built* as a meta-analytic dataset (one Fisher-$z$ per species, aggregated from many effect-size estimates). We use only $\mathrm{Zr}$ here; the $v_i$ column is set aside. That's a pedagogical choice — a clean phylo-LMM is easier to read than a meta-analytic phylo-LMM, and it keeps this article focused on the structural-dependence axis.
3. `phylolm` fits a **marginalised** version of this same model: it absorbs $u_p$ into the residual to give $\mathrm{Zr} \sim \mathcal N(\beta_0 \mathbf 1, \sigma^2 \mathbf C(\alpha))$ where $\sigma^2 \mathbf C(\alpha) = \sigma_p^2 \mathbf A + \sigma_e^2 \mathbf I$ in the Pagel-$\lambda$ parametrisation. The §"Face 3" Face below makes the bridge explicit.

The phylogenetic correlation matrix $\mathbf A$ carries realistic biology — most species pairs are distantly related (median off-diagonal $\approx 0.03$) while a handful are close (top 1 % above 0.85). Compare with the degenerate `rcoal(15)` simulation used in earlier drafts where 75 % of pairs were above 0.91.

### Face 1 (deep dive): `MCMCglmm` with `ginverse = list(species = Ainv)`

`MCMCglmm` uses the **all-nodes** representation natively.

```{r mcmcglmm-fit, message = FALSE, results = "hide"}
fit_mcmc <- MCMCglmm(
  Zr ~ 1,
  random   = ~ species,
  ginverse = list(species = A_all$Ainv),
  data     = dat,
  family   = "gaussian",
  pr       = TRUE,        # store the per-species BLUPs for the widget
  nitt = 5000, burnin = 1000, thin = 5, verbose = FALSE
)
sym_mcmc <- symbolize(fit_mcmc, data = dat)
```

```{r mcmcglmm-meta}
sym_mcmc$metadata$phylo_representation   # "all_nodes"
sym_mcmc$metadata$detected_signals       # "phylo"
sym_mcmc$variance_components             # phylo + residual
sym_mcmc$metadata$heritability           # h^2 derived automatically
```

#### Three-views widget for the MCMCglmm fit

```{r mcmcglmm-widget, results = "asis", message = FALSE}
# All-nodes shim per Hadfield's course notes section 8.2.1:
# MCMCglmm fits with `ginverse = list(species = Ainv$Ainv)` where
# `inverseA(nodes = "ALL")` operates internally on the augmented
# (tips + internal nodes) parameterisation. For k tips the augmented
# random-effect vector has 2k - 2 entries (k tips + k - 1 internals -
# 1 root that inverseA drops). MCMCglmm stores BLUPs for ALL of these
# in `fit$Sol` when `pr = TRUE`. The Cov(u) prior is sigma^2 * A_all
# (dense (2k-2)x(2k-2)); Z is n_obs x (2k-2) with one-hot 1s only at
# the tip columns.
shim_mcmcglmm <- function(fit, dat, Ainv_obj) {
  node_order <- rownames(Ainv_obj$Ainv)         # internals first, then tips
  n_total    <- length(node_order)              # 2k - 2 for k tips
  beta_hat   <- mean(fit$Sol[, "(Intercept)"])
  sp_cols    <- grep("^species[.]", colnames(fit$Sol), value = TRUE)
  u_named    <- colMeans(fit$Sol[, sp_cols, drop = FALSE])
  names(u_named) <- sub("^species[.]", "", names(u_named))
  # Align to node_order (so u entries match Cov(u) rows and Z columns)
  u <- u_named[node_order]
  # Z: one-hot only at tip columns for the species each observation
  # belongs to; zeros for the n_internals - 1 internal-node columns
  # since internals have no direct observation
  Z_g <- matrix(0, nrow = nrow(dat), ncol = n_total,
                dimnames = list(NULL, node_order))
  for (i in seq_len(nrow(dat))) {
    Z_g[i, as.character(dat$species[i])] <- 1
  }
  X <- matrix(1, nrow = nrow(dat), ncol = 1,
              dimnames = list(NULL, "(Intercept)"))
  mu_hat <- as.numeric(X %*% beta_hat + Z_g %*% u)
  # Cov(u) = sigma^2 * A_all, where A_all is the dense inverse of the
  # sparse precision matrix that inverseA(nodes = "ALL") produces. This
  # is the (2k-2)x(2k-2) augmented covariance MCMCglmm operates on.
  M_dense <- as.matrix(Matrix::solve(Ainv_obj$Ainv))
  dimnames(M_dense) <- list(node_order, node_order)
  list(y = dat$Zr, X = X, beta = beta_hat,
       Z_g = Z_g, u = u, mu_hat = mu_hat,
       fitted = mu_hat, residuals = dat$Zr - mu_hat,
       M = M_dense)
}
sym_mcmc$expanded <- shim_mcmcglmm(fit_mcmc, dat, A_all)
as_html_three_views(sym_mcmc, id = "mcmc")
```

```{r mcmcglmm-pdf, results = "asis", message = FALSE}
# Helper -- write the PDF next to the rendered HTML so the `<a href>` link
# resolves whether the article is built by rmarkdown::render (vignettes
# dir) or pkgdown::build_article (docs/articles). knitr's
# `opts_knit$get("output.dir")` returns the Rmd source dir, NOT the HTML
# destination, so we emit one PDF in the chunk cwd (= source dir,
# where rmarkdown::render's link resolves) AND if a sibling
# `../docs/articles` exists (pkgdown build), copy a second copy there.
pdf_alongside_html <- function(sym, basename, title) {
  as_pdf_three_views(sym, file = basename, title = title)
  docs_articles <- "../docs/articles"
  if (dir.exists(docs_articles)) {
    file.copy(basename, file.path(docs_articles, basename), overwrite = TRUE)
  }
  cat(sprintf('<p><a href="%s" class="btn btn-primary">Download as PDF</a></p>\n', basename))
  invisible(basename)
}
pdf_alongside_html(sym_mcmc, "fig-mcmc-phylo.pdf",
                   "Phylogenetic phylo-LMM (MCMCglmm) -- three views")
```

The same extractor branch handles **animal models** (where `A` comes from a pedigree, not a tree) — see *§Animal-model unification* below.

### Face 2 (light): `brms::gr(species, cov = A)`

`brms` (Bürkner 2017) exposes the same structured-effect model via the `gr()` group factor inside an lme4-style RE bar. The fit takes ~30 s of Stan compile-and-sample (cached after the first build).

```{r brms-fit, message = FALSE, results = "hide"}
library(brms)
data2_list <- list(A = A_tips)
fit_brms <- brm(
  Zr ~ 1 + (1 | gr(species, cov = A)),
  data       = dat,
  data2      = data2_list,
  family     = gaussian(),
  chains     = 2, iter = 1000, refresh = 0, silent = 2,
  file       = "fig-brms-phylo-cache",       # brms auto-caches as .rds
  file_refit = "on_change"
)
sym_brms <- symbolize(fit_brms)
```

```{r brms-meta}
sym_brms$metadata$phylo_representation   # "tips_only"
sym_brms$metadata$detected_signals       # "phylo"
sym_brms$variance_components             # species + residual
```

`symbolizer` reports `phylo_representation = "tips_only"` because `brms` works directly with the $k \times k$ tips-only $\mathbf A$ — no internal-node augmentation. Compare with the MCMCglmm Face above where `phylo_representation = "all_nodes"`; the variance estimate $\hat\sigma_p^2$ is the same in expectation.

```{r brms-assumptions}
assumption_table(sym_brms)
```

### Face 3 (light): `phylolm::phylolm()` — the marginal PGLS form

`phylolm` (Ho & Ané 2014) fits the same conceptual model in a *marginalised* form: instead of writing $y = \beta_0 + u_p + e$ with two separate variance components, `phylolm` absorbs $u_p$ into the residual to give

$$
\mathrm{Zr}_i \;=\; \beta_0 + e'_i,\quad
\mathbf e' \sim \mathcal N(\mathbf 0,\; \sigma^2\,\mathbf C(\alpha))
$$

where $\mathbf C(\alpha)$ is a tree-derived correlation matrix. For Pagel's $\lambda$ model the connection to the RE form is precise:

$$
\mathbf C(\lambda) = \lambda\,\mathbf A + (1 - \lambda)\,\mathbf I,
\qquad
\sigma_p^2 = \lambda\,\sigma^2,
\qquad
\sigma_e^2 = (1 - \lambda)\,\sigma^2.
$$

So $\lambda = \sigma_p^2 / (\sigma_p^2 + \sigma_e^2)$ — the **phylogenetic heritability** $H^2$ from §"Animal-model unification" — and the two parameterisations agree on every prediction. `phylolm`'s advantage is speed: a closed-form ML fit in a fraction of a second instead of Stan's compile + sample.

```{r phylolm-fit, message = FALSE}
library(phylolm)
dat_pl <- dat
rownames(dat_pl) <- dat_pl$species
fit_pl <- phylolm(Zr ~ 1, data = dat_pl, phy = tree, model = "lambda")
sym_pl <- symbolize(fit_pl, data = dat_pl)
```

```{r phylolm-meta}
sym_pl$metadata$phylo_representation   # "pgls_marginal"
sym_pl$metadata$phylo_model            # "lambda"
sym_pl$metadata$phylo_param            # estimated lambda
sym_pl$metadata$detected_signals       # "phylo"
sym_pl$variance_components             # single sigma^2 row
```

```{r phylolm-summary}
summary(fit_pl)
```

The bridge with the RE form: take the estimated $\hat\sigma^2$ and $\hat\lambda$ from the summary above, then `sigma_p^2_hat = lambda_hat * sigma^2_hat` and `sigma_e^2_hat = (1 - lambda_hat) * sigma^2_hat`. These will be very close to the MCMCglmm / `brms` variance-component estimates (modulo MCMC sampling noise) — try it.

```{r phylolm-assumptions}
assumption_table(sym_pl)
```

The cross-package agreement story is the point: every fit above attaches the same conceptual object — a phylogenetic correlation matrix $\mathbf A$ — even though the syntactic surface differs (RE bar in `MCMCglmm` and `brms`, marginalised residual covariance in `phylolm`). `symbolizer` produces the same `metadata$detected_signals = "phylo"`, the same symbol-dictionary row for $\mathbf A$, the same gated phylogenetic assumption rows, no matter which parameterisation fitted the model.

## Animal-model unification

The phylogenetic random effect is the **same mathematical object** as the additive-genetic random effect in a quantitative-genetics animal model. The matrix $\mathbf A$ comes from a different source (a pedigree of dam-sire-offspring rather than a phylogenetic tree), but its meaning in the model is identical:

$$
\mathbf u_a \;\sim\; \mathcal N(\mathbf 0, \sigma_A^2\,\mathbf A),\qquad
h^2 \;=\; \frac{\sigma_A^2}{\sigma_A^2 + \sigma_e^2}.
$$

The only interpretation difference is what $\sigma_A^2$ *means*: phylogenetic variance among species in a phylogenetic comparative method, additive genetic variance among individuals in a quantitative-genetics animal model. `symbolizer`'s `symbolize.MCMCglmm()` handles both via the same `ginverse` detection branch, and `drmTMB` ships separate `phylo(term, tree)` and `animal(term, pedigree)` markers that share an internal implementation. For deep-dive treatments of either case, see drmTMB's own [`structural-dependence.Rmd`](https://itchyshin.github.io/drmTMB/articles/structural-dependence.html), [`phylogenetic-models.Rmd`](https://itchyshin.github.io/drmTMB/articles/phylogenetic-models.html), and [`animal-models.Rmd`](https://itchyshin.github.io/drmTMB/articles/animal-models.html) — `symbolizer` cross-links rather than duplicates them.

## Tips-only vs all-nodes

Why does the all-nodes representation matter when the tips-only correlation matrix has all the data information? Three reasons:

1. **Ancestral states**. The all-nodes form lets you query a posterior for the trait value at any internal node — useful for inferring ancestral character states.
2. **Numerical efficiency on large trees**. The all-nodes A-inverse is *sparse* (each internal node connects to its parent and children only). Inverting the tips-only $k \times k$ matrix is $\mathcal O(k^3)$; the sparse A-inverse path is much faster for $k > \sim 200$.
3. **Partial observations**. If some species have measured traits and others are unobserved (but on the tree), the all-nodes form handles this naturally.

On a balanced setup with all species observed, both representations give the same $\hat\sigma_p^2$ to machine precision — the Hadfield-Nakagawa equivalence.

## Spatial: same grammar, different matrix

For spatial dependence, $\mathbf M$ is built from pairwise geographic distances and a kernel. The standard choices are:

| Kernel | Formula | Decay |
|---|---|---|
| Exponential | $C(d) = \exp(-d/\rho)$ | Power |
| Squared-exponential (Gaussian) | $C(d) = \exp(-d^2/\rho^2)$ | Sharp |
| Matérn | $C(d; \kappa, \nu)$ | Tunable smoothness |

Two fits demonstrate the structurally identical grammar with $\boldsymbol\Omega$ in place of $\mathbf A$:

```{r spatial-show, eval = FALSE}
library(sdmTMB)
fit_sp <- sdmTMB(
  y ~ 1,
  data = dat_spatial,
  mesh = mesh,
  spatial = "on",
  family = gaussian()
)
sym_sp <- symbolize(fit_sp)
sym_sp$metadata$spatial_representation  # "package_managed"

library(mgcv)
fit_gp <- gam(y ~ s(x, y, bs = "gp"), data = dat_spatial)
sym_gp <- symbolize(fit_gp)
```

`symbolizer`'s metafor extractor also handles `rma.mv(struct = "SPEXP", dist = list(location = D))` — Mizuno et al. (2026, *Research Synthesis Methods*) is the conceptual anchor for the unified phylogenetic + spatial meta-analytic story.

## When unification breaks: identifiability

The unification works mathematically, but practical inference depends on the data. When you stack a phylogenetic tier and a non-phylogenetic species-level tier on the same grouping column,

$$
y_i = \beta_0 + u_{p_{k[i]}} + u_{s_{k[i]}} + e_i,\qquad
u_p \sim \mathcal N(0, \sigma_p^2 \mathbf A),\quad
u_s \sim \mathcal N(0, \sigma_s^2 \mathbf I),
$$

the variance components $\sigma_p^2$ and $\sigma_s^2$ are *weakly identified*: their sum is precisely estimable but the split between them is not (Hadfield & Nakagawa 2010 §3.2; Mizuno et al. 2026 §4). With strong phylogenetic signal (Pagel's λ near 1) the split is informative; with weak signal (λ near 0) the data have almost no leverage to separate the two. `symbolizer` emits a `phylo_nonphylo_unidentifiable` row in `warning_table()` when this configuration is detected. Profile on λ or fix one tier to resolve.

## Forward links

- **v0.22 — Phylogenetic meta-analysis (Mizuno et al. 2026)**. This article fits the general phylo-GLMM with $\sigma_e^2$ *estimated*. The meta-analytic case where the sampling variance $v_i$ is **known** per study is its own surface — `metafor::rma.mv(V = V, ...)` is the canonical interface, with `glmmTMB::equalto` (reserved) and `brms::se(sqrt(vi))` as bridges.
- **v0.24 — Location-scale on $\mathbf M$ (Nakagawa et al. 2025 *MEE*)**. The dependence parameter itself can have a submodel: $\sigma_p \sim z$ across clades, $\rho \sim z$ across environments. Symbolizer's drmTMB extractor is the right home for the phylogenetic location-scale model.
- **drmTMB deep dives**: `vignette("phylogenetic-models", package = "drmTMB")`, `vignette("animal-models", package = "drmTMB")`, `vignette("structural-dependence", package = "drmTMB")`.
- **gllvmTMB deep dive**: `vignette("phylogenetic-gllvm", package = "gllvmTMB")`.

## References

- Bürkner, P.-C. (2017). brms: An R Package for Bayesian multilevel models using Stan. *Journal of Statistical Software*, 80(1), 1–28.
- Cinar, O., Nakagawa, S. & Viechtbauer, W. (2022). Phylogenetic multilevel meta-analysis: A simulation study on the importance of modelling the phylogeny. *Methods in Ecology and Evolution*, 13, 383–395.
- Hadfield, J. D. (2010). MCMC methods for multi-response generalized linear mixed models: the `MCMCglmm` R package. *Journal of Statistical Software*, 33(2), 1–22.
- Hadfield, J. D. & Nakagawa, S. (2010). General quantitative genetic methods for comparative biology: phylogenies, taxonomies and multi-trait models for continuous and categorical characters. *Journal of Evolutionary Biology*, 23, 494–508.
- Ho, L. S. T. & Ané, C. (2014). A linear-time algorithm for Gaussian and non-Gaussian trait evolution models. *Systematic Biology*, 63(3), 397–408.
- Lynch, M. (1991). Methods for the analysis of comparative data in evolutionary biology. *Evolution*, 45(5), 1065–1080.
- Mizuno, A., Williams, C., Lagisz, M., Senior, A. M. & Nakagawa, S. (2026). A unified framework for phylogenetic and spatial meta-analysis: concepts, implementation, and practical guidance. *Research Synthesis Methods*.
- Nakagawa, S., Mizuno, A., Williams, C., Lagisz, M., Yang, Y. & Drobniak, S. M. (2025). Quantifying macro-evolutionary patterns of trait mean and variance with phylogenetic location-scale models. *Methods in Ecology and Evolution*, 16, 2585–2602.
- Williams, C., McGillycuddy, M., Drobniak, S. M., Bolker, B. M., Warton, D. I. & Nakagawa, S. (2025). Fast phylogenetic generalised linear mixed-effects modelling using the `glmmTMB` R package. *bioRxiv* 2025.12.20.695312.
