---
title: "Get started with symbolizer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with symbolizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_drmTMB <- requireNamespace("drmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_drmTMB
)
```

## Why structured symbolic models

[`equatiomatic`](https://datalorax.github.io/equatiomatic/) renders a fitted
model as a LaTeX string. That is enough when the only goal is a publishable
equation. `symbolizer` returns a richer object: a *structured symbolic model*
that other renderers can read. From the same object you can pull the equation,
the symbol dictionary, the assumption table, the formula-to-math bridge, and
the per-coefficient interpretation on link, natural, variance, and biological
scales.

The package is educator-first. It exposes equations in both index (scalar) and
matrix notation side by side, and it labels every assumption as *stated*,
*implied*, or *not checked* so a reader can see what the fit guarantees and
what it does not.

**Takeaway.** The product is the `symbolized_model` object; everything else is
a renderer of that object.

## A first `symbolize()` call

The canonical v0.1 example is a Gaussian location-scale model fit with
[`drmTMB`](https://itchyshin.github.io/drmTMB/): one submodel for the mean
(`mu`), one for the residual standard deviation (`sigma`).

```{r fit}
library(symbolizer)
library(drmTMB)

set.seed(1)
n <- 80
temperature <- runif(n, 10, 25)
dat <- data.frame(
  body_mass   = rnorm(n, 30 + 0.4 * temperature, exp(0.5 + 0.1 * temperature)),
  temperature = temperature
)

fit <- drmTMB(
  drm_formula(body_mass ~ temperature, sigma ~ temperature),
  family = gaussian(),
  data   = dat
)
```

`symbolize()` is the single entry point. Pass user-facing symbols and units so
the rendered equations carry biological meaning rather than R variable names:

```{r symbolize}
sym <- symbolize(
  fit,
  symbols = c(body_mass = "W_i", temperature = "T_i"),
  units   = c(body_mass = "g",   temperature = "C"),
  context = "avian body-size location-scale model"
)
```

**Takeaway.** One call gives you the structured object; the next sections
unpack what it carries.

## Equations in both notations

`equations()` returns one row per renderable block (conditional distribution,
each linear predictor, and any random-effect distribution). It always carries
both notations as columns; the `notation` argument only controls how the
result prints.

```{r equations}
equations(sym)
```

`as_latex()` produces the LaTeX string ready to splice into a document. With
`notation = "both"` the index and matrix forms are stacked in two `aligned`
blocks. Pasting the output into a math-aware renderer (the pkgdown site
emits MathML through pandoc; Quarto and most LaTeX renderers use KaTeX or
MathJax) shows them as proper math:

```{r as-latex-print}
cat(as_latex(sym, notation = "both"))
```

```{r as-latex-math, results = 'asis'}
cat("$$", as_latex(sym, notation = "both"), "$$", sep = "\n")
```

The matrix form follows a simple convention: bold lowercase is a vector
($\boldsymbol{\mu}$, $\boldsymbol{\beta}$), bold uppercase is a matrix
($\mathbf{X}$, $\mathbf{Z}$), and Greek vectors keep their Greek glyph.

`notation_bridge()` is the educator-facing table that pairs every piece of the
model with its counterpart in the other notation and tags the shape:

```{r notation-bridge}
notation_bridge(sym)
```

Read it as follows: the `dimension` column is the shape rule
(`\mathbb{R}^n`, `\mathbb{R}^{n \times p_\mu}`), and the `dimension_concrete`
column plugs in the actual sizes from this fit (`\mathbb{R}^{80}`,
`\mathbb{R}^{80 \times 2}`).

**Takeaway.** Both notations always coexist; the bridge teaches the reader to
move between them without translating by hand.

## Symbols, assumptions, formula bridge

`symbol_table()` lists every symbol that appears in the equations together
with its role, units, dimension, and description. Dimensions appear in both
abstract and concrete forms.

```{r symbol-table}
symbol_table(sym)
```

`assumption_table()` distinguishes three statuses: **stated** assumptions are
in the formula (distribution, link, linear predictor); **implied** assumptions
follow from the parameterisation (positivity of $\sigma$, conditional
independence); **not_checked** assumptions are still the user's responsibility
(missing-at-random).

```{r assumption-table}
assumption_table(sym)
```

`formula_bridge()` translates R syntax to mathematics. Each submodel has its
R formula on the left, a plain-English meaning, and the corresponding math in
both notations.

```{r formula-bridge}
formula_bridge(sym)
```

**Takeaway.** What the model says, what it assumes, and how its R syntax maps
to math live in three separate tables — so reviewers can audit each one
independently.

## Parameter interpretation

`parameter_interpretation()` reads each fixed-effect estimate on multiple
scales. By default (`scale = "all"`) it shows link, natural, variance (for
$\sigma$ coefficients), and biological readings together:

```{r parameter-interpretation}
parameter_interpretation(sym)
```

The `scale` argument filters to a single reading. The `"biological"` scale is
the one that goes in a results paragraph:

```{r parameter-interpretation-biological}
parameter_interpretation(sym, scale = "biological")
```

**Takeaway.** Every coefficient has an interpretation on every scale that
makes sense for its submodel; the package writes them so you do not have to.

## Random intercepts (first slice)

v0.1 supports Gaussian random intercepts of the form `(1 | group)` as a *first
slice*. The same `symbolize()` call works; the equation set picks up the
random-intercept term and a new distribution row, and the structured object
gains a `random_effects` and `variance_components` tibble.

```{r re-fit}
set.seed(2)
n        <- 80
n_groups <- 6
group    <- factor(rep(letters[seq_len(n_groups)], length.out = n))
temperature <- runif(n, 10, 25)
re <- rnorm(n_groups, sd = 1)
dat_re <- data.frame(
  body_mass = rnorm(
    n,
    30 + 0.4 * temperature + re[as.integer(group)],
    exp(0.5 + 0.05 * temperature)
  ),
  temperature = temperature,
  group       = group
)

fit_re <- drmTMB(
  drm_formula(body_mass ~ temperature + (1 | group), sigma ~ temperature),
  family = gaussian(),
  data   = dat_re
)

sym_re <- symbolize(
  fit_re,
  symbols = c(body_mass = "W_i", temperature = "T_i"),
  units   = c(body_mass = "g",   temperature = "C")
)
```

The mu linear predictor now carries a $+ u_{group(i)}$ term, and a new
distributional row appears for the random intercepts:

```{r re-equations}
equations(sym_re)
```

```{r re-latex, results = 'asis'}
cat("$$", as_latex(sym_re, notation = "both"), "$$", sep = "\n")
```

The grouping structure is exposed as tidy tibbles, ready to feed into reports
or other renderers:

```{r re-tables}
sym_re$random_effects
sym_re$variance_components
```

**Takeaway.** The same object carries everything a reader needs to discuss
random structure — what is grouped, how many levels, and which variance
component goes with which submodel.

## What's supported, what's planned

`symbolizer_capabilities()` is the registry that gates `symbolize()`. Each row
carries one of five status words: **Stable**, **First slice**, **Opt-in
control**, **Planned or reserved**, or **Unsupported**. v0.1 marks
`drmTMB` Gaussian `mu` and `sigma` as Stable, Gaussian `random_effects` as
First slice, and reserves the rest of the matrix for later versions.

```{r capabilities}
symbolizer_capabilities()
```

The roadmap in `README.md` lists the planned version per family / package. If
`symbolize()` errors with a capability message, the registry will tell you
which version is scheduled to lift it.

**Takeaway.** The registry is the single source of truth for "what works".
Read the status word before designing around an unsupported tuple.
