---
title: "Understanding a drmTMB fit with symbolizer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding a drmTMB fit with symbolizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_drmTMB <- requireNamespace("drmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_drmTMB
)
```

> *symbolizer does not just print equations for drmTMB; it makes a fitted
> distributional model auditable by connecting the R formula, symbolic
> model, parameter scales, assumptions, coefficient readings, and next
> diagnostic steps.*

This vignette is the drmTMB-first tour. It assumes you have a `drmTMB`
fit on the table and you want to know exactly what it claims, what it
assumes, and what to look at next. If you want the package-wide tour
first, read `vignette("symbolizer")`. The richer worked example with
factors, transformations, and a multi-level random intercept now lives
in Section 6 of this same vignette.

## 1. Why distributional models need a structured story

`drmTMB` fits Gaussian location-scale models — and, since v0.2,
bivariate Gaussian location-scale-correlation models (Student-t,
lognormal, gamma, and beta remain on the roadmap). A univariate
location-scale fit has *two* linear predictors: one for the mean `mu`,
one for the residual standard deviation `sigma`. Both depend on
covariates. Both have a link function. Both have their own coefficient
vector.

That structure makes the question *what does this fit actually say?*
genuinely hard to answer in one line. The mean coefficient on
temperature lives on the response scale. The sigma coefficient on
temperature lives on the log scale. The residual SD is no longer a
single number; it is a vector with one entry per observation. A reader
who only sees `summary(fit)` has to keep all of that in their head while
reading the table.

`symbolizer` returns a `symbolized_model`: a structured object that
connects the R formula to a symbolic equation, to assumptions, to
per-coefficient readings on every relevant scale, and to a short list of
diagnostic next steps. Every renderer in the package consumes that same
object. No prose is generated by a language model at runtime; every
phrase comes from a template in `inst/extdata/`.

**Takeaway.** A drmTMB fit has more structure than a single regression
table. The `symbolized_model` is what you read instead of holding it all
in your head.

## 2. From `drm_formula()` to equations

A small fit, by way of example. `body_mass` depends on `temperature` on
both submodels:

```{r fit}
library(symbolizer)
library(drmTMB)

set.seed(20260523)
n <- 120
temperature <- runif(n, 10, 25)
dat <- data.frame(
  body_mass   = rnorm(n, 30 + 0.4 * temperature, exp(0.5 + 0.08 * temperature)),
  temperature = temperature
)

fit <- drmTMB(
  drm_formula(body_mass ~ temperature, sigma ~ temperature),
  family = gaussian(),
  data   = dat
)
```

The two arguments to `drm_formula()` map directly to two linear
predictors. `body_mass ~ temperature` defines the mu submodel;
`sigma ~ temperature` defines the sigma submodel. `symbolize()` reads
the fit and returns the structured object:

```{r symbolize}
sym <- symbolize(
  fit,
  symbols = c(body_mass = "W_i", temperature = "T_i"),
  units   = c(body_mass = "g",   temperature = "C"),
  context = "Gaussian location-scale body-size fit"
)
```

`equations(sym)` exposes the model as one row per renderable block. The
`notation = "both"` argument controls how the result prints; both
notations always live in the object.

```{r equations}
equations(sym, notation = "both")
```

In math, the index form is

$$W_i \mid \mu_i,\, \sigma_i \sim \mathrm{Normal}(\mu_i,\, \sigma_i^2),$$
$$\mu_i = \beta_0 + \beta_1\, T_i, \qquad \log(\sigma_i) = \gamma_0 + \gamma_1\, T_i,$$

and the matrix form is

$$\mathbf{w} \mid \boldsymbol{\mu},\, \boldsymbol{\sigma}
  \sim \mathcal{N}(\boldsymbol{\mu},\, \mathrm{diag}(\boldsymbol{\sigma}^2)),$$
$$\boldsymbol{\mu} = \mathbf{X}\,\boldsymbol{\beta}, \qquad
  \log(\boldsymbol{\sigma}) = \mathbf{Z}\,\boldsymbol{\gamma}.$$

Each piece of the R call produced one piece of math. The left half of
`drm_formula()` (`body_mass ~ temperature`) became the mu line; the
right half (`sigma ~ temperature`) became the log-sigma line; the
`family = gaussian()` choice became the distribution line and pinned
the identity link on mu.

**Takeaway.** Each argument to `drm_formula()` produces one line of
math. The mapping is one-to-one; nothing is implicit.

## 3. What sigma means in a location-scale model

In an introductory regression, sigma is a scalar — the residual SD,
estimated once for the whole dataset. In a `drmTMB` location-scale fit
it is a *vector*: each observation has its own predicted SD, given by
a linear predictor on the log scale.

That choice is not cosmetic. It changes the units of every coefficient
in the sigma submodel. `drmTMB` locks in the log link on sigma, so

$$\log(\sigma_i) = \gamma_0 + \gamma_1 T_i \quad\Longleftrightarrow\quad
  \sigma_i = \exp(\gamma_0)\,\exp(\gamma_1 T_i).$$

Two readings follow.

- $\gamma_0$ is $\log(\sigma)$ at the reference (here, $T_i = 0$).
  $\exp(\gamma_0)$ is the residual SD itself at that reference value.
- $\gamma_1$ is the change in $\log(\sigma)$ per one-unit change in
  temperature. The biological reading is multiplicative: a one-unit
  change in temperature multiplies the residual SD by $\exp(\gamma_1)$.
  If $\gamma_1 > 0$, hotter sites are more variable; if
  $\gamma_1 < 0$, they are less variable.

`parameter_interpretation()` reads each coefficient on every relevant
scale at once:

```{r parameter-interpretation}
parameter_interpretation(sym)
```

The `variance_scale_reading` column is the one that does the work for
sigma coefficients: it states the multiplicative effect on the residual
SD itself, with the fitted value of $\exp(\hat{\gamma}_1)$ already
substituted in.

**Takeaway.** Each observation has its own predicted sigma. The
sigma coefficients live on the log scale; `exp()` of the slope is the
multiplicative factor on the residual SD.

## 4. What is and isn't assumed

A reader auditing the fit needs to know which assumptions are in the
formula, which follow automatically from the parameterisation, and
which are still the user's responsibility. `assumption_table()` splits
them into three buckets:

```{r assumption-table}
assumption_table(sym)
```

The three statuses map onto three honest questions.

- **`stated`** — the formula declares this. The Gaussian distribution
  on $W_i$, the identity link on mu, the log link on sigma, and the
  linear-predictor structure of each submodel are all *stated*: they
  are what `drm_formula()` and `family = gaussian()` mean. There is
  nothing for the user to verify; this is what the model *is*.
- **`implied`** — these follow from the parameterisation, not from
  data. The log link on sigma forces $\sigma_i > 0$ automatically;
  conditional independence given the predictors is the default of any
  fixed-effects regression. The package surfaces these so they are not
  invisible, but they are not separate empirical claims.
- **`not_checked`** — these are the user's responsibility. Missing-at-
  random, for example, cannot be inferred from the fit; `symbolizer`
  states it as an assumption the analyst still owns.

Three things `symbolizer` deliberately does *not* check. Residual
normality on mu. Homoscedasticity *within* the sigma model (the model
allows sigma to vary; whether the fitted sigma actually absorbs all the
heteroscedasticity is a diagnostic question, not a symbolic one).
Missing-at-random. The first two belong in
section 6 (diagnostics); the third is documented in the assumption
table itself.

**Takeaway.** `assumption_table()` is the audit lens. Stated and
implied are guarantees of the parameterisation; not_checked is on the
analyst.

## 5. Random intercepts and the assumption swap

When a random intercept enters the mu submodel, the independence
statement automatically updates. `symbolizer` swaps the unconditional
"observations are conditionally independent given the predictors" row
for the conditional "observations are conditionally independent given
the predictors **and** the random effects" row.

A small fit demonstrates:

```{r re-fit}
set.seed(20260523)
n <- 120
n_sites <- 6
site <- factor(rep(letters[seq_len(n_sites)], length.out = n))
temperature <- runif(n, 10, 25)
site_re <- rnorm(n_sites, sd = 1.0)
dat_re <- data.frame(
  body_mass   = rnorm(
    n,
    30 + 0.4 * temperature + site_re[as.integer(site)],
    exp(0.5 + 0.05 * temperature)
  ),
  temperature = temperature,
  site        = site
)

fit_re <- drmTMB(
  drm_formula(body_mass ~ temperature + (1 | site), sigma ~ temperature),
  family = gaussian(),
  data   = dat_re
)

sym_re <- symbolize(
  fit_re,
  symbols = c(body_mass = "W_i", temperature = "T_i"),
  units   = c(body_mass = "g",   temperature = "C")
)
```

The mu equation picks up a $+ u_{\mathrm{site}(i)}$ term, and a new
random-effect distribution row appears in the equation block:

```{r re-equations}
equations(sym_re, notation = "both")
```

The assumption table swaps the independence row. Side by side:

```{r re-assumptions}
# Without random effects
assumption_table(sym)[assumption_table(sym)$assumption %in%
                       c("independence", "independence_given_random_effects"), ]
# With (1 | site)
assumption_table(sym_re)[assumption_table(sym_re)$assumption %in%
                          c("independence", "independence_given_random_effects"), ]
```

The structured object also exposes the random-effect structure as tidy
tibbles, ready for downstream renderers:

```{r re-tables}
sym_re$random_effects
sym_re$variance_components
```

`sd(site)` is the between-site standard deviation; the symbol table
reports it as $\sigma_{\mathrm{site}}$.

**Takeaway.** Adding `(1 | group)` automatically reworks the
independence assumption from unconditional to conditional. The audit
trail is built into the template, not into prose.

## 6. A richer worked example: four predictor roles in one fit

The previous sections walked the basics on a simple two-predictor fit.
A real model an ecologist would run usually has more going on:
continuous predictors, log-transformed predictors, factor contrasts,
and a random intercept on the mean. This section shows that the same
`symbolize()` surfaces scale up — with two educator-facing surfaces
(`formula_bridge` and `notation_bridge`) that get more interesting
when the fit is more interesting.

```{r rich-fit}
set.seed(20260523)
nr           <- 200
temperature  <- runif(nr, 8, 30)
food         <- rlnorm(nr, meanlog = 1.5, sdlog = 0.6)
sex          <- factor(sample(c("female", "male"), nr, replace = TRUE))
site         <- factor(sample(letters[1:8], nr, replace = TRUE))

site_re <- rnorm(8, sd = 1.5)
mu_r    <- 5 + 0.5 * temperature + 2 * log(food) +
             3 * (sex == "male") + site_re[site]
sg_r    <- exp(0.3 + 0.05 * temperature + 0.1 * (sex == "male"))
dat_r   <- data.frame(
  body_mass = rnorm(nr, mu_r, sg_r),
  temperature = temperature, food = food, sex = sex, site = site
)

fit_rich <- drmTMB(
  drm_formula(
    body_mass ~ temperature + log(food) + sex + (1 | site),
    sigma ~ temperature + sex
  ),
  family = gaussian(),
  data   = dat_r
)
sym_rich <- symbolize(
  fit_rich,
  symbols = c(body_mass = "W_i", temperature = "T_i",
              food      = "F_i", sex         = "S_i"),
  units   = c(body_mass = "g",   temperature = "C", food = "g/day"),
  context = "structured body-size location-scale model"
)
```

`formula_bridge()` is the educator-facing translation table from R
formula syntax to mathematics. Two rows (one per submodel), in both
notations:

```{r rich-formula-bridge}
formula_bridge(sym_rich)
```

For the mu submodel, the R syntax
`body_mass ~ temperature + log(food) + sex + (1 | site)` maps to the
indexed equation in the `mathematics` column and to
$\boldsymbol{\mu} = \mathbf{X} \boldsymbol{\beta} + \mathbf{u}$ in the
`mathematics_matrix` column. The random-intercept piece appears as
$+ \mathbf{u}$ in the matrix form and as $+ u_{\mathrm{site}(i)}$ in
the indexed form. A reader can compare the two columns and learn the
translation without doing it by hand.

`notation_bridge()` is the deeper teaching surface: every symbol *and*
every equation pairs its index form with its matrix form and tags
both shapes:

```{r rich-notation-bridge}
notation_bridge(sym_rich)
```

The `dimension` column carries the shape rule (e.g. $\mathbb{R}^n$,
$\mathbb{R}^{n \times p_\mu}$). The `dimension_concrete` column plugs
in the actual numbers from this fit (e.g. $\mathbb{R}^{200}$,
$\mathbb{R}^{200 \times 4}$). A reader learning matrix notation can
stare at this table and see, in one place, that the
$\boldsymbol{\mu}$ vector has 200 entries, the $\boldsymbol{\beta}$
vector has four, and $\mathbf{X}$ is the $200 \times 4$ matrix that
connects them.

**Takeaway.** The same renderers (`equations`, `formula_bridge`,
`notation_bridge`) scale from a two-predictor toy fit to a four-role
mixed model. Abstract shape rules on one side, concrete numbers from
this fit on the other.

## 7. Two responses at once: bivariate Gaussian

When you measure two responses on the same subject — paired growth
and reproduction, two morphometric traits, a behaviour and its
physiological correlate — a univariate fit fits each response in
isolation and ignores the residual correlation between them.
`drmTMB`'s `biv_gaussian()` family models both responses jointly,
with a residual correlation $\rho_{12}$ that itself can depend on
covariates.

```{r biv-fit}
set.seed(20260524)
nb <- 80
x1 <- rnorm(nb)
x2 <- rnorm(nb)
# Cholesky factorisation for two correlated standard normals (rho = 0.6),
# no extra package needed:
e1 <- rnorm(nb)
e2 <- 0.6 * e1 + sqrt(1 - 0.6^2) * rnorm(nb)
dat_biv <- data.frame(
  growth = 30 + 1.5 * x1 + e1,
  repro  = 10 + 0.8 * x2 + e2,
  x1     = x1,
  x2     = x2
)

fit_biv <- drmTMB(
  drm_formula(
    mu1    = growth ~ x1,
    mu2    = repro  ~ x2,
    sigma1 = ~ 1,
    sigma2 = ~ 1,
    rho12  = ~ 1
  ),
  family = biv_gaussian(),
  data   = dat_biv
)

sym_biv <- symbolize(fit_biv)
```

The `symbolized_model` carries five submodels instead of two. Each
appears in `formula_bridge()` with its own R syntax, statistical
meaning, and symbolic form:

```{r biv-formula-bridge}
formula_bridge(sym_biv)
```

`mu1`, `mu2` are the two mean predictors; `sigma1`, `sigma2` are the
two log-SD predictors (one residual SD per response); `rho12` is the
residual correlation on the Fisher-z scale (so the linear predictor
can range over all of $\mathbb{R}$ while the correlation stays in
$(-1, 1)$).

The joint conditional distribution is rendered in both index and matrix
forms:

```{r biv-equations}
equations(sym_biv, notation = "both")
```

The matrix form is the cleaner reading: $\mathbf{Y}$ is the $n \times 2$
response matrix, $\mathbf{M}$ stacks the per-observation mean vectors,
and $\boldsymbol{\Sigma}$ is the per-observation $2 \times 2$ covariance
matrix with $\sigma_{1,i}\,\sigma_{2,i}\,\rho_{12,i}$ on the
off-diagonal.

The rho12 coefficient gets the same multi-scale treatment as every
other coefficient. The reading row points the reader at the Fisher-z
scale (where the coefficient lives), the natural correlation scale
(`tanh()` of the link), and the biological meaning (residual co-variation
between the two responses):

```{r biv-rho-interpretation}
rho_rows <- parameter_interpretation(sym_biv)
rho_rows[rho_rows$submodel == "rho12", ]
```

**Takeaway.** The same `symbolized_model` interface scales from two
submodels to five. `formula_bridge` shows the new submodels, the joint
distribution exposes $\boldsymbol{\Sigma}$, and
`parameter_interpretation` gives the rho12 row the same multi-scale
reading as every other coefficient.

## 8. What to inspect next

The point of the structured object is not just to render equations; it
is to tell the reader *what to look at next*. `model_card(sym)`
packages the full teaching bundle in one object, including a class-
specific list of extraction calls. For a `drmTMB` fit, the extraction
calls are ordered diagnostics-first:

```{r model-card-calls}
mc <- model_card(sym)
mc$extraction_calls
```

The ordering is intentional. A reader should know whether the fit is
trustworthy before they interpret a single coefficient.

1. **`drmTMB::check_drm(fit)`** — convergence and identifiability.
   Did the optimiser converge? Are any parameters at the boundary? Are
   the standard errors finite?
2. **`residuals(fit)`** and **`predict(fit)`** — residual diagnostics.
   Plot residuals against the mu predictors (any structure left in the
   mean?) and absolute residuals against the sigma predictors (did the
   sigma submodel absorb the heteroscedasticity?).
3. **`drmTMB::ranef(fit)`** — random-effect BLUPs, when there are
   any. The distribution of BLUPs is worth a histogram or a QQ plot.
4. **`drmTMB::fixef(fit, dpar = "mu")`** and **`drmTMB::fixef(fit, dpar = "sigma")`**
   — the actual coefficient tables, by submodel. Read these *after*
   the diagnostics, not before.

`model_card()` also ships a set of recommended-plot recipes, in the
same diagnostics-first order:

```{r model-card-plots}
mc$recommended_plots
```

These are text recipes only -- `symbolizer` does not draw the plots
itself. The recipes name the diagnostic plot the reader should make
next, in their own plotting framework of choice.

**Takeaway.** Diagnostics before coefficients. The extraction-calls
list is the order a reader should work through the fit.

## 9. Drafting the Methods paragraph

Once you've inspected the fit and you're ready to write the paper,
`methods_text(sym)` returns a draft Methods-section paragraph that
matches what your model actually does. The prose is composed from a
CSV template plus slots filled from the `symbolized_model` — there's
no LLM in the loop, so every phrase traces to the template row.

```{r methods-text}
methods_text(sym)
```

Two things to notice in the output:

- The paragraph names the response, the predictors per submodel, the
  random-effect group (when present), the package and version, the
  number of observations, and the CI method. Each of those traces
  back to a slot you can inspect on `methods_text(sym)$slots`.
- The print method appends *Reminders before you paste* — a short
  checklist that nudges you to confirm the package version, soften
  or sharpen the prose, and reword the random-effect sentence if
  you used a structure more complex than `(1 | group)`.

The same call works for the richer fit from Section 6 (a 4-role
location-scale model with a random intercept):

```{r methods-text-rich}
methods_text(sym_rich)
```

and for the bivariate-Gaussian fit from Section 7:

```{r methods-text-biv}
methods_text(sym_biv)
```

The biv_gaussian template carries dedicated slots for `mu1`, `mu2`,
`sigma1`, `sigma2`, and `rho12` so the paragraph names both
responses and reports the Fisher-z scale on the residual
correlation.

**Takeaway.** `methods_text(sym)` produces a draft, not a final
paragraph. Treat the output as starting material to edit; the
reminders block is the editorial checklist.

## 10. The capability gate

`symbolize()` does not silently dispatch on every `drmTMB` fit. The
capability registry gates each `(class, family, component)` tuple with
one of five status words: **Stable**, **First slice**, **Opt-in
control**, **Planned or reserved**, **Unsupported or blocked**.

```{r capabilities}
caps <- symbolizer_capabilities()
caps[caps$class == "drmTMB", ]
```

The relevant rows are:

- `drmTMB / gaussian / mu` — **Stable**
- `drmTMB / gaussian / sigma` — **Stable**
- `drmTMB / gaussian / random_effects` — **First slice**
  (intercept-only `(1 | group)`)
- `drmTMB / biv_gaussian / mu1`, `mu2`, `sigma1`, `sigma2`, `rho12` —
  **First slice** (see Section 6)
- All non-Gaussian families — **Planned or reserved**

If you call `symbolize()` on an unsupported tuple, the package errors
with a friendly *"Today we can read…"* message that names the row in
the registry. The registry, not the prose, decides what `symbolize()`
accepts; check the status word before designing a workflow around a
tuple.

**Takeaway.** The capability registry is the single source of truth.
For `drmTMB` it covers the Gaussian location-scale family (including
random intercepts and random slopes), the full bivariate Gaussian
surface (`mu1`, `mu2`, `sigma1`, `sigma2`, `rho12`), the Student-t /
lognormal / Gamma / beta / beta_binomial / Poisson / nbinom2 /
truncated_nbinom2 / cumulative_logit families, and the zero-inflation
and hurdle submodels. See `vignette("symbolizer-roadmap")` for the
full matrix.

## 11. Where to read next

- `vignette("symbolizer")` — the package-wide Get Started tour.
- `vignette("symbolizer-factors")` — the categorical pedagogy guide:
  dummies, contrasts, interactions, and six common pitfalls.
- `vignette("symbolizer-gllvm")` — the parallel story for
  latent-variable models in `gllvmTMB`.
- `vignette("symbolizer-compare")` — comparing two symbolized models
  with `compare_symbolic()`.
