---
title: "Understanding a drmTMB fit with symbolizer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding a drmTMB fit with symbolizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_drmTMB <- requireNamespace("drmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_drmTMB
)
```

> *symbolizer does not just print equations for drmTMB; it makes a fitted
> distributional model auditable by connecting the R formula, symbolic
> model, parameter scales, assumptions, coefficient readings, and next
> diagnostic steps.*

This vignette is the drmTMB-first tour. It assumes you have a `drmTMB`
fit on the table and you want to know exactly what it claims, what it
assumes, and what to look at next. If you want the package-wide tour
first, read `vignette("symbolizer")`. If you want a richer worked
example with multiple predictor types, read `vignette("symbolizer-large")`.

## 1. Why distributional models need a structured story

`drmTMB` fits Gaussian location-scale models (and, on the roadmap,
Student-t, lognormal, gamma, beta, and bivariate Gaussian). A
location-scale fit has *two* linear predictors: one for the mean `mu`,
one for the residual standard deviation `sigma`. Both depend on
covariates. Both have a link function. Both have their own coefficient
vector.

That structure makes the question *what does this fit actually say?*
genuinely hard to answer in one line. The mean coefficient on
temperature lives on the response scale. The sigma coefficient on
temperature lives on the log scale. The residual SD is no longer a
single number; it is a vector with one entry per observation. A reader
who only sees `summary(fit)` has to keep all of that in their head while
reading the table.

`symbolizer` returns a `symbolized_model`: a structured object that
connects the R formula to a symbolic equation, to assumptions, to
per-coefficient readings on every relevant scale, and to a short list of
diagnostic next steps. Every renderer in the package consumes that same
object. No prose is generated by a language model at runtime; every
phrase comes from a template in `inst/extdata/`.

**Takeaway.** A drmTMB fit has more structure than a single regression
table. The `symbolized_model` is what you read instead of holding it all
in your head.

## 2. From `drm_formula()` to equations

A small fit, by way of example. `body_mass` depends on `temperature` on
both submodels:

```{r fit}
library(symbolizer)
library(drmTMB)

set.seed(20260523)
n <- 120
temperature <- runif(n, 10, 25)
dat <- data.frame(
  body_mass   = rnorm(n, 30 + 0.4 * temperature, exp(0.5 + 0.08 * temperature)),
  temperature = temperature
)

fit <- drmTMB(
  drm_formula(body_mass ~ temperature, sigma ~ temperature),
  family = gaussian(),
  data   = dat
)
```

The two arguments to `drm_formula()` map directly to two linear
predictors. `body_mass ~ temperature` defines the mu submodel;
`sigma ~ temperature` defines the sigma submodel. `symbolize()` reads
the fit and returns the structured object:

```{r symbolize}
sym <- symbolize(
  fit,
  symbols = c(body_mass = "W_i", temperature = "T_i"),
  units   = c(body_mass = "g",   temperature = "C"),
  context = "Gaussian location-scale body-size fit"
)
```

`equations(sym)` exposes the model as one row per renderable block. The
`notation = "both"` argument controls how the result prints; both
notations always live in the object.

```{r equations}
equations(sym, notation = "both")
```

In math, the index form is

$$W_i \mid \mu_i,\, \sigma_i \sim \mathrm{Normal}(\mu_i,\, \sigma_i^2),$$
$$\mu_i = \beta_0 + \beta_1\, T_i, \qquad \log(\sigma_i) = \gamma_0 + \gamma_1\, T_i,$$

and the matrix form is

$$\mathbf{w} \mid \boldsymbol{\mu},\, \boldsymbol{\sigma}
  \sim \mathcal{N}(\boldsymbol{\mu},\, \mathrm{diag}(\boldsymbol{\sigma}^2)),$$
$$\boldsymbol{\mu} = \mathbf{X}\,\boldsymbol{\beta}, \qquad
  \log(\boldsymbol{\sigma}) = \mathbf{Z}\,\boldsymbol{\gamma}.$$

Each piece of the R call produced one piece of math. The left half of
`drm_formula()` (`body_mass ~ temperature`) became the mu line; the
right half (`sigma ~ temperature`) became the log-sigma line; the
`family = gaussian()` choice became the distribution line and pinned
the identity link on mu.

**Takeaway.** Each argument to `drm_formula()` produces one line of
math. The mapping is one-to-one; nothing is implicit.

## 3. What sigma means in a location-scale model

In an introductory regression, sigma is a scalar — the residual SD,
estimated once for the whole dataset. In a `drmTMB` location-scale fit
it is a *vector*: each observation has its own predicted SD, given by
a linear predictor on the log scale.

That choice is not cosmetic. It changes the units of every coefficient
in the sigma submodel. `drmTMB` locks in the log link on sigma, so

$$\log(\sigma_i) = \gamma_0 + \gamma_1 T_i \quad\Longleftrightarrow\quad
  \sigma_i = \exp(\gamma_0)\,\exp(\gamma_1 T_i).$$

Two readings follow.

- $\gamma_0$ is $\log(\sigma)$ at the reference (here, $T_i = 0$).
  $\exp(\gamma_0)$ is the residual SD itself at that reference value.
- $\gamma_1$ is the change in $\log(\sigma)$ per one-unit change in
  temperature. The biological reading is multiplicative: a one-unit
  change in temperature multiplies the residual SD by $\exp(\gamma_1)$.
  If $\gamma_1 > 0$, hotter sites are more variable; if
  $\gamma_1 < 0$, they are less variable.

`parameter_interpretation()` reads each coefficient on every relevant
scale at once:

```{r parameter-interpretation}
parameter_interpretation(sym)
```

The `variance_scale_reading` column is the one that does the work for
sigma coefficients: it states the multiplicative effect on the residual
SD itself, with the fitted value of $\exp(\hat{\gamma}_1)$ already
substituted in.

**Takeaway.** Each observation has its own predicted sigma. The
sigma coefficients live on the log scale; `exp()` of the slope is the
multiplicative factor on the residual SD.

## 4. What is and isn't assumed

A reader auditing the fit needs to know which assumptions are in the
formula, which follow automatically from the parameterisation, and
which are still the user's responsibility. `assumption_table()` splits
them into three buckets:

```{r assumption-table}
assumption_table(sym)
```

The three statuses map onto three honest questions.

- **`stated`** — the formula declares this. The Gaussian distribution
  on $W_i$, the identity link on mu, the log link on sigma, and the
  linear-predictor structure of each submodel are all *stated*: they
  are what `drm_formula()` and `family = gaussian()` mean. There is
  nothing for the user to verify; this is what the model *is*.
- **`implied`** — these follow from the parameterisation, not from
  data. The log link on sigma forces $\sigma_i > 0$ automatically;
  conditional independence given the predictors is the default of any
  fixed-effects regression. The package surfaces these so they are not
  invisible, but they are not separate empirical claims.
- **`not_checked`** — these are the user's responsibility. Missing-at-
  random, for example, cannot be inferred from the fit; `symbolizer`
  states it as an assumption the analyst still owns.

Three things `symbolizer` deliberately does *not* check. Residual
normality on mu. Homoscedasticity *within* the sigma model (the model
allows sigma to vary; whether the fitted sigma actually absorbs all the
heteroscedasticity is a diagnostic question, not a symbolic one).
Missing-at-random. The first two belong in
section 6 (diagnostics); the third is documented in the assumption
table itself.

**Takeaway.** `assumption_table()` is the audit lens. Stated and
implied are guarantees of the parameterisation; not_checked is on the
analyst.

## 5. Random intercepts and the assumption swap

When a random intercept enters the mu submodel, the independence
statement automatically updates. `symbolizer` swaps the unconditional
"observations are conditionally independent given the predictors" row
for the conditional "observations are conditionally independent given
the predictors **and** the random effects" row.

A small fit demonstrates:

```{r re-fit}
set.seed(20260523)
n <- 120
n_sites <- 6
site <- factor(rep(letters[seq_len(n_sites)], length.out = n))
temperature <- runif(n, 10, 25)
site_re <- rnorm(n_sites, sd = 1.0)
dat_re <- data.frame(
  body_mass   = rnorm(
    n,
    30 + 0.4 * temperature + site_re[as.integer(site)],
    exp(0.5 + 0.05 * temperature)
  ),
  temperature = temperature,
  site        = site
)

fit_re <- drmTMB(
  drm_formula(body_mass ~ temperature + (1 | site), sigma ~ temperature),
  family = gaussian(),
  data   = dat_re
)

sym_re <- symbolize(
  fit_re,
  symbols = c(body_mass = "W_i", temperature = "T_i"),
  units   = c(body_mass = "g",   temperature = "C")
)
```

The mu equation picks up a $+ u_{\mathrm{site}(i)}$ term, and a new
random-effect distribution row appears in the equation block:

```{r re-equations}
equations(sym_re, notation = "both")
```

The assumption table swaps the independence row. Side by side:

```{r re-assumptions}
# Without random effects
assumption_table(sym)[assumption_table(sym)$assumption %in%
                       c("independence", "independence_given_random_effects"), ]
# With (1 | site)
assumption_table(sym_re)[assumption_table(sym_re)$assumption %in%
                          c("independence", "independence_given_random_effects"), ]
```

The structured object also exposes the random-effect structure as tidy
tibbles, ready for downstream renderers:

```{r re-tables}
sym_re$random_effects
sym_re$variance_components
```

`sd(site)` is the between-site standard deviation; the symbol table
reports it as $\sigma_{\mathrm{site}}$.

**Takeaway.** Adding `(1 | group)` automatically reworks the
independence assumption from unconditional to conditional. The audit
trail is built into the template, not into prose.

## 6. What to inspect next

The point of the structured object is not just to render equations; it
is to tell the reader *what to look at next*. `model_card(sym)`
packages the full teaching bundle in one object, including a class-
specific list of extraction calls. For a `drmTMB` fit, the extraction
calls are ordered diagnostics-first:

```{r model-card-calls}
mc <- model_card(sym)
mc$extraction_calls
```

The ordering is intentional. A reader should know whether the fit is
trustworthy before they interpret a single coefficient.

1. **`drmTMB::check_drm(fit)`** — convergence and identifiability.
   Did the optimiser converge? Are any parameters at the boundary? Are
   the standard errors finite?
2. **`residuals(fit)`** and **`predict(fit)`** — residual diagnostics.
   Plot residuals against the mu predictors (any structure left in the
   mean?) and absolute residuals against the sigma predictors (did the
   sigma submodel absorb the heteroscedasticity?).
3. **`drmTMB::ranef(fit)`** — random-effect BLUPs, when there are
   any. The distribution of BLUPs is worth a histogram or a QQ plot.
4. **`drmTMB::fixef(fit, dpar = "mu")`** and **`drmTMB::fixef(fit, dpar = "sigma")`**
   — the actual coefficient tables, by submodel. Read these *after*
   the diagnostics, not before.

`model_card()` also ships a set of recommended-plot recipes, in the
same diagnostics-first order:

```{r model-card-plots}
mc$recommended_plots
```

These are text recipes only; the v0.1 surface does not plot live. The
recipes name the plot the reader should make next.

**Takeaway.** Diagnostics before coefficients. The extraction-calls
list is the order a reader should work through the fit.

## 7. The capability gate

`symbolize()` does not silently dispatch on every `drmTMB` fit. The
capability registry gates each `(class, family, component)` tuple with
one of five status words: **Stable**, **First slice**, **Opt-in
control**, **Planned or reserved**, **Unsupported or blocked**.

```{r capabilities}
caps <- symbolizer_capabilities()
caps[caps$class == "drmTMB", ]
```

For v0.1 the relevant rows are:

- `drmTMB / gaussian / mu` — **Stable**
- `drmTMB / gaussian / sigma` — **Stable**
- `drmTMB / gaussian / random_effects` — **First slice**
  (intercept-only `(1 | group)`)
- `drmTMB / gaussian / rho12` — **Planned or reserved**
  (bivariate residual correlation; next on the roadmap)
- All non-Gaussian families — **Planned or reserved**

If you call `symbolize()` on an unsupported tuple, the package errors
with a friendly *"Today we can read…"* message that names the row in
the registry. The registry, not the prose, decides what `symbolize()`
accepts; check the status word before designing a workflow around a
tuple.

The roadmap beyond `mu` and `sigma` is the **coscale** surface: the
bivariate Gaussian family in `drmTMB` carries two means
($\mu_1, \mu_2$), two scales ($\sigma_1, \sigma_2$), and a residual
correlation $\rho_{12}$. The corresponding symbolizer surface will
render the sigma-correlation block alongside the per-trait submodels,
with assumptions and readings to match.

**Takeaway.** The capability registry is the single source of truth.
v0.1 ships Gaussian `mu`, `sigma`, and intercept-only random effects;
the rest is on the published roadmap.

## 8. Where to read next

- `vignette("symbolizer")` — the package-wide Get Started tour.
- `vignette("symbolizer-large")` — a richer worked example with
  factors, transformations, and a multi-level random intercept.
- `vignette("symbolizer-gllvm")` — the parallel story for
  latent-variable models in `gllvmTMB`.
- `vignette("symbolizer-gllvm-design")` — design notes for the
  in-development GLLVM surface.
