---
title: "Structural dependence: phylogenetic, animal-model, and spatial random effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structural dependence: phylogenetic, animal-model, and spatial random effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 6,
  fig.height = 3.5
)
library(symbolizer)
set.seed(1)
```

If you have ever written `(1 | species)` for an animal model, `gr(species, cov = A)` in `brms`, or `s(x, y, bs = "gp")` for a spatial process, you have been writing the **same model** three times — only the matrix that encodes the dependence changes. This article is the first in `symbolizer`'s **concept-axis series**; it teaches the *dependence-structure* axis of GLMM grammar through five worked examples across six R packages.

## The shared grammar

A generalised linear mixed model writes

$$
y_i \mid \mathbf b \;\sim\; \mathcal D(\mu_i, \phi),\qquad
g(\mu_i) \;=\; \mathbf x_i^\top \boldsymbol\beta + \mathbf z_i^\top \mathbf b,
$$

with random effects

$$
\mathbf b \;\sim\; \mathcal N(\mathbf 0,\; \sigma^2\,\mathbf M).
$$

The standard textbook GLMM takes $\mathbf M = \mathbf I$ — the random effects are independent draws. Structural dependence is what you get when $\mathbf M \ne \mathbf I$. Three named instantiations recur:

| Context | Symbol | What $\mathbf M$ encodes |
|---|---|---|
| Phylogenetic | $\mathbf A$ | Shared evolutionary history (tree → correlation) |
| Pedigree (animal model) | $\mathbf A$ | Genetic relatedness (pedigree → correlation) |
| Spatial | $\boldsymbol\Omega$ | Geographic proximity (distance + kernel) |

In abstract / cross-package teaching contexts (this article's prose), we write $\mathbf M$ for the generic role. In a fitted model with a known context (e.g. `drmTMB::phylo()` calls), `symbolizer` chooses the domain-specific letter — $\mathbf A$ for phylogenetic / animal-model, $\boldsymbol\Omega$ for spatial. The symbol-dictionary description for $\boldsymbol\Omega$ explicitly disambiguates it from (a) Wishart precision matrices and (b) CAR / SAR spatial weights matrices.

## Where the matrix comes from

For phylogenetic dependence under a Brownian-motion model of trait evolution, $\mathbf A$ is the **phylogenetic correlation matrix**:

$$
A_{ij} \;=\; \frac{T_{ij}}{T},
$$

where $T_{ij}$ is the shared branch length between species $i$ and $j$ measured from the root, and $T$ is the tree height. For an ultrametric tree all tips are equidistant from the root and $A_{ii} = 1$ by construction. The matrix is symmetric positive-definite.

Two operational representations exist:

```{r build-A, message = FALSE}
suppressPackageStartupMessages({
  library(ape)
  library(MCMCglmm)
})

# A toy ultrametric tree with k = 15 species.
tree <- rcoal(15, tip.label = paste0("sp", 1:15))

# Representation 1 -- tips-only k x k correlation matrix
A_tips <- vcv.phylo(tree, corr = TRUE)
dim(A_tips)

# Representation 2 -- all-nodes A-inverse (tips + internal nodes,
# sparse precision; Hadfield-Nakagawa 2010).
A_all  <- inverseA(tree, nodes = "ALL")
dim(A_all$Ainv)
```

The two representations encode the **same** Brownian-motion prior. The tips-only version is the operational default for `metafor`, `glmmTMB::propto`, and `brms::gr(cov = A)`. The all-nodes version is what `MCMCglmm` uses internally (`ginverse = list(animal = Ainv)`) and what `drmTMB::phylo()` builds under the hood (`drmTMB` cites Hadfield & Nakagawa 2010 explicitly in its `?phylo` help page). The all-nodes form is *richer*: it gives access to ancestral-state BLUPs at internal nodes; the tips-only form marginalises those out. In a balanced setup the two agree on the variance estimate $\hat\sigma_p^2$ — a result we'll return to in §"Tips vs all-nodes".

## Six packages, one phylogenetic model

We simulate a small phylogenetic trait dataset, then fit the same model six different ways. The model is

$$
y_i \;=\; \beta_0 + u_{p_{k[i]}} + e_i,\qquad
u_p \sim \mathcal N(\mathbf 0, \sigma_p^2\,\mathbf A),\quad
e_i \sim \mathcal N(0, \sigma_e^2).
$$

```{r sim-phylo}
k <- nlevels(factor(tree$tip.label))
sigma_p <- 0.7; sigma_e <- 0.3
u_p <- as.numeric(MASS::mvrnorm(1, mu = rep(0, k), Sigma = sigma_p^2 * A_tips))
dat <- data.frame(
  species = factor(tree$tip.label, levels = tree$tip.label),
  y       = u_p + rnorm(k, 0, sigma_e)
)
# Sampling variance "v_i" for the meta-analysis-style fits.
dat$vi <- runif(k, 0.05, 0.15)
```

### Face 1: `metafor::rma.mv` with `R = list(species = A)`

```{r metafor-fit, message = FALSE}
suppressPackageStartupMessages(library(metafor))

fit_meta <- rma.mv(
  yi      = y,
  V       = vi,                # known sampling variances
  random  = list(~ 1 | species),
  R       = list(species = A_tips),
  data    = dat,
  sparse  = TRUE
)
sym_meta <- symbolize(fit_meta)
sym_meta$metadata$phylo_representation
sym_meta$metadata$detected_signals
```

```{r metafor-equations}
equations(sym_meta, notation = "index")
```

The R-matrix tier's group name (`species`) matches the canonical phylogenetic lexicon, so `symbolizer`'s `metafor` extractor sets `metadata$phylo_representation = "tips_only"`, declares `detected_signals = "phylo"`, and adds a row to the symbol dictionary for $\mathbf A$ with role `structured_correlation_phylo`. The assumption table now lists the phylogenetic prior assumptions (Brownian motion, ultrametricity, branch-length scale — your responsibility):

```{r metafor-assumptions}
assumption_table(sym_meta)
```

### Face 2: `MCMCglmm` with `ginverse = list(species = Ainv)`

`MCMCglmm` uses the **all-nodes** representation natively.

```{r mcmcglmm-fit, message = FALSE, results = "hide"}
fit_mcmc <- MCMCglmm(
  y ~ 1,
  random   = ~ species,
  ginverse = list(species = A_all$Ainv),
  data     = dat,
  family   = "gaussian",
  nitt = 5000, burnin = 1000, thin = 5, verbose = FALSE
)
sym_mcmc <- symbolize(fit_mcmc, data = dat)
```

```{r mcmcglmm-meta}
sym_mcmc$metadata$phylo_representation   # "all_nodes"
sym_mcmc$metadata$detected_signals       # "phylo"
sym_mcmc$variance_components             # phylo + residual
sym_mcmc$metadata$heritability           # h^2 derived automatically
```

The same extractor branch handles **animal models** (where `A` comes from a pedigree, not a tree) — see *§Animal-model unification* below.

### Face 3: `glmmTMB::propto`

`glmmTMB` attaches $\mathbf A$ via the `propto()` covariance structure. There is one subtle catch: `propto()` parameterises $\boldsymbol\Sigma = \sigma_p^2\,\mathbf A$ with $\sigma_p^2$ **estimated as a free scalar**, and `glmmTMB` adds an independent residual $\sigma_\text{res}^2\,\mathbf I$ unless you pass `dispformula = ~ 0`. So the full conditional covariance under a propto fit is

$$
\operatorname{Cov}(\mathbf y \mid \mathbf u) \;=\; \sigma_p^2\,\mathbf A + \sigma_\text{res}^2\,\mathbf I,
$$

i.e. **two** free scalars on top of $\mathbf A$. This is the **phylogenetic / pedigree** pattern (Hadfield & Nakagawa 2010; Williams et al. 2025, bioRxiv) — *not* the meta-analysis pattern. The meta-analysis pattern is $\boldsymbol\Sigma = \mathbf V$ exactly (no scalar multiplier) — `glmmTMB`'s reserved `equalto()` will deliver that, but `equalto` isn't yet implemented as of `glmmTMB` ≤ 1.1.11.

```{r glmmTMB-show, eval = FALSE}
library(glmmTMB)
dat$g <- factor(1)
A_blk <- A_tips
rownames(A_blk) <- colnames(A_blk) <- dat$species

fit_glmm <- glmmTMB(
  y ~ 1 + propto(0 + species | g, A_blk),
  data = dat
)
sym_glmm <- symbolize(fit_glmm)
sym_glmm$metadata$phylo_representation  # "tips_only"
warning_table(sym_glmm)                 # propto_two_scalars info row
```

### Face 4: `brms::gr(species, cov = A)`

`brms` exposes the same matrix via the `gr()` group factor:

```{r brms-show, eval = FALSE}
library(brms)
A_blk <- A_tips
data2 <- list(A = A_blk)

fit_brms <- brm(
  y ~ 1 + (1 | gr(species, cov = A)),
  data   = dat,
  data2  = data2,
  family = gaussian(),
  chains = 2, iter = 2000, refresh = 0
)
sym_brms <- symbolize(fit_brms)
sym_brms$metadata$phylo_representation  # "tips_only"
```

### Face 5: `drmTMB::phylo()`

`drmTMB` exposes the marker directly in the formula:

```{r drmtmb-show, eval = FALSE}
library(drmTMB)
fit_drm <- drmTMB(
  y ~ phylo(1 | species, tree = tree),
  family = gaussian(),
  data   = dat
)
sym_drm <- symbolize(fit_drm)
sym_drm$metadata$phylo_representation  # "all_nodes" (HN sparse precision)
```

### Face 6: `gllvmTMB::phylo_unique()`

`gllvmTMB` adds the phylogenetic layer to a latent-variable structure (community phylogenetics; Williams et al. 2025, *MEE*). The latent axes themselves are phylogenetically structured:

```{r gllvmtmb-show, eval = FALSE}
library(gllvmTMB)
fit_gllvm <- gllvmTMB(
  y_long ~ 0 + trait + phylo_unique(0 + trait | species, tree = tree),
  family = gaussian(),
  data   = dat_long
)
sym_gllvm <- symbolize(fit_gllvm)
sym_gllvm$metadata$phylo_representation  # "package_managed"
```

The cross-package agreement story is the point: every fit above attaches the same conceptual object — a phylogenetic correlation matrix $\mathbf A$ on a random-effect block — even though the syntactic surface looks different in each package. `symbolizer` produces the same `metadata$detected_signals = "phylo"`, the same symbol-dictionary row for $\mathbf A$, the same gated phylogenetic assumption rows, no matter which package fitted the model.

## Animal-model unification

The phylogenetic random effect is the **same mathematical object** as the additive-genetic random effect in a quantitative-genetics animal model. The matrix $\mathbf A$ comes from a different source (a pedigree of dam-sire-offspring rather than a phylogenetic tree), but its meaning in the model is identical:

$$
\mathbf u_a \;\sim\; \mathcal N(\mathbf 0, \sigma_A^2\,\mathbf A),\qquad
h^2 \;=\; \frac{\sigma_A^2}{\sigma_A^2 + \sigma_e^2}.
$$

The only interpretation difference is what $\sigma_A^2$ *means*: phylogenetic variance among species in a phylogenetic comparative method, additive genetic variance among individuals in a quantitative-genetics animal model. `symbolizer`'s `symbolize.MCMCglmm()` handles both via the same `ginverse` detection branch, and `drmTMB` ships separate `phylo(term, tree)` and `animal(term, pedigree)` markers that share an internal implementation. For deep-dive treatments of either case, see drmTMB's own [`structural-dependence.Rmd`](https://itchyshin.github.io/drmTMB/articles/structural-dependence.html), [`phylogenetic-models.Rmd`](https://itchyshin.github.io/drmTMB/articles/phylogenetic-models.html), and [`animal-models.Rmd`](https://itchyshin.github.io/drmTMB/articles/animal-models.html) — `symbolizer` cross-links rather than duplicates them.

## Tips-only vs all-nodes

Why does the all-nodes representation matter when the tips-only correlation matrix has all the data information? Three reasons:

1. **Ancestral states**. The all-nodes form lets you query a posterior for the trait value at any internal node — useful for inferring ancestral character states.
2. **Numerical efficiency on large trees**. The all-nodes A-inverse is *sparse* (each internal node connects to its parent and children only). Inverting the tips-only $k \times k$ matrix is $\mathcal O(k^3)$; the sparse A-inverse path is much faster for $k > \sim 200$.
3. **Partial observations**. If some species have measured traits and others are unobserved (but on the tree), the all-nodes form handles this naturally.

On a balanced setup with all species observed, both representations give the same $\hat\sigma_p^2$ to machine precision — the Hadfield-Nakagawa equivalence.

## Spatial: same grammar, different matrix

For spatial dependence, $\mathbf M$ is built from pairwise geographic distances and a kernel. The standard choices are:

| Kernel | Formula | Decay |
|---|---|---|
| Exponential | $C(d) = \exp(-d/\rho)$ | Power |
| Squared-exponential (Gaussian) | $C(d) = \exp(-d^2/\rho^2)$ | Sharp |
| Matérn | $C(d; \kappa, \nu)$ | Tunable smoothness |

Two fits demonstrate the structurally identical grammar with $\boldsymbol\Omega$ in place of $\mathbf A$:

```{r spatial-show, eval = FALSE}
library(sdmTMB)
fit_sp <- sdmTMB(
  y ~ 1,
  data = dat_spatial,
  mesh = mesh,
  spatial = "on",
  family = gaussian()
)
sym_sp <- symbolize(fit_sp)
sym_sp$metadata$spatial_representation  # "package_managed"

library(mgcv)
fit_gp <- gam(y ~ s(x, y, bs = "gp"), data = dat_spatial)
sym_gp <- symbolize(fit_gp)
```

`symbolizer`'s metafor extractor also handles `rma.mv(struct = "SPEXP", dist = list(location = D))` — Mizuno et al. (2026, *Research Synthesis Methods*) is the conceptual anchor for the unified phylogenetic + spatial meta-analytic story.

## When unification breaks: identifiability

The unification works mathematically, but practical inference depends on the data. When you stack a phylogenetic tier and a non-phylogenetic species-level tier on the same grouping column,

$$
y_i = \beta_0 + u_{p_{k[i]}} + u_{s_{k[i]}} + e_i,\qquad
u_p \sim \mathcal N(0, \sigma_p^2 \mathbf A),\quad
u_s \sim \mathcal N(0, \sigma_s^2 \mathbf I),
$$

the variance components $\sigma_p^2$ and $\sigma_s^2$ are *weakly identified*: their sum is precisely estimable but the split between them is not (Hadfield & Nakagawa 2010 §3.2; Mizuno et al. 2026 §4). With strong phylogenetic signal (Pagel's λ near 1) the split is informative; with weak signal (λ near 0) the data have almost no leverage to separate the two. `symbolizer` emits a `phylo_nonphylo_unidentifiable` row in `warning_table()` when this configuration is detected. Profile on λ or fix one tier to resolve.

## Forward links

- **v0.22 — Phylogenetic meta-analysis (Mizuno et al. 2026)**. This article fits the general phylo-GLMM with $\sigma_e^2$ *estimated*. The meta-analytic case where the sampling variance $v_i$ is **known** per study is its own surface — `metafor::rma.mv(V = V, ...)` is the canonical interface, with `glmmTMB::equalto` (reserved) and `brms::se(sqrt(vi))` as bridges.
- **v0.24 — Location-scale on $\mathbf M$ (Nakagawa et al. 2025 *MEE*)**. The dependence parameter itself can have a submodel: $\sigma_p \sim z$ across clades, $\rho \sim z$ across environments. Symbolizer's drmTMB extractor is the right home for the phylogenetic location-scale model.
- **drmTMB deep dives**: `vignette("phylogenetic-models", package = "drmTMB")`, `vignette("animal-models", package = "drmTMB")`, `vignette("structural-dependence", package = "drmTMB")`.
- **gllvmTMB deep dive**: `vignette("phylogenetic-gllvm", package = "gllvmTMB")`.

## References

- Hadfield, J. D. (2010). MCMC methods for multi-response generalized linear mixed models: the `MCMCglmm` R package. *Journal of Statistical Software*, 33(2), 1–22.
- Hadfield, J. D. & Nakagawa, S. (2010). General quantitative genetic methods for comparative biology: phylogenies, taxonomies and multi-trait models for continuous and categorical characters. *Journal of Evolutionary Biology*, 23, 494–508.
- Lynch, M. (1991). Methods for the analysis of comparative data in evolutionary biology. *Evolution*, 45(5), 1065–1080.
- Mizuno, A., Williams, C., Lagisz, M., Senior, A. M. & Nakagawa, S. (2026). A unified framework for phylogenetic and spatial meta-analysis: concepts, implementation, and practical guidance. *Research Synthesis Methods*.
- Nakagawa, S., Mizuno, A., Williams, C., Lagisz, M., Yang, Y. & Drobniak, S. M. (2025). Quantifying macro-evolutionary patterns of trait mean and variance with phylogenetic location-scale models. *Methods in Ecology and Evolution*, 16, 2585–2602.
- Williams, C., McGillycuddy, M., Drobniak, S. M., Bolker, B. M., Warton, D. I. & Nakagawa, S. (2025). Fast phylogenetic generalised linear mixed-effects modelling using the `glmmTMB` R package. *bioRxiv* 2025.12.20.695312.
