---
title: "Multi-node structural equation models: piecewiseSEM and drmSEM"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Multi-node structural equation models: piecewiseSEM and drmSEM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The live path in this article fits a `piecewiseSEM::psem()` from base-R
# `lm()` nodes, so it runs wherever piecewiseSEM is installed. The drmSEM
# section is shown as reference code only (eval = FALSE) -- it needs the
# off-CRAN drmSEM + a compiled drmTMB fit -- so the vignette stays
# build-safe, matching the sibling cross-package articles.
has_psem <- requireNamespace("piecewiseSEM", quietly = TRUE) &&
            requireNamespace("symbolizer", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_psem
)
library(symbolizer)
set.seed(1)
```

> *A piecewise structural equation model is a **system** of regressions —
> one fitted model per response, wired together into a causal graph. This
> article shows how `symbolize()` reads the whole system: it walks each
> node, symbolizes it with the node's own class method, and hands back a
> collator whose equations, assumption table, and HTML widget are stacked
> node by node. Two flavours are covered: mean-only piecewise SEMs from
> `piecewiseSEM`, and distributional (location-scale) piecewise SEMs from
> `drmSEM`.*

## A SEM is a list of regressions

Where a single GLM has one response, a piecewise SEM has several — each
its own fitted model, each a **node** in a directed graph. A path that
appears as a predictor in one node is a response in another. The model
of avian fecundity below is three statements:

- `body_size ~ temperature` — body size tracks temperature,
- `fecundity ~ body_size + rainfall` — fecundity tracks body size and rainfall,
- `rainfall %~~% temperature` — the two climate drivers are correlated
  (a residual covariance arc, **not** a regression).

`piecewiseSEM::psem()` glues those fitted nodes into one object.
`symbolize()` reads it as a unit.

```{r psem-fit, message = FALSE}
library(piecewiseSEM)

n <- 120
d <- data.frame(temperature = rnorm(n), rainfall = rnorm(n))
d$body_size <- 0.5 * d$temperature + rnorm(n)
d$fecundity <- 0.4 * d$body_size + 0.3 * d$rainfall + rnorm(n)

m_size <- lm(body_size ~ temperature, data = d)
m_fec  <- lm(fecundity ~ body_size + rainfall, data = d)

sem <- psem(m_size, m_fec, rainfall %~~% temperature, data = d)

sym <- symbolize(
  sem,
  symbols = c(body_size = "S_i", fecundity = "F_i",
              temperature = "T_i", rainfall = "R_i")
)
sym
```

`symbolize()` returns a `symbolized_psem` — a collator holding one
`symbolized_model` per node (`sym$parts`), the node names in declared
order (`sym$node_names`), and any `%~~%` arcs (`sym$cov_arcs`). Every
node was dispatched on its own class:

```{r psem-parts}
sym$node_names
vapply(sym$parts, function(p) p$model$class, character(1))
```

## The equations, node by node

`equations()` row-binds every node's equation block with a leading
`node` column, so you can see the whole system at once:

```{r psem-equations}
equations(sym)
```

`as_latex()` concatenates the per-node LaTeX under `## Node:` headers,
ready to splice into a manuscript:

```{r psem-latex}
cat(as_latex(sym))
```

## The assumptions — including the covariance arc

`assumption_table()` row-binds each node's assumptions, again with a
leading `node` column. The `%~~%` arc surfaces here as a
`Residual cov(...)` row — and **only** here. A bidirected arc is a
covariance, not a regression, so it never appears as an equation:

```{r psem-assumptions}
assumption_table(sym)
```

## Three views of the whole SEM

`as_html_three_views()` renders the interactive *Index / Matrix /
Matrix-with-data* widget for **every node**, stacked under node headers.
Each node carries its own real data through the third tab:

```{r psem-widget, results = "asis", message = FALSE}
as_html_three_views(sym, id = "psem")
```

## The distributional flavour: `drmSEM`

A `piecewiseSEM` node is mean-only — one linear predictor per response.
When a path needs a **distributional** structure (a scale submodel, a
zero-inflation component, a residual correlation between two responses),
the node is a `drmTMB` location-scale fit and the SEM is built with
[`drmSEM::drm_sem()`](https://itchyshin.github.io/drmSEM/). `symbolize()`
reads it through the `drmSEM`-side `symbolize.drm_sem` method, which
delegates each node to `symbolize.drmTMB`:

```{r drmsem-ref, eval = FALSE}
library(drmSEM)
library(drmTMB)

sem <- drm_sem(
  size = drm_node(drm_formula(size ~ temp, sigma ~ temp),
                  family = gaussian()),
  fec  = drm_node(drm_formula(fec ~ size + temp),
                  family = gaussian()),
  data = dat
)

sym <- symbolize(sem)   # dispatches to drmSEM::symbolize.drm_sem
equations(sym)          # node column + per-node mu AND sigma rows
assumption_table(sym)   # per-node assumptions, distributional
as_html_three_views(sym)  # same stacked widget, location-scale nodes
```

The `sigma ~ temp` submodel on the `size` node is what `piecewiseSEM`
cannot express: it models the **spread** of body size as a function of
temperature, not just the mean. Everything downstream — the equations,
the assumption table, the HTML widget — works identically, because both
collators share one parent class.

## Why the same methods work on both

`symbolize.psem` tags its output `symbolized_model_set`; `drmSEM`'s
`symbolize.drm_sem` tags its output the same way. Both store their nodes
in a **named** `parts` list. So the rendering generics dispatch on the
shared parent and read `names(x$parts)` — one implementation, two
collators. The capability registry records which package owns each
method:

```{r registry}
caps <- symbolizer_capabilities()
caps[caps$class %in% c("piecewiseSEM", "drm_sem"),
     c("class", "status", "since", "lives_in", "notes")]
```

| | `piecewiseSEM` (`psem`) | `drmSEM` (`drm_sem`) |
|---|---|---|
| Per-node model | `lm` / `glm` / `lmer` / `glmer` / `glmmTMB` / `lme` | `drmTMB` location-scale |
| Structure per node | mean only (one linear predictor) | distributional (`mu`, `sigma`, `nu`, `zi`, `hu`, `rho12`) |
| Method lives in | `symbolizer` (bundled) | `drmSEM` |
| Collator class | `symbolized_psem` → `symbolized_model_set` | `symbolized_drm_sem` → `symbolized_model_set` |
| `%~~%` / residual covariance | `Residual cov(...)` assumption row | per-node `rho12` distributional component |

Reach for `piecewiseSEM` when each response is adequately summarised by
its mean; reach for `drmSEM` when a path's **variance, skew, or
zero-inflation** is itself part of the story.
