---
title: "Building up: from lm to location-scale"
description: >
  One synthetic dataset, four models of increasing complexity, the same
  `symbolize()` call. Watch the symbolic equation grow one row at a
  time as you add a factor, then a random effect, then a heterogeneity
  submodel.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Building up: from lm to location-scale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_lme4    <- requireNamespace("lme4",   quietly = TRUE)
has_drmTMB  <- requireNamespace("drmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_lme4 && has_drmTMB
)
```

## The idea

> **New to symbolizer?** Skim `vignette("symbolizer")` first — a five-minute
> quickstart on a single model. This article is the next step up: it climbs from
> `lm` to a location-scale model so you can watch the symbolic story grow.

`symbolizer` turns any fitted model into a structured symbolic
specification. To see *what* that buys you, the easiest path is to
fit a sequence of models — each adding one feature — and watch the
symbolic specification grow rung by rung.

Throughout this article we use **one dataset** — body mass measured
across temperature, sex, and site — and climb four rungs:

| Rung | Model | What's new in the symbolic story |
| --- | --- | --- |
| 1 | `lm(body_mass ~ temperature)` | distribution + linear predictor |
| 2 | `lm(body_mass ~ temperature + sex)` | a factor: dummy-encoded contrast |
| 3 | lmer(body_mass ~ temperature + sex + (1 \| site)) | a random intercept + variance components |
| 4 | drmTMB(body_mass ~ temperature + sex + (1 \| site), sigma ~ temperature) | heteroscedasticity: variance depends on temperature |

The reader-facing surface stays exactly the same: at every rung you
call `symbolize(fit)`, then `as_latex()` / `symbol_table()` /
`parameter_interpretation()` to read it. The *contents* of those
outputs grow.

## Setup — one shared dataset

```{r data}
library(symbolizer)
set.seed(1)

n <- 200
n_sites <- 12
sites <- factor(rep(paste0("S", sprintf("%02d", seq_len(n_sites))),
                    length.out = n))
site_effect <- rnorm(n_sites, 0, 1.5)[as.integer(sites)]
sex <- factor(sample(c("F", "M"), n, replace = TRUE))
temperature <- runif(n, 10, 25)

# Mean: linear in temperature, +0.7 for males, plus a site offset.
# Variance: heteroscedastic — residual SD grows with temperature.
mu_true    <- 30 + 0.4 * temperature + 0.7 * (sex == "M") + site_effect
sigma_true <- exp(0.3 + 0.05 * temperature)
body_mass  <- rnorm(n, mean = mu_true, sd = sigma_true)

dat <- data.frame(
  body_mass   = body_mass,
  temperature = temperature,
  sex         = sex,
  site        = sites
)
head(dat, 4)
```

The "truth" the ladder will gradually uncover:

- Slope of temperature on the mean: `0.4` g per °C
- Sex contrast (M − F) on the mean: `+0.7` g
- Site-to-site SD on the mean: `1.5` g
- Heteroscedasticity: residual SD grows by a factor of `exp(0.05) ≈ 1.05` per °C

## Rung 1 — A linear model

The simplest fit. Just temperature on the mean.

```{r rung1-fit}
fit1 <- lm(body_mass ~ temperature, data = dat)
sym1 <- symbolize(fit1)
```

The equation:

```{r rung1-latex}
equations(sym1, notation = "index")
```

The symbol dictionary, listing every variable the model uses:

```{r rung1-symbols}
symbol_table(sym1)
```

What each coefficient means, on each scale `symbolizer` knows about:

```{r rung1-interp}
parameter_interpretation(sym1, scale = "biological")
```

**What we have so far.** A distribution line for the response, a
linear predictor for $\mu_i$, and one slope to interpret. No
variance modelling beyond a single residual SD. No grouping. The
biological reading is "a unit change in temperature shifts the
expected body_mass by `r if (exists("fit1")) round(coef(fit1)[2], 3) else "0.4"`."

## Rung 2 — Adding a factor

What if mass differs systematically between sexes? Add `sex` to the
mean.

```{r rung2-fit}
fit2 <- lm(body_mass ~ temperature + sex, data = dat)
sym2 <- symbolize(fit2)
```

```{r rung2-latex}
equations(sym2, notation = "index")
```

The equation now carries a contrast term. The symbol dictionary
adds a row for the dummy-encoded contrast:

```{r rung2-symbols}
symbol_table(sym2)
```

`sex` is a factor with reference level `F` (alphabetical). The
single dummy column `[sex = M]` enters the linear predictor with
its own slope $\beta_2$. The interpretation reads the contrast
explicitly:

```{r rung2-interp}
pi2 <- parameter_interpretation(sym2, scale = "biological")
pi2[pi2$coefficient_role == "factor_contrast", c("term_label", "estimate", "biological_reading")]
```

**What just got added.** One row in the symbol table (the factor
contrast), one row in `parameter_interpretation()`, and one term in
the LaTeX. The model class didn't change (`lm` to `lm`), but the
symbolic story grew exactly where the model grew.

## Rung 3 — Adding random effects

Sites probably differ from one another, even at the same
temperature. Add a random intercept per site.

```{r rung3-fit}
library(lme4)
fit3 <- lmer(body_mass ~ temperature + sex + (1 | site), data = dat)
sym3 <- symbolize(fit3)
```

```{r rung3-latex}
equations(sym3, notation = "index")
```

Two new rows in the equation block:

1. `u_{site(i)}` is added to the linear predictor for $\mu_i$.
2. A second distributional line: `u_{site} ~ N(0, σ²_site)`.

Variance components are now first-class:

```{r rung3-vc}
sym3$variance_components
```

You can read off the between-site SD and the residual SD directly.
The fixed-effect interpretation rows for temperature and sex still
read on the response scale; what changed is that the model now
*partitions* the variability into between-site and within-site
parts.

**What just got added.** A `u_{site(i)}` symbol, a random-effect
distributional line, and a `variance_components` tibble showing the
two SDs. Same `symbolize()` call, richer object.

## Rung 4 — Location-scale (heteroscedasticity)

Look at the residuals against temperature — they fan out:

```{r rung4-resid, fig.width = 5, fig.height = 3.2, fig.alt = "Residuals against temperature with a wedge-shaped spread"}
plot(dat$temperature, residuals(fit3),
     xlab = "temperature (°C)", ylab = "residual (g)",
     pch = 16, col = "#a0282b")
abline(h = 0, col = "#666666")
```

The variance is changing with temperature. A standard mixed model
forces residual variance to be constant, so the inference on
$\beta_1$ ignores that. This is called a **location-scale model**
(also known as **distributional regression** or, when the variance
component is what changes, *heteroscedastic regression*): instead of
treating the residual SD as a single fixed number, we let it depend
on its own predictors with its own coefficients. Climb the last rung:
fit $\log(\sigma_i) = \gamma_0 + \gamma_1\, T_i$.

```{r rung4-fit}
library(drmTMB)
fit4 <- drmTMB(
  drm_formula(body_mass ~ temperature + sex + (1 | site),
              sigma     ~ temperature),
  family = gaussian(),
  data   = dat
)
sym4 <- symbolize(fit4)
```

```{r rung4-latex}
equations(sym4, notation = "index")
```

The equation block grew by **one new line** — the
$\log(\sigma_i) = \gamma_0 + \gamma_1\, T_i$ row. That single
additional line is the heteroscedasticity story.

Reading on the variability scale:

```{r rung4-sigma-read}
pi4 <- parameter_interpretation(sym4, scale = "biological")
pi4[pi4$submodel == "sigma", c("term_label", "estimate", "biological_reading")]
```

The slope $\gamma_1$ reads as "residual SD multiplies by
`exp(γ_1)` per °C". For our simulated data the truth was
`exp(0.05) ≈ 1.05`, so each additional °C inflates the residual SD
by about 5 %.

## Three views of the same fit

For the full educator-facing surface — equation, index expansion,
matrix form, with the actual numeric arrays running alongside —
ask for the three-view widget:

```{r three-views, eval = FALSE}
as_html_three_views(sym4)
```

```{r three-views-live, echo = FALSE, results = "asis"}
if (has_drmTMB) {
  # `as_html_three_views()` emits raw HTML via `cat()`; in a
  # `results = "asis"` chunk this becomes the rendered widget. Do
  # NOT wrap in `htmltools::tagList()` -- tagList treats the
  # character return value as a text node and re-emits the same
  # HTML escaped, which leaks raw tags into the page.
  invisible(as_html_three_views(sym4))
}
```

(The widget renders live in this page. In an R session,
`cat(as_html_three_views(sym))` inside a `results = "asis"` chunk
will render the same widget into a knitted document.)

## What just happened: a recap

```{r recap, echo = FALSE, results = "asis"}
recap <- data.frame(
  # Use explicit <code> tags rather than markdown backticks: when the
  # surrounding chunk is `results = "asis"` + kable(format = "html",
  # escape = FALSE), pandoc still parses markdown inside the cells, and
  # backticks around `(1|site)` turn into <code>(1&#124;site)</code> --
  # which then gets `&` -> `&amp;` re-escaped and surfaces as literal
  # `&amp;#124;`. Raw <code> bypasses the markdown reader.
  Rung     = c("1: <code>lm</code>",
               "2: <code>lm + sex</code>",
               "3: <code>lmer + (1|site)</code>",
               "4: <code>drmTMB</code> location-scale"),
  Equation = c(
    "$y_i \\sim \\mathcal{N}(\\mu_i,\\, \\sigma^2)$; $\\mu_i = \\beta_0 + \\beta_1 T_i$",
    "$\\mu_i = \\beta_0 + \\beta_1 T_i + \\beta_2 [\\mathit{sex} = M]$",
    "$\\mu_i = \\beta_0 + \\beta_1 T_i + \\beta_2 [\\mathit{sex} = M] + u_{site(i)}$;<br>$u_{site} \\sim \\mathcal{N}(0,\\, \\sigma^2_{site})$",
    "as Rung 3, plus $\\log(\\sigma_i) = \\gamma_0 + \\gamma_1 T_i$"
  ),
  `New rows in symbolize output` = c(
    "distribution + linear predictor",
    "+ factor contrast row in `symbol_table` / interp",
    "+ random-effect symbol + `variance_components` rows",
    "+ `sigma` submodel + heteroscedasticity reading"
  ),
  check.names = FALSE
)
knitr::kable(recap, format = "html", escape = FALSE)
```

Three observations:

1. **The same `symbolize()` call works on every fit.** No
   class-specific accessors. The four fits cover three different
   packages (`stats`, `lme4`, `drmTMB`) and the user-facing surface
   is identical.
2. **The equation block grows exactly where the model grows.** Each
   rung adds *one* concept; the LaTeX line count tells you which.
3. **The symbolic surface is incrementally readable.** A student
   who just learned `lm` can read Rung 1; a student who just learned
   random effects can read Rung 3 *and* see what Rung 4 adds on
   top.

The same approach scales to every model class `symbolizer`
covers — `glmmTMB`, `brms`, `MCMCglmm`, `sdmTMB`, `metafor`,
`mgcv`. The educator-facing surface stays one verb deep.

## Where to next

- For factor / interaction pedagogy in depth, see
  `vignette("symbolizer-factors")`.
- For a tour of distributional families (Student-t, Gamma, beta,
  binomial, …), see `vignette("symbolizer-families")`.
- For meta-analysis with `metafor` and `drmTMB` location-scale
  (plus `glmmTMB`'s `propto()` as the related phylogenetic /
  structured-covariance pattern, not strict meta-analysis), see
  `vignette("symbolizer-meta-analysis")`.
- For the full capability matrix and what's planned next, see
  `vignette("symbolizer-roadmap")`.
