---
title: "A tour of non-Gaussian families"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A tour of non-Gaussian families}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_drmTMB <- requireNamespace("drmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_drmTMB
)

# Render a single LaTeX string as a display-math block. Use this for the
# distribution line on each family. The returned value is asis_output, so
# it must be the last expression in the chunk.
render_math <- function(x) {
  knitr::asis_output(paste0("\n$$\n", x, "\n$$\n"))
}

# Write a PDF rendering of the three-views widget next to the rendered
# HTML, and emit a "Download as PDF" link. Mirrors the helper used in
# vignettes/symbolizer-structural-dependence.Rmd -- write into the
# chunk cwd (= vignettes/) and copy into ../docs/articles when that
# exists (pkgdown build path) so the <a href> resolves under both
# rmarkdown::render() and pkgdown::build_article().
pdf_alongside_html <- function(sym, basename, title) {
  symbolizer::as_pdf_three_views(sym, file = basename, title = title)
  docs_articles <- "../docs/articles"
  if (dir.exists(docs_articles)) {
    file.copy(basename, file.path(docs_articles, basename), overwrite = TRUE)
  }
  cat(sprintf(
    '<p><a href="%s" class="btn btn-primary">Download as PDF</a></p>\n',
    basename
  ))
  invisible(basename)
}
```

> *This tour walks through the non-Gaussian distributional families
> available for `drmTMB`. The biology of each is different — what
> kind of response it fits, what the link does, and how to read the
> `mu` coefficient. The same `symbolize(fit)` interface now applies
> to ten package families (`drmTMB`, `gllvmTMB`, `glmmTMB`, `brms`,
> `lme4`, `MCMCglmm`, `sdmTMB`, `stats::lm` / `stats::glm`, `metafor`,
> `mgcv`); see the [Roadmap article](symbolizer-roadmap.html) for the
> full capability matrix.*

The non-Gaussian families covered here all use the same
`symbolize(fit)` interface. The differences are in the LaTeX, the
biological reading on each coefficient, and the assumption table.
Every phrase you see below is templated from
`inst/extdata/*.csv` — no LLM at runtime.

For Gaussian and bivariate Gaussian, see
`vignette("symbolizer-drmtmb")`. For the matching pedagogy on
factors and interactions, see `vignette("symbolizer-factors")`.

```{r setup-libs}
library(symbolizer)
library(drmTMB)
```

## Continuous responses

### Student-t — robust regression

A heavier-tailed Gaussian alternative. The `nu` parameter controls
tail heaviness: `nu` large means almost Gaussian, `nu` small means
heavy outliers tolerated. drmTMB parameterises `nu` on the
`log(nu - 2)` scale so the variance is always finite.

```{r student}
set.seed(1); n <- 100
dat <- data.frame(
  y = rt(n, df = 5) * 2 + 10 + 0.5 * rnorm(n),
  x = rnorm(n)
)
fit <- drmTMB(drm_formula(y ~ x, sigma ~ 1, nu ~ 1),
              family = student(), data = dat)
sym <- symbolize(fit)
render_math(sym$distribution$latex)
```

**Coefficient reading on mu:** identical to Gaussian — a slope on
the response scale.

### Lognormal — positive continuous, right-skewed

`Y > 0`, and `log(Y)` is Gaussian. Coefficients on `mu` live on the
log-response scale, so `exp(beta)` is the multiplicative effect on
the **geometric mean** of `Y`.

```{r lognormal}
set.seed(1); n <- 100
dat <- data.frame(y = exp(rnorm(n, 2 + 0.3 * rnorm(n), 0.4)),
                  x = rnorm(n))
fit <- drmTMB(drm_formula(y ~ x, sigma ~ 1),
              family = lognormal(), data = dat)
sym <- symbolize(fit)
parameter_interpretation(sym, scale = "biological")
```

**Coefficient reading on mu:** a unit change in `x` multiplies the
geometric mean by `exp(beta)`.

#### Three views of the lognormal fit

```{r lognormal-widget, results = "asis", message = FALSE}
as_html_three_views(sym, id = "lognormal")
```

```{r lognormal-pdf, results = "asis", message = FALSE}
pdf_alongside_html(sym, "fig-lognormal.pdf",
                   "Lognormal regression -- three views")
```

### Gamma — positive continuous, skewed

`Y > 0`, log link by convention (`stats::Gamma(link = "log")`).
Coefficients are multiplicative on the mean response, like
lognormal — but the variance scales as `mu^2 * sigma^2`, not
`log(Y)`.

```{r gamma}
set.seed(1); n <- 100
dat <- data.frame(y = rgamma(n, shape = 2, rate = 0.5),
                  x = rnorm(n))
fit <- drmTMB(drm_formula(y ~ x, sigma ~ 1),
              family = stats::Gamma(link = "log"), data = dat)
sym <- symbolize(fit)
render_math(sym$distribution$latex)
```

**When to choose Gamma over lognormal?** Gamma is a member of the
exponential family (it has nice GLM properties — link, canonical
form). Lognormal is fully implied by `log(Y) ~ Normal`, which makes
some predictive plots simpler but loses the GLM toolkit.

## Proportion responses

### Beta — proportions in (0, 1)

`Y` is strictly between 0 and 1. Mean via logit link, precision
via log link (large `sigma` means a *tighter* distribution around
the mean).

```{r beta}
set.seed(1); n <- 100
dat <- data.frame(y = rbeta(n, 2, 5),
                  x = rnorm(n))
fit <- drmTMB(drm_formula(y ~ x, sigma ~ 1),
              family = beta(), data = dat)
sym <- symbolize(fit)
render_math(sym$distribution$latex)
```

**Coefficient reading on mu:** `exp(beta)` is the *odds ratio* of
the success probability. A coefficient of `0.3` means a unit
increase in `x` multiplies the odds by `exp(0.3) ≈ 1.35`.

#### Three views of the Beta fit

```{r beta-widget, results = "asis", message = FALSE}
as_html_three_views(sym, id = "beta")
```

```{r beta-pdf, results = "asis", message = FALSE}
pdf_alongside_html(sym, "fig-beta.pdf",
                   "Beta regression -- three views")
```

### Beta-binomial — overdispersed binomial counts

`Y` is the count of successes out of `N_i` trials. Specify via
`cbind(successes, failures)`. `sigma` controls overdispersion: large
`sigma` approaches binomial (no overdispersion), small `sigma`
means strong overdispersion.

```{r beta-binomial}
set.seed(1); n <- 80
trials <- sample(5:20, n, replace = TRUE)
p <- plogis(0.5 + 0.5 * rnorm(n))
dat <- data.frame(
  successes = rbinom(n, trials, p),
  failures  = trials - rbinom(n, trials, p),
  x         = rnorm(n)
)
fit <- drmTMB(drm_formula(cbind(successes, failures) ~ x, sigma ~ 1),
              family = beta_binomial(), data = dat)
sym <- symbolize(fit)
render_math(sym$distribution$latex)
```

**Coefficient reading on mu:** same logit reading as Beta —
`exp(beta)` is the odds ratio for the success probability.

## Count responses

### Poisson — counts with no dispersion parameter

`Y in {0, 1, 2, ...}`, log link on the mean count rate. The Poisson
family forces `variance = mean`; if your data show overdispersion
(variance > mean), prefer `nbinom2`.

```{r poisson}
set.seed(1); n <- 100
dat <- data.frame(y = rpois(n, lambda = exp(1 + 0.3 * rnorm(n))),
                  x = rnorm(n))
fit <- drmTMB(drm_formula(y ~ x),
              family = stats::poisson(), data = dat)
sym <- symbolize(fit)
render_math(sym$distribution$latex)
```

**Coefficient reading on mu:** `exp(beta)` is the *rate ratio* — a
multiplicative effect on the expected count.

#### Three views of the Poisson fit

```{r poisson-widget, results = "asis", message = FALSE}
as_html_three_views(sym, id = "poisson")
```

```{r poisson-pdf, results = "asis", message = FALSE}
pdf_alongside_html(sym, "fig-poisson.pdf",
                   "Poisson regression -- three views")
```

### Negative binomial (nbinom2) — counts with overdispersion

`Y in {0, 1, 2, ...}`, log link on the mean count rate. `sigma`
parameterises overdispersion: `size = exp(sigma)`. Small `size`
means highly overdispersed (much higher variance than mean); large
`size` approaches Poisson.

```{r nbinom2}
set.seed(1); n <- 100
dat <- data.frame(y = rnbinom(n, size = 3, mu = exp(1 + 0.3 * rnorm(n))),
                  x = rnorm(n))
fit <- drmTMB(drm_formula(y ~ x, sigma ~ 1),
              family = nbinom2(), data = dat)
sym <- symbolize(fit)
render_math(sym$distribution$latex)
```

**Coefficient reading on mu:** same as Poisson — `exp(beta)` is
the rate ratio.

### Zero-truncated nbinom2 — counts that are never zero

`Y in {1, 2, 3, ...}`. The truncation is built into the likelihood,
not a data filter. The modelled `mu` is the mean of the *underlying
untruncated* distribution; the observed truncated mean is larger.

```{r truncated-nbinom2}
set.seed(1); n <- 100
dat <- data.frame(
  y = pmax(1L, rnbinom(n, size = 2, mu = exp(1 + 0.3 * rnorm(n)))),
  x = rnorm(n)
)
fit <- drmTMB(drm_formula(y ~ x, sigma ~ 1),
              family = truncated_nbinom2(), data = dat)
sym <- symbolize(fit)
render_math(sym$distribution$latex)
```

**Coefficient reading on mu:** `exp(beta)` is the rate ratio on
the untruncated mean. The observed (truncated) mean has a different
expression that depends on `size`.

## Bivariate Gaussian — two responses jointly

For paired continuous responses with a residual correlation
between them, see `vignette("symbolizer-drmtmb")` Section 7. The
bivariate Gaussian shape is structurally different — five
submodels (`mu1`, `mu2`, `sigma1`, `sigma2`, `rho12`) instead of
two or three — and has its own dedicated section in the drmTMB
vignette.

## Picking a family

Six rules of thumb:

1. **Continuous, roughly symmetric, fixed scale** → `gaussian`.
2. **Continuous, symmetric, heavier tails (outliers tolerated)** →
   `student`.
3. **Continuous, positive, right-skewed** → `lognormal` or
   `Gamma`. Use lognormal when interpretation is naturally on the
   `log(Y)` scale (e.g., body mass, lifespan). Use Gamma when the
   GLM framework matters (e.g., for diagnostics).
4. **Proportion in (0, 1)** → `beta`. **Count of successes out of
   a known number of trials** → `beta_binomial`.
5. **Counts with variance = mean** → `poisson`. **Counts with
   variance > mean** → `nbinom2`.
6. **Counts that are structurally non-zero** → `truncated_nbinom2`.

If none of these fit, check
`symbolizer_capabilities()` — `symbolize()` will refuse with a
clear message naming what's not yet supported.

## Where to read next

- `vignette("symbolizer-drmtmb")` — the deep drmTMB tour
  (Gaussian + bivariate Gaussian + random effects).
- `vignette("symbolizer-factors")` — categorical pedagogy (works
  the same way for any family).
- `vignette("symbolizer-compare")` — comparing two fits with
  `compare_symbolic()`.
