---
title: "Where the variation lives: ICC and repeatability"
description: "Partition variance into components, then read ICC and repeatability -- and which scale each number lives on."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Where the variation lives: ICC and repeatability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_lme4    <- requireNamespace("lme4", quietly = TRUE)
has_glmmTMB <- requireNamespace("glmmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_lme4
)
library(symbolizer)
```

> *A mixed model's payoff for a biologist is two questions: **where does the
> variation live**, and **how repeatable is the trait**. `symbolizer` reads
> those answers off the fitted model -- with an explicit, honest account of
> which scale each number lives on.*

When you fit a mixed model with a grouping factor -- individuals measured more
than once, nests, populations, sites -- the model splits the total variation
into a *between-group* part and a *within-group* (residual) part.
`variance_partition()` reports that split, and `icc()` reports the
**intraclass correlation**: the share of variation that is between groups.
For repeated measurements of the same individuals, the ICC is exactly the
**repeatability** $R$ -- the quantity the
[rptR](https://cran.r-project.org/package=rptR) package was built to estimate
(Nakagawa & Schielzeth 2010; Stoffel, Nakagawa & Schielzeth 2017).

This vignette uses ordinary GLMMs from `lme4` and `glmmTMB` -- the engines a
biologist actually reaches for. Every number below is a point estimate; no
confidence intervals are shown (see the package's uncertainty contract). As
everywhere in `symbolizer`, the readings come from templates in
`inst/extdata/`, not from a language model.

## 1. Gaussian repeatability with `lme4`

Suppose boldness is scored three times for each of 30 individuals, with the
assay temperature recorded each time. We want the **repeatability** of
boldness: how much of the variation is consistent differences *between*
individuals, versus variation *within* an individual across repeats.

```{r gaussian-fit}
set.seed(1)
n_ind <- 30L; n_rep <- 3L
ind   <- factor(rep(seq_len(n_ind), each = n_rep))
temp  <- round(runif(n_ind * n_rep, 14, 24), 1)
u_ind <- rnorm(n_ind, sd = 1.2)                      # between-individual
boldness <- 2 + 0.10 * temp + u_ind[ind] + rnorm(n_ind * n_rep, sd = 1.0)
dat_g <- data.frame(boldness, temp, ind)

fit_g <- lme4::lmer(boldness ~ temp + (1 | ind), data = dat_g)
sym_g <- symbolize(fit_g)
```

`variance_partition()` shows where the variation lives -- one row per source,
as a share of the total:

```{r gaussian-partition}
variance_partition(sym_g)
```

And `icc()` turns that into a single repeatability number. Because this is a
Gaussian-identity fit with one random intercept, the ICC is a **true
proportion of variance in the observed response** -- the *data scale*:

```{r gaussian-icc}
icc(sym_g)
```

**Takeaway.** A few tenths of the boldness variation is consistent between
individuals; the rest is within-individual. That between-individual share *is*
the repeatability of boldness.

## 2. Binomial repeatability with `lme4`

Now a binary outcome: whether each offspring survived, with several offspring
per nest. The "repeatability of survival across nests" is a binomial GLMM with
a `(1 | nest)` random intercept.

```{r binomial-fit}
set.seed(2)
n_nest <- 25L; brood <- 5L
nest   <- factor(rep(seq_len(n_nest), each = brood))
temp_n <- round(runif(n_nest * brood, 14, 24), 1)
u_nest <- rnorm(n_nest, sd = 1.0)                    # between-nest, logit scale
eta    <- -0.2 + 0.05 * temp_n + u_nest[nest]
surv   <- rbinom(n_nest * brood, size = 1, prob = plogis(eta))
dat_b  <- data.frame(surv, temp_n, nest)

fit_b <- lme4::glmer(surv ~ temp_n + (1 | nest), data = dat_b,
                     family = binomial)
sym_b <- symbolize(fit_b)
```

For a binomial model there is no residual variance on the observed 0/1 scale.
The model lives on the **latent (logit) scale**, where the residual variance
is the known constant $\pi^2/3$. So `icc()` reports a **latent-scale** ICC,
and labels it as such -- with the caption that keeps you honest:

```{r binomial-icc}
icc(sym_b)
```

This is exactly rptR's *link-scale* (a.k.a. latent-scale) repeatability. It is
the repeatability of the underlying propensity to survive, **not** a
proportion of variance in the realized 0/1 outcome. `symbolizer` refuses to
pretend otherwise.

**Takeaway.** Proportion and binary data get a repeatability too -- but on the
latent scale, with the residual fixed at $\pi^2/3$ (logit) or 1 (probit).

## 3. The same reading from a different engine: `glmmTMB`

`variance_partition()` and `icc()` read the fitted object, not the package
that produced it. Refit the same binomial model with `glmmTMB` and the
latent-scale repeatability is the same:

```{r glmmtmb-icc, eval = has_glmmTMB}
fit_b2 <- glmmTMB::glmmTMB(surv ~ temp_n + (1 | nest), data = dat_b,
                           family = binomial)
icc(symbolize(fit_b2))
```

The reading -- the quantity, the scale, the caption -- is identical whichever
engine you used. (For *Gaussian* fits the point estimate can differ slightly
between engines because `lme4::lmer` defaults to REML while `glmmTMB` defaults
to maximum likelihood; the variance partition is computed the same way from
whatever the fit reports.)

## 4. When the ICC is not defined -- and `symbolizer` says so

A single-number ICC needs a single residual variance on a known scale. When
that does not exist, `symbolizer` returns `NA` with a reason rather than a
misleading number:

* **Count models (Poisson, negative binomial)** have no fixed residual
  constant on the data scale, so the data-scale ICC is undefined.
* **Location-scale Gaussian** fits (where the residual SD itself varies across
  observations) have no single within-group variance.
* **More than one random-effect term**: the variance partition is still
  meaningful and shown, but a single-number ICC is not.

```{r poisson-refusal}
set.seed(3)
dat_p <- dat_b
dat_p$count <- rpois(nrow(dat_p), lambda = 3)
fit_p <- lme4::glmer(count ~ temp_n + (1 | nest), data = dat_p,
                     family = poisson)
icc(symbolize(fit_p))
```

The partition still shows the random-effect variance; only the share and the
single-number ICC are withheld.

**Takeaway.** `symbolizer` reports repeatability where it is well defined,
names the scale when it is on the latent scale, and refuses -- with a reason --
when it is not defined at all. That refusal is the feature.

## See also

* `variance_partition()` and `icc()` -- the accessors used here.
* `explain()` and `model_card()` -- both surface a "How the variation splits"
  section for any fit with random effects.
* The three-views widget (`as_html_three_views()`) carries the same
  "where does the variation live?" panel on its first tab.

### References

Nakagawa, S. & Schielzeth, H. (2010). Repeatability for Gaussian and
non-Gaussian data: a practical guide for biologists. *Biological Reviews*,
85, 935--956.

Stoffel, M. A., Nakagawa, S. & Schielzeth, H. (2017). rptR: repeatability
estimation and variance decomposition by generalized linear mixed-effects
models. *Methods in Ecology and Evolution*, 8, 1639--1644.
