---
title: "Reading factors, dummies, and interactions"
description: "How factors, dummy coding, contrasts, and interactions appear in the equation -- and the pitfalls that trip up interpretation."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading factors, dummies, and interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_drmTMB <- requireNamespace("drmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_drmTMB
)
set.seed(20260524)
```

> *Categorical predictors are where model output most often misleads a
> biologist — `sexmale` is not "average male body mass." This article is
> for anyone fitting models with factors and interactions; by the end
> you'll be able to read every dummy-coded coefficient and interaction
> term for what it actually says, and see how `symbolize()` spells out
> the contrast each one encodes.*

A biologist sees `body_mass ~ sex + temperature` and the model fits. The
output table lists a coefficient for `sexmale`. The obvious reading is
"average male body mass" — and the obvious reading is wrong. The
`sexmale` coefficient is something else. This vignette unpacks the
something else, then builds up to interactions, where the unpacking
becomes essential.

We walk through six steps that build on each other. Each step fits a
small `drmTMB` model, looks at the first few rows of the design matrix,
and asks `symbolize()` to read each coefficient on the biological scale.
The point is to make the dummy-variable, contrast, and interaction
machinery concrete enough that a reader can apply the same logic to
their own fits. A closing **Common pitfalls** section catches the
mistakes biologists most often make when reading these tables.

```{r libs}
library(symbolizer)
library(drmTMB)
```

## Step 1: A binary factor (`body_mass ~ sex`)

Two-level factor, no other predictors. `sex` is `female` or `male`. We
simulate `n = 120` observations with female mass centred at 30 g and
male mass centred at 35 g, then fit:

```{r step1-fit}
n <- 120
sex <- factor(sample(c("female", "male"), n, replace = TRUE))
dat1 <- data.frame(
  body_mass = rnorm(n, 30 + 5 * (sex == "male"), 3),
  sex       = sex
)

fit1 <- drmTMB(
  drm_formula(body_mass ~ sex, sigma ~ 1),
  family = gaussian(),
  data   = dat1
)

sym1 <- symbolize(
  fit1,
  symbols = c(body_mass = "W_i"),
  units   = c(body_mass = "g"),
  context = "binary factor: sex on body mass"
)
```

`expand(sym1)$X` is the design matrix the computer actually multiplies
against the coefficient vector. The first five rows tell the whole
story:

```{r step1-X}
head(expand(sym1)$X, 5)
```

Two columns. The `(Intercept)` column is all 1s, by design. The
`sexmale` column is 0 for females and 1 for males. There is no `sexfemale`
column — `female` is the **reference level**, absorbed into the
intercept.

The symbol dictionary marks the reference level explicitly:

```{r step1-symbols}
symbol_table(sym1)
```

Look at the `sex [factor]` row's description: `factor (female
[reference], male)`. That `[reference]` marker is the single most
important piece of context for reading every coefficient that follows.

Now the readings:

```{r step1-interp}
parameter_interpretation(sym1, scale = "biological")
```

Two coefficients on the mu submodel:

- The **intercept** estimate (`r if (exists("sym1")) round(sym1$fixed_effects$estimate[1], 2) else "30"`) is the expected `body_mass` for the **reference level**,
  i.e. females. It is *not* the overall mean of `body_mass` — if the
  sexes were unbalanced the overall mean would differ from the female
  mean.
- The **`sexmale` contrast** estimate (`r if (exists("sym1")) round(sym1$fixed_effects$estimate[2], 2) else "5"`) is the **difference** between male
  and female means. It is *not* the male mean. To get the male mean you
  add the intercept and the contrast: `intercept + sexmale` = `r if (exists("sym1")) round(sum(sym1$fixed_effects$estimate[1:2]), 2) else "35"`.

The biological reading in the table — *"Average body_mass differs
between male and the reference"* — is the same statement in templated
prose.

This is the rule the rest of the vignette builds on. Write it down.

**Takeaway.** A binary factor turns into one 0/1 dummy column. The
intercept is the reference-level mean; the contrast coefficient is the
*difference* from the reference, not the other group's mean.

## Step 2: A multi-level factor (`body_mass ~ site`)

`sex` had two levels and produced one dummy column. What happens with
four levels? `site` runs `A`, `B`, `C`, `D`:

```{r step2-fit}
site <- factor(sample(c("A", "B", "C", "D"), n, replace = TRUE))
site_mean <- c(A = 30, B = 32, C = 34, D = 27)
dat2 <- data.frame(
  body_mass = rnorm(n, site_mean[site], 3),
  site      = site
)

fit2 <- drmTMB(
  drm_formula(body_mass ~ site, sigma ~ 1),
  family = gaussian(),
  data   = dat2
)

sym2 <- symbolize(
  fit2,
  symbols = c(body_mass = "W_i"),
  units   = c(body_mass = "g"),
  context = "four-level factor: site on body mass"
)
```

The design matrix:

```{r step2-X}
head(expand(sym2)$X, 5)
```

Four levels, three dummy columns: `siteB`, `siteC`, `siteD`. Reference
is `A`, absorbed into the intercept. The pattern is mechanical: each
row has a 1 in the column matching its level, or zeros everywhere if it
is at the reference.

The symbol dictionary again carries the reference marker:

```{r step2-symbols}
symbol_table(sym2)
```

The `site [factor]` row reads `factor (A [reference], B, C, D)`. And
the coefficients are:

```{r step2-interp}
parameter_interpretation(sym2, scale = "biological")
```

Reading the table by hand:

- **Intercept** = mean of site `A`.
- **`siteB`** = mean of site `B` minus mean of site `A`.
- **`siteC`** = mean of site `C` minus mean of site `A`.
- **`siteD`** = mean of site `D` minus mean of site `A`.

None of the contrast coefficients are group means. Every one is a
*difference from the reference*. If you want the mean of site `C`, you
add `intercept + siteC`. If you want the difference between sites `B`
and `C`, you compute `siteC - siteB` by hand; the table does not give
it directly.

The rule generalises: **`k` levels become `k − 1` dummy columns, which
become `k − 1` contrasts, all measured against the reference**. A
factor with twelve levels would produce eleven dummies — never one per
level.

**Takeaway.** A `k`-level factor becomes `k − 1` dummy columns. Each
non-reference contrast is a *difference from the reference*, not a
group mean.

## Step 3: Factor plus continuous predictor (`body_mass ~ sex + body_size`)

Adding a continuous predictor changes what the intercept means. We
re-use `sex` and add `body_size`:

```{r step3-fit}
body_size <- runif(n, 50, 150)
dat3 <- data.frame(
  body_mass = rnorm(n, 30 + 5 * (sex == "male") + 0.2 * body_size, 3),
  sex       = sex,
  body_size = body_size
)

fit3 <- drmTMB(
  drm_formula(body_mass ~ sex + body_size, sigma ~ 1),
  family = gaussian(),
  data   = dat3
)

sym3 <- symbolize(
  fit3,
  symbols = c(body_mass = "W_i", body_size = "L_i"),
  units   = c(body_mass = "g",   body_size = "mm"),
  context = "additive: sex contrast + body_size slope"
)
```

Design matrix:

```{r step3-X}
head(expand(sym3)$X, 5)
```

Three coefficient columns: `(Intercept)`, `sexmale`, `body_size`. The
shape rule for `X` is now `R^{120 x 3}`. The symbol table makes that
concrete in its design-matrix row:

```{r step3-symbols}
symbol_table(sym3)
```

Look for the `(design_matrix)` row at the bottom: `dimension:
R^{n × p_mu}  (= R^{120 × 3})`. Three coefficients, three columns.

The coefficient readings:

```{r step3-interp}
parameter_interpretation(sym3, scale = "biological")
```

The single most important shift from Step 1 is what the intercept now
means. In Step 1 it was *the mean of females*. In Step 3 it is **the
expected `body_mass` for a female at `body_size = 0`**. The intercept
is the reference cell anchored to *both* the reference factor level
*and* the zero of every continuous predictor.

For this fit, `body_size = 0` is not biologically meaningful (no animal
has zero body size), which is why centring continuous predictors is
common practice. After centring, the intercept becomes "expected female
mass at average body size", which usually is meaningful.

The `sexmale` reading is unchanged in form — *"Average body_mass
differs between male and the reference"* — but its arithmetic is now
"at the same `body_size`". The contrast is a *vertical shift* of the
female regression line: the male line has the same slope but a
different intercept.

**Takeaway.** Adding a continuous predictor shifts what the intercept
means: it is now the expected response at the reference factor level
*and* at zero of every continuous predictor. The factor contrast is
still a difference from the reference, now read "at the same level of
the continuous predictor(s)".

### Three views: how the dummy column sits in the X matrix

For a `sex + body_size` fit the design matrix has **three columns** —
intercept, `sexmale` (0 for female, 1 for male), and `body_size` — and
the three-views widget shows exactly that on the matrix-with-data tab:
the `0`s and `1`s of the dummy column sit next to the continuous
`body_size` column, with the coefficient vector
$\boldsymbol{\beta} = (\beta_0,\,\beta_1,\,\beta_2)^\top$ multiplying
through:

```{r step3-three-views, eval = FALSE}
as_html_three_views(sym3)
```

```{r step3-three-views-live, echo = FALSE, results = "asis"}
if (requireNamespace("drmTMB", quietly = TRUE)) {
  invisible(as_html_three_views(sym3))
}
```

## Step 4: Continuous-by-factor interaction (`body_mass ~ sex * body_size`)

Step 3 forced the male and female regression lines to be parallel.
Adding an interaction lets them have different slopes:

```{r step4-fit}
dat4 <- data.frame(
  body_mass = rnorm(
    n,
    30 + 5 * (sex == "male") + 0.2 * body_size +
      0.05 * (sex == "male") * body_size,
    3
  ),
  sex       = sex,
  body_size = body_size
)

fit4 <- drmTMB(
  drm_formula(body_mass ~ sex * body_size, sigma ~ 1),
  family = gaussian(),
  data   = dat4
)

sym4 <- symbolize(
  fit4,
  symbols = c(body_mass = "W_i", body_size = "L_i"),
  units   = c(body_mass = "g",   body_size = "mm"),
  context = "interaction: sex slope on body_size"
)
```

`sex * body_size` expands to `sex + body_size + sex:body_size`. The
design matrix gains a fourth column for the interaction:

```{r step4-X}
head(expand(sym4)$X, 5)
```

Four columns: `(Intercept)`, `sexmale`, `body_size`, `sexmale:body_size`.
The interaction column is the **product** of the `sexmale` dummy and
the continuous `body_size` — it is zero for every female (because the
`sexmale` dummy is zero), and equals `body_size` for every male.

The coefficient table:

```{r step4-fixed}
sym4$fixed_effects[, c("term_label", "role", "contrast_level", "estimate")]
```

Four coefficients to read. Let us call them
$\beta_0$ (intercept), $\beta_1$ (`sexmale`), $\beta_2$ (`body_size`),
$\beta_3$ (`sexmale:body_size`). The model is

$$\mathrm{E}(W_i) = \beta_0 + \beta_1 \cdot \mathrm{male}_i + \beta_2 \cdot L_i + \beta_3 \cdot \mathrm{male}_i \cdot L_i.$$

Split it into the two regression lines by setting the `sexmale` dummy
to 0 or 1:

- **Females** (`sexmale = 0`): $\mathrm{E}(W_i) = \beta_0 + \beta_2 L_i$.
  Intercept $\beta_0$, slope $\beta_2$.
- **Males** (`sexmale = 1`): $\mathrm{E}(W_i) = (\beta_0 + \beta_1) +
  (\beta_2 + \beta_3) L_i$. Intercept $\beta_0 + \beta_1$, slope
  $\beta_2 + \beta_3$.

So:

- $\beta_2$ is the **female slope on `body_size`**.
- $\beta_2 + \beta_3$ is the **male slope**.
- $\beta_3$ — the interaction coefficient — is the **difference between
  the two slopes**, not the male slope itself.
- $\beta_1$ — the `sexmale` contrast — is the **difference in
  intercepts** at `body_size = 0`. It is no longer "the average male
  mass" *or* "the male-female mass difference at average body size".

`parameter_interpretation()` reads all four templated rows, including
the dedicated continuous-by-factor interaction reading:

```{r step4-interp}
parameter_interpretation(sym4, scale = "biological")
```

The interaction row carries its own templated reading. It is
ecologically the most interesting coefficient in the fit — the slope
of `body_mass` on `body_size` *depends on* `sex`, which is exactly what
"interaction" means.

**Takeaway.** A continuous-by-factor interaction lets each factor level
have its own slope. The interaction coefficient is the *difference in
slopes* between the non-reference level and the reference. The bare
factor contrast is now the difference in intercepts at zero of the
continuous predictor.

## Step 5: Continuous-by-continuous interaction (`body_mass ~ temperature * body_size`)

A continuous-by-continuous interaction lets the slope of one continuous
predictor change with the level of another. There are no dummy columns
here, but the same arithmetic — *the interaction coefficient is a
difference of slopes* — still applies, just on a sliding rather than a
group-by-group basis.

```{r step5cc-fit}
n_cc <- 80
temperature <- runif(n_cc, 10, 30)
body_size_cc <- runif(n_cc, 50, 150)
dat5cc <- data.frame(
  body_mass   = rnorm(
    n_cc,
    20 + 0.1 * temperature + 0.15 * body_size_cc +
      0.01 * temperature * body_size_cc,
    3
  ),
  temperature = temperature,
  body_size   = body_size_cc
)

fit5cc <- drmTMB(
  drm_formula(body_mass ~ temperature * body_size, sigma ~ 1),
  family = gaussian(),
  data   = dat5cc
)

sym5cc <- symbolize(
  fit5cc,
  symbols = c(body_mass = "W_i", temperature = "T_i", body_size = "L_i"),
  units   = c(body_mass = "g",   temperature = "°C", body_size = "mm"),
  context = "continuous-by-continuous interaction"
)
```

`temperature * body_size` expands to
`temperature + body_size + temperature:body_size`. The design matrix has
four coefficient columns — no zeros and ones this time, just two real
numbers and their product:

```{r step5cc-X}
head(expand(sym5cc)$X, 5)
```

The four columns are `(Intercept)`, `temperature`, `body_size`, and
`temperature:body_size`. The interaction column is literally the product
of the other two predictor columns.

Call the four coefficients $\beta_0, \beta_1, \beta_2, \beta_3$ in
column order. The model is

$$\mathrm{E}(W_i) = \beta_0 + \beta_1 T_i + \beta_2 L_i + \beta_3 T_i L_i.$$

Re-arrange to isolate the slope of `temperature`:

$$\mathrm{E}(W_i) = \beta_0 + (\beta_1 + \beta_3 L_i) T_i + \beta_2 L_i.$$

The slope of `temperature` is $\beta_1 + \beta_3 L_i$ — it depends on
`body_size`. Reading the four coefficients in those terms:

- **$\beta_0$ (intercept)** = expected `body_mass` when
  `temperature = 0` and `body_size = 0`.
- **$\beta_1$ (`temperature`)** = the `temperature` slope *at*
  `body_size = 0`. Not "the temperature slope". The slope at one
  specific value of `body_size`.
- **$\beta_2$ (`body_size`)** = the `body_size` slope *at*
  `temperature = 0`. Same reasoning.
- **$\beta_3$ (`temperature:body_size`)** = how much the `temperature`
  slope changes per one-unit increase in `body_size`. Equivalently, how
  much the `body_size` slope changes per one-unit increase in
  `temperature`. The two readings are symmetric.

For the cont-by-factor interaction in Step 4 we could call
`group_slopes(sym, continuous = "body_size")` to see one slope per sex.
The continuous-by-continuous equivalent asks for the slope of
`temperature` at *specific values* of `body_size`:

```{r step5cc-group-slopes}
group_slopes(
  sym5cc,
  continuous = "temperature",
  at = list(body_size = c(50, 100, 150))
)
```

Three rows, each the `temperature` slope at a different `body_size`,
with confidence intervals. The slope rises as `body_size` rises — the
sign of $\beta_3$ tells you whether the interaction amplifies or
dampens the `temperature` effect.

**Takeaway.** A continuous-by-continuous interaction means *the slope
of one continuous predictor changes with the level of another*. Use
`group_slopes(sym, continuous = ..., at = list(other = ...))` to read
the slope at specific values, the same way `group_means()` and
`group_slopes(at = factor)` work for categorical strata.

## Step 6: Factor-by-factor interaction (`body_mass ~ site * sex`)

The hardest case: two factors interacting. `site` has four levels and
`sex` has two, so the fitted-cell table has `4 × 2 = 8` entries. We
generate data with site-specific male offsets so the interaction is not
zero:

```{r step6-fit}
site_offset_male <- c(A = 0, B = 1.5, C = -1.0, D = 2.5)
dat5 <- data.frame(
  body_mass = rnorm(
    n,
    30 + site_mean[site] - 30 + 5 * (sex == "male") +
      site_offset_male[site] * (sex == "male"),
    3
  ),
  site = site,
  sex  = sex
)

fit5 <- drmTMB(
  drm_formula(body_mass ~ site * sex, sigma ~ 1),
  family = gaussian(),
  data   = dat5
)

sym5 <- symbolize(
  fit5,
  symbols = c(body_mass = "W_i"),
  units   = c(body_mass = "g"),
  context = "two factors interacting: site and sex"
)
```

`site * sex` expands to `site + sex + site:sex` and produces eight
coefficient columns. The first five rows:

```{r step6-X}
head(expand(sym5)$X, 5)
```

Eight columns: `(Intercept)`, `siteB`, `siteC`, `siteD`, `sexmale`,
`siteB:sexmale`, `siteC:sexmale`, `siteD:sexmale`. Two references —
`A` for site, `female` for sex — are absorbed into the intercept. The
interaction columns are products of the site dummy and the `sexmale`
dummy: a `B-male` row has 1s in both `siteB` and `siteB:sexmale`.

The coefficient table:

```{r step6-fixed}
sym5$fixed_effects[, c("term_label", "role", "contrast_level", "estimate")]
```

Reading the eight coefficients in cell terms (write
$\bar W_{s,x}$ for the population mean at site `s`, sex `x`):

- **Intercept** = $\bar W_{A,\mathrm{female}}$ — site `A`, female.
- **`siteB`**, **`siteC`**, **`siteD`** = site mean minus site `A` mean,
  *for females*: $\bar W_{B,\mathrm{female}} - \bar W_{A,\mathrm{female}}$, and so on.
- **`sexmale`** = male minus female *at site `A`*: $\bar W_{A,\mathrm{male}} -
  \bar W_{A,\mathrm{female}}$.
- **`siteB:sexmale`** is the part that needs the most care. It is
  $(\bar W_{B,\mathrm{male}} - \bar W_{B,\mathrm{female}}) -
   (\bar W_{A,\mathrm{male}} - \bar W_{A,\mathrm{female}})$.

That last line is the *difference of two differences*. The male-female
gap at site `B`, minus the male-female gap at site `A`. If
`siteB:sexmale > 0`, the male advantage is *bigger* at site `B` than
at site `A`. If it is negative, the male advantage is *smaller* (or
reversed) at `B` compared to `A`.

This is what "the effect of sex depends on site" means in coefficients.
It is also why a factor-by-factor interaction is hard to read off a
summary table without the cell-mean translation.

The biological readings `parameter_interpretation()` carries cover the
intercept and the per-term contrasts; interaction-row prose is shipped
via the `interpretation-templates.csv` and `group_slopes()` / `group_means()`
gives you the cell-mean translation on demand:

```{r step6-interp}
parameter_interpretation(sym5, scale = "biological")
```

Each of the three interaction rows carries its own templated
*difference-of-differences* reading; `group_means()` gives the cell-mean
translation on demand.

**Takeaway.** A factor-by-factor interaction coefficient is a
*difference of differences*: how much the gap between two non-reference
cells differs from the corresponding gap in the reference rows. The
bare factor contrasts now apply *only* at the other factor's reference
level.

## Common pitfalls

The six steps above cover the mechanics. The six pitfalls below cover
the *readings* — every one is a sentence we have heard a biologist say
out loud about their own fit. Each pitfall has the same structure:
**Symptom** (what goes wrong), **Diagnosis** (why), a code block showing
the wrong reading then the right one, and a one-sentence **Rule of
thumb** to carry away.

### Pitfall 1: The intercept is not "the average response"

**Symptom.** A reviewer asks what the intercept means; the author
answers "the average body mass". The number is way off compared to
`mean(dat$body_mass)`, and now everyone is confused.

**Diagnosis.** With `body_mass ~ sex + body_size`, the intercept is the
expected response *at the reference level of every factor and at zero on
every continuous predictor*. For a fit with `sex = factor(female, male)`
that means "expected `body_mass` for a FEMALE at `body_size = 0`". It is
not the overall mean unless the sexes are balanced and `body_size` is
already centred at zero.

```{r pitfall1-fit}
n_p1 <- 80
sex_p1 <- factor(sample(c("female", "male"), n_p1, replace = TRUE))
bs_p1  <- runif(n_p1, 50, 150)
dat_p1 <- data.frame(
  body_mass = rnorm(n_p1, 30 + 5 * (sex_p1 == "male") + 0.2 * bs_p1, 3),
  sex       = sex_p1,
  body_size = bs_p1
)
fit_p1 <- drmTMB(
  drm_formula(body_mass ~ sex + body_size, sigma ~ 1),
  family = gaussian(), data = dat_p1
)
sym_p1 <- symbolize(fit_p1, context = "intercept reading")

# WRONG: read the intercept as the grand mean.
#   "the average body mass is 12.0 g" -- it is not.
# RIGHT: read the intercept as the reference cell.
#   symbol_table() flags the reference level so this is auditable.
symbol_table(sym_p1)
```

**Rule of thumb.** *Read the intercept as the expected response at the
reference factor level and zero on every continuous predictor.*

### Pitfall 2: A factor contrast is not the group's mean

**Symptom.** A paper sentence says "male body mass is 5 g (sexmale
coefficient)". A reader checks `mean(body_mass[sex == "male"])` and
gets ~35 g. The sentence is wrong.

**Diagnosis.** The `sexmale` coefficient is the *difference* between
the male and female cell means, not the male mean. The male mean is
`(Intercept) + sexmale`. `group_means(sym, by = "sex")` returns each
cell mean directly with a confidence band, which is what most readers
actually wanted.

```{r pitfall2-fit}
n_p2 <- 80
sex_p2 <- factor(sample(c("female", "male"), n_p2, replace = TRUE))
dat_p2 <- data.frame(
  body_mass = rnorm(n_p2, 30 + 5 * (sex_p2 == "male"), 3),
  sex       = sex_p2
)
fit_p2 <- drmTMB(
  drm_formula(body_mass ~ sex, sigma ~ 1),
  family = gaussian(), data = dat_p2
)
sym_p2 <- symbolize(fit_p2, context = "contrast vs cell mean")

# WRONG: read the sexmale coefficient as the male mean.
sym_p2$fixed_effects[, c("term_label", "estimate")]

# RIGHT: ask for the cell means directly.
group_means(sym_p2, by = "sex")
```

**Rule of thumb.** *Coefficients on factor levels are differences from
the reference; use `group_means(sym)` to see the group means
themselves.*

### Pitfall 3: An interaction is not "the effect of A on B"

**Symptom.** A methods section says "the interaction tests whether sex
has an effect on body size". That is a category error — sex does not
*have an effect on* body size; the interaction tests whether the slope
of one predictor depends on the level of the other.

**Diagnosis.** For `body_mass ~ sex * body_size`, the
`sexmale:body_size` coefficient is the *difference between the male and
female slopes of `body_size`*, not "the effect of sex on body_size".
The honest reading is "the slope of `body_size` is X for females and
X + Y for males". `group_slopes(sym, continuous = "body_size")`
returns the per-group slopes directly.

```{r pitfall3-fit}
n_p3 <- 80
sex_p3 <- factor(sample(c("female", "male"), n_p3, replace = TRUE))
bs_p3  <- runif(n_p3, 50, 150)
dat_p3 <- data.frame(
  body_mass = rnorm(
    n_p3,
    30 + 5 * (sex_p3 == "male") + 0.2 * bs_p3 +
      0.05 * (sex_p3 == "male") * bs_p3,
    3
  ),
  sex       = sex_p3,
  body_size = bs_p3
)
fit_p3 <- drmTMB(
  drm_formula(body_mass ~ sex * body_size, sigma ~ 1),
  family = gaussian(), data = dat_p3
)
sym_p3 <- symbolize(fit_p3, context = "interaction wording")

# WRONG: "the sexmale:body_size coefficient is the effect of sex on body_size."
# RIGHT: it is the difference in body_size slopes between male and female.
group_slopes(sym_p3, continuous = "body_size")
```

**Rule of thumb.** *The interaction coefficient is the difference
between the two effects, not either effect alone; use
`group_slopes(sym, continuous = ...)` to read each group's slope.*

### Pitfall 4: Wald CIs can be too narrow with few groups

**Symptom.** A fit with `(1 | site)` and only six sites reports very
tight confidence bands on the fixed effects. The bands look more
precise than the data could justify, and a reviewer asks how they were
computed.

**Diagnosis.** drmTMB's default is `ci_method = "wald"`, which uses an
asymptotic normal approximation. When `sd(site)` is estimated from
only a handful of groups, the asymptotic regime has not kicked in, and
Wald intervals over-state precision. Profile-likelihood intervals are
honest but slower to compute. The flag lives on `symbolize()` so the
choice is recorded on `x$metadata$ci_method` and propagates to every
downstream reading.

```{r pitfall4-skeleton, eval = FALSE}
# WRONG: take the default Wald CI when only a few groups inform sd(site).
sym_wald <- symbolize(fit, ci_method = "wald")

# RIGHT: ask for a profile CI — slower, but honest.
sym_prof <- symbolize(fit, ci_method = "profile")
```

**Rule of thumb.** *When `sd(group)` is estimated from few groups, pass
`ci_method = "profile"` to `symbolize()` for an honest interval.*

### Pitfall 5: Dropping the intercept doesn't always do what you think

**Symptom.** A user wants every cell mean directly, types
`y ~ 0 + sex + site` thinking it will give all `sex × site` cell
means, and then is surprised when the coefficients still look like
contrasts.

**Diagnosis.** `y ~ 0 + sex` with one factor *does* give cell means
directly (one coefficient per level). But adding a second predictor,
e.g. `y ~ 0 + sex + site`, gives "sex cell means PLUS site contrasts
from the reference site". The non-orthogonality between two factors
without an intercept is rarely what the reader wanted.
`group_means(sym, by = c("sex", "site"))` returns every cell mean
correctly regardless of how the formula was written.

```{r pitfall5-fit}
n_p5 <- 100
sex_p5  <- factor(sample(c("female", "male"), n_p5, replace = TRUE))
site_p5 <- factor(sample(c("A", "B", "C"), n_p5, replace = TRUE))
dat_p5 <- data.frame(
  body_mass = rnorm(n_p5, 30 + 5 * (sex_p5 == "male"), 3),
  sex       = sex_p5,
  site      = site_p5
)

# WRONG: drop the intercept and hope to read all 6 cell means off the table.
fit_p5_wrong <- drmTMB(
  drm_formula(body_mass ~ 0 + sex + site, sigma ~ 1),
  family = gaussian(), data = dat_p5
)

# RIGHT: keep the intercept in, fit ~ sex * site, then use group_means().
fit_p5_right <- drmTMB(
  drm_formula(body_mass ~ sex * site, sigma ~ 1),
  family = gaussian(), data = dat_p5
)
sym_p5 <- symbolize(fit_p5_right, context = "all cell means")
group_means(sym_p5, by = c("sex", "site"))
```

**Rule of thumb.** *With two or more factors, the safest way to get
all cell means is `group_means(sym, by = c(...))`, not a hand-built
intercept-less formula.*

### Pitfall 6: `poly(x, 2)` and `I(x^2)` are not the same

**Symptom.** A paper reports "the slope of body_size was 0.18 g/mm
(`poly(body_size, 2)1` coefficient)". The number happens to match an
expected linear slope, so it looks fine — but the column is the *first
orthogonal polynomial of `body_size`*, not the raw linear term.

**Diagnosis.** `poly(x, 2)` produces two ORTHOGONAL columns: a
combination of $x$ and $x^2$ scaled so the two columns are
uncorrelated. The first column is *not* "the slope of $x$". By
contrast `I(x^2)` puts the literal $x^2$ column into the design matrix
alongside a separate linear $x$ term, and now the linear coefficient
*is* the slope at $x = 0$. The two parameterisations carry the same
information but their coefficient *interpretations* differ. Use
whichever fits your story.

```{r pitfall6-fit}
n_p6 <- 80
bs_p6 <- runif(n_p6, -1, 1)
dat_p6 <- data.frame(
  body_mass = rnorm(n_p6, 30 + 0.2 * bs_p6 + 0.5 * bs_p6^2, 1),
  body_size = bs_p6
)

# WRONG: read the poly() column as a literal linear slope.
fit_p6_poly <- drmTMB(
  drm_formula(body_mass ~ poly(body_size, 2), sigma ~ 1),
  family = gaussian(), data = dat_p6
)
sym_p6_poly <- symbolize(fit_p6_poly, context = "orthogonal polynomial")

# RIGHT: use I(x^2) when you want a literal quadratic reading.
fit_p6_raw <- drmTMB(
  drm_formula(body_mass ~ body_size + I(body_size^2), sigma ~ 1),
  family = gaussian(), data = dat_p6
)
sym_p6_raw <- symbolize(fit_p6_raw, context = "raw quadratic")

symbol_table(sym_p6_poly)
symbol_table(sym_p6_raw)
```

**Rule of thumb.** *Use `poly(x, 2)` when you want orthogonal columns
(uncorrelated polynomial terms); use `I(x^2)` when you want a literal
quadratic interpretation.*

## One call: `explain_factors()`

Steps 1--6 unpack the coding by hand. In day-to-day use you can get the
whole story in a single call. `explain_factors()` states each factor's
coding scheme in plain language, then shows the per-group means and the
pairwise comparisons; for interactions it prints the cells or per-group
slopes directly, so you never have to add coefficients by hand.

```{r explain-factors, eval = requireNamespace("emmeans", quietly = TRUE)}
explain_factors(sym2)
```

The pairwise question --- *which sites differ from each **other**?*, not
just each versus the reference --- is answered by `group_contrasts()`. It
reports compatibility bands and never p-values; pass `adjust = "tukey"`
for simultaneous (family-wise) intervals.

```{r group-contrasts, eval = requireNamespace("emmeans", quietly = TRUE)}
group_contrasts(sym2, by = "site")
```

For a web-facing version, `as_html_factor_views(sym2)` renders the same
story as a four-tab interactive widget --- coding scheme, group means,
pairwise, interactions --- with Confidence-Eye uncertainty bands.

```{r factor-views-widget, results = "asis", eval = requireNamespace("emmeans", quietly = TRUE)}
as_html_factor_views(sym2)
```

## Closing: a checklist for reading any model with factors

When you see a coefficient table from a model that contains factors,
apply this three-step audit before you interpret anything.

1. **Identify the reference levels.** Run `symbol_table(sym)` and look
   for the `[reference]` marker in each factor's `description`. By
   default R uses the alphabetically-first level, but
   `relevel()` or `contrasts()` can shift that without changing the
   apparent variable name. The reference level is the silent half of
   every contrast.

2. **The intercept is the reference cell.** It is the expected response
   when every factor sits at its reference level *and* every continuous
   predictor sits at zero. Centring continuous predictors makes the
   intercept biologically meaningful; until then it is the cell at
   `body_size = 0`, `temperature = 0`, and so on.

3. **Every contrast and interaction coefficient is a difference from a
   reference.**
   - A bare factor contrast is a difference between two cells of
     *that* factor, at the reference levels of every other factor.
   - An interaction coefficient is a *difference of differences*: the
     gap at the non-reference combination, minus the gap in the
     reference rows.

   No coefficient (except possibly the intercept after centring) is a
   group mean by itself. To get a group mean, add the relevant
   intercept-plus-contrast(s) by hand.

The point of `symbolize()` is to make every layer of this translation
visible: the design matrix shows the dummies, the symbol table marks
the references, and `parameter_interpretation()` reads each templated
coefficient on the biological scale — including the continuous-by-factor
and factor-by-factor interaction rows. `group_means()` / `group_slopes()`
give the cell-mean and per-level-slope translations on demand.
