---
title: "Latent variables in ecology: a gllvmTMB worked example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Latent variables in ecology: a gllvmTMB worked example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_pkgs <- requireNamespace("gllvmTMB", quietly = TRUE) &&
            requireNamespace("drmTMB",   quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_pkgs
)
```

## 1. The biological question

A behavioural ecologist measures five traits — boldness in a novel arena,
exploration in a maze, aggression to a mirror, activity in the home cage,
and time-out-of-shelter — on each of forty fish, three times each. Two
themes keep coming up in the data. Bolder fish also tend to be more
exploratory. Active fish tend to spend more time out of shelter. The
question is not whether any single pair of traits correlates, but whether
the **whole battery of traits is organised by a small number of stable
axes of among-individual variation** — the things the syndromes literature
calls *behavioural syndromes* (Sih et al. 2004) or, in a broader trait
context, *phenotypic integration* (Pigliucci 2003).

This is the natural job for a generalised linear latent variable model
(GLLVM). Each trait gets its own intercept and its own residual variance.
A small number of unobserved latent variables — two, say — describe the
shared structure: each trait *loads* on each latent axis, and an
individual's score on each axis is what makes their trait values move
together. The output a biologist wants is the loading matrix
$\boldsymbol{\Lambda}_B$, the proportion of each trait's
between-individual variance that the shared axes explain (the
*communality*), and the residual trait-trait correlation matrix.

**Takeaway.** The biological question is *what stable axes of
among-individual variation organise these traits*, and the answer is a
loading matrix plus a communality vector.

## 2. The data

We simulate from a known truth so the biological reading at the end can
be checked. The truth puts traits 1-3 on a first axis (the
*boldness-exploration-aggression* syndrome) and traits 4-5 on a second
axis (the *activity-shelter-use* syndrome), with extra
trait-specific between-individual variance on top.

```{r data}
library(symbolizer)
library(gllvmTMB)

set.seed(20260523)
n_ind  <- 40   # individuals
n_tr   <- 5    # traits
n_sess <- 3    # repeated sessions per individual

# True between-individual loading matrix Lambda_B (5 traits x 2 latents)
Lam_B <- matrix(c(
  1.0, 0.0,   # trait 1 loads on axis 1 only
  0.7, 0.0,   # trait 2 loads on axis 1
  0.6, 0.0,   # trait 3 loads on axis 1
  0.0, 1.0,   # trait 4 loads on axis 2 only
  0.0, 0.7    # trait 5 loads on axis 2
), nrow = n_tr, byrow = TRUE)

sim <- simulate_site_trait(
  n_sites               = n_ind,
  n_species             = n_sess,
  n_traits              = n_tr,
  n_predictors          = 1,
  mean_species_per_site = n_sess,
  sigma2_eps            = 0.3,
  Lambda_B              = Lam_B,
  psi_B                 = rep(0.4, n_tr),
  seed                  = 1
)

# Relabel for syndromes vocabulary
dat <- sim$data
dat$individual <- dat$site          # unit
dat$session    <- dat$species       # repeated session
dat$obs        <- dat$site_species  # one row per (individual, session)
head(dat, 6)
```

`simulate_site_trait()` is the gllvmTMB recovery-test simulator. It uses
site/species names by default; we rename them to the syndromes
vocabulary. Each row is one (individual, session, trait) cell.

**Takeaway.** Forty individuals, three sessions each, five traits — long
format, one row per (individual, session, trait).

## 3. The model in symbols

The same model can be written two ways. The **long form** matches the
way `gllvmTMB()` actually fits the data — one row per (individual,
occasion, trait) cell, with a scalar response. The **wide form**
matches the factor-analytic textbooks — one row per observation, with a
trait-vector response. The two panels below place them side by side.

```{r model-long-wide, results = 'asis', echo = FALSE, eval = TRUE}
cat('<div style="display: flex; gap: 1rem; flex-wrap: wrap; margin: 1em 0;">

<div style="flex: 1; min-width: 320px;">

#### Long form

Each row of the data is one (individual, occasion, trait) cell. The
scalar response is $y_{ijt}$.

$$
y_{ijt} \\mid \\boldsymbol{\\mu},\\, \\boldsymbol{\\Lambda},\\, \\mathbf{z}_i,\\, \\boldsymbol{\\Psi}
  \\sim \\mathcal{N}\\!\\left(\\mu_t + \\sum_{k=1}^{d_B} \\lambda_{tk}\\, z_{ik},\\; \\sigma^2_{e}\\right)
$$

Native to `gllvmTMB`\'s formula interface. Each row\'s contribution is a
scalar Gaussian.

</div>

<div style="flex: 1; min-width: 320px;">

#### Wide form

Each row is one observation; the response is a $T$-vector
$\\mathbf{Y}_{ij}$.

$$
\\mathbf{Y}_{ij} \\mid \\boldsymbol{\\mu},\\, \\boldsymbol{\\Lambda},\\, \\mathbf{z}_i,\\, \\boldsymbol{\\Psi}
  \\sim \\mathcal{MN}\\!\\left(\\boldsymbol{\\mu} + \\boldsymbol{\\Lambda}\\,\\mathbf{z}_i,\\; \\boldsymbol{\\Sigma}\\right),
\\quad \\boldsymbol{\\Sigma} = \\boldsymbol{\\Lambda}\\boldsymbol{\\Lambda}^{\\!\\top} + \\boldsymbol{\\Psi}
$$

Native to factor-analytic textbooks. The covariance decomposition
$\\boldsymbol{\\Sigma} = \\boldsymbol{\\Lambda}\\boldsymbol{\\Lambda}^{\\!\\top} + \\boldsymbol{\\Psi}$
is the headline result.

</div>

</div>
')
```

*Bridge.* The two forms describe the same model. The wide form makes
the covariance decomposition explicit; the long form is what
`gllvmTMB()` actually fits.

Index $i$ ranges over individuals, $j$ over occasions, $t$ (or $j$ in
the wide form) over traits, $k$ over latent axes. $\mu_t$ is the trait
$t$ mean; $\lambda_{tk}$ is the loading of trait $t$ on axis $k$;
$z_{ik}$ is individual $i$'s position on axis $k$;
$\boldsymbol{\Psi} = \mathrm{diag}(\psi_1^2, \ldots, \psi_T^2)$ collects
the trait-specific between-individual uniquenesses; $\sigma^2_e$ is
the residual scalar variance in the long form.

**Takeaway.** Long form is one scalar equation per (individual,
occasion, trait); wide form is one $T$-vector equation per observation,
with the covariance decomposition written explicitly. Same model, two
notations.

## 4. The fit

`gllvmTMB()` takes a long-format frame and a glmmTMB-style formula.
`latent(0 + trait | individual, d = 2)` is the between-individual
reduced-rank decomposition with two latent axes. `unique(0 + trait |
individual)` adds the trait-specific between-individual uniqueness
$\mathbf{S}$. `unique(0 + trait | obs)` absorbs the within-individual
(across-session) residual.

```{r fit}
fit <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | individual, d = 2) +
          unique(0 + trait | individual) +
          unique(0 + trait | obs),
  data     = dat,
  family   = gaussian(),
  trait    = "trait",
  unit     = "individual",
  unit_obs = "obs",
  cluster  = "session",
  silent   = TRUE
)
class(fit)
```

The fit returns a `gllvmTMB_multi` object — the multi-trait variant.
Convergence takes about a second on this dataset; the recovery-test
simulator is intentionally well-posed.

**Takeaway.** One `gllvmTMB()` call. The formula carries four things:
trait intercepts, a 2-axis latent decomposition, between-individual
uniquenesses, and a within-individual residual.

## 5. Symbolize it

`symbolize.gllvmTMB` is a v0.4 First slice still being wired in (Agent
G1's parallel work). The vignette is structured so it works either
way: the call is wrapped in `tryCatch()` and downstream chunks guard on
the return value.

```{r symbolize, error = TRUE}
sym <- tryCatch(
  symbolize(
    fit,
    symbols = c(value = "y_{ij}", trait = "j", individual = "i"),
    context = "behavioural-syndromes GLLVM"
  ),
  error = function(e) {
    cat(
      "Note: symbolize.gllvmTMB is a v0.4 First slice still being wired in.\n",
      "Once it lands, calling symbolize() here will return a structured\n",
      "symbolized_model that the same accessors below consume.\n",
      sep = ""
    )
    NULL
  }
)
```

When the method lands, the returned `symbolized_model` will carry the
same surfaces as the `drmTMB` examples in the other vignettes:

- `equations(sym)` will return one row for the conditional
  distribution, one for the linear predictor, and one for the latent
  variable distribution $\mathbf{z}_i \sim \mathcal{N}(\mathbf{0},
  \mathbf{I}_{d_B})$.
- `symbol_table(sym)` will list each symbol — $\mathbf{Y}$,
  $\boldsymbol{\Lambda}$, $\mathbf{Z}$, $\boldsymbol{\mu}$, $\mathbf{S}$
  — with its abstract dimension (`\mathbb{R}^{n \times T}`,
  `\mathbb{R}^{T \times d_B}`, `\mathbb{R}^{n \times d_B}`) and its
  concrete dimension for this fit (`\mathbb{R}^{40 \times 5}`,
  `\mathbb{R}^{5 \times 2}`, `\mathbb{R}^{40 \times 2}`).
- `assumption_table(sym)` will state the standard-Gaussian prior on
  the latent variables, the trait-specific Gaussian residuals, and the
  lower-triangular identification convention on $\boldsymbol{\Lambda}$.
- `parameter_interpretation(sym)` will read each loading as the
  contribution of a one-SD shift on axis $k$ to trait $j$.

For now we read the loadings directly from gllvmTMB's own accessors.

**Takeaway.** The structural object is on the v0.4 roadmap. The
accessors in section 6 are the gllvmTMB-side surface that
`symbolize.gllvmTMB` will mirror.

## 6. Reading the latent axes biologically

The raw loading matrix is identified only up to rotation when $d_B >
1$, which is the usual factor-analysis story. A varimax rotation makes
the axes interpretable: each axis tries to load strongly on a small set
of traits and near zero on the rest.

```{r loadings}
getLoadings(fit, rotate = "varimax")
```

Reading down the columns of the rotated loading matrix, the first axis
loads on traits 4 and 5 (the *activity-shelter-use* syndrome in the
simulated truth); the second axis loads on traits 1, 2, and 3 (the
*boldness-exploration-aggression* syndrome). The numerical labels swap
because varimax does not preserve column order — but the *block
structure* matches the truth.

The communality is the proportion of each trait's between-individual
variance that the shared axes capture:

```{r communality}
extract_communality(fit)
```

Traits with communality near 1 are well-explained by the shared axes;
traits with low communality have most of their between-individual
variability in the trait-specific uniqueness $s_j^2$.

The residual between-individual correlation matrix on the latent scale
falls out of `extract_correlations()`:

```{r correlations}
co <- extract_correlations(fit)
co_B <- co[co$tier == "B", c("trait_i", "trait_j", "correlation", "lower", "upper")]
co_B
```

The two within-syndrome correlations (traits 1-2-3 among themselves and
traits 4-5 among themselves) are clearly positive; the
across-syndrome pairs are smaller.

**Takeaway.** Two rotated axes pull apart the two simulated syndromes;
communalities tell you which traits live in the shared structure;
trait-trait correlations show the syndrome blocks directly.

## 7. Where the symbolized story helps

With five traits and two latents, the per-element index form
$y_{ij} = \mu_j + \sum_{k} \lambda_{jk} z_{ik} + \epsilon_{ij}$ requires
the reader to mentally instantiate fifty equations (forty individuals
times five traits, with two latents apiece). The matrix block

$$
\mathbf{Y} = \mathbf{1}\boldsymbol{\mu}^{\!\top} + \mathbf{Z}\,\boldsymbol{\Lambda}^{\!\top} + \mathbf{E}
$$

says exactly the same thing in one line. A biologist can *see* that
$\mathbf{Z}$ stores one row of latent scores per individual,
$\boldsymbol{\Lambda}$ stores one row of loadings per trait, and the
product is the deterministic part of the trait matrix.

This is the educator-first surface symbolizer is built around: the
structural object always carries both notations, and the
`symbol_table()` always carries both abstract and concrete dimensions.
For latent-variable models, where the matrix form is genuinely the more
readable one, that bridge is where the package earns its keep.

**Takeaway.** Index form scales like $n \times T$; matrix form scales
like one block. For GLLVMs the matrix form is the reader-friendly one.

## 8. What v0.4 unlocks

The accessors in section 6 — `getLoadings()`, `extract_communality()`,
`extract_correlations()` — are gllvmTMB-side. v0.4 wraps them in the
symbolizer surface so the same code that builds the Methods section
of a location-scale paper also builds the Methods section of a
latent-variable paper. Three concrete payoffs:

1. **Two-model comparison.** `compare_symbolic(fit_d1, fit_d2)` will
   diff the structural specifications of $d_B = 1$ versus $d_B = 2$
   side by side — which lines change, which symbols change, which
   assumption rows change — so a reader can audit the model choice.
2. **Automated syndromes reading.**
   `parameter_interpretation(sym, scale = "biological")` will produce
   one row per (trait, axis) pair, reading each $\lambda_{jk}$ on the
   natural scale of trait $j$.
3. **Uncertainty on $\boldsymbol{\Lambda}$.** `bootstrap_Sigma()` from
   gllvmTMB already returns parametric-bootstrap draws of the loading
   matrix; v0.4 surfaces them through `symbolize()` so the
   `symbol_table()` rows for $\boldsymbol{\Lambda}$ and
   $\boldsymbol{\Sigma}_B$ carry confidence regions, not just point
   estimates.

**Takeaway.** v0.4 is the slice that makes latent-variable models
first-class in symbolizer; the gllvmTMB side already returns the
structural pieces.
