---
title: "Latent variables in ecology: a gllvmTMB worked example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Latent variables in ecology: a gllvmTMB worked example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_pkgs <- requireNamespace("gllvmTMB", quietly = TRUE) &&
            requireNamespace("drmTMB",   quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_pkgs
)
```

## 1. The biological question

A behavioural ecologist measures five traits — boldness in a novel arena,
exploration in a maze, aggression to a mirror, activity in the home cage,
and time-out-of-shelter — on each of forty fish, three times each. Two
themes keep coming up in the data. Bolder fish also tend to be more
exploratory. Active fish tend to spend more time out of shelter. The
question is not whether any single pair of traits correlates, but whether
the **whole battery of traits is organised by a small number of stable
axes of among-individual variation** — the things the syndromes literature
calls *behavioural syndromes* (Sih et al. 2004) or, in a broader trait
context, *phenotypic integration* (Pigliucci 2003).

This is the natural job for a generalised linear latent variable model
(GLLVM). Each trait gets its own intercept and its own residual variance.
A small number of unobserved latent variables — two, say — describe the
shared structure: each trait *loads* on each latent axis, and an
individual's score on each axis is what makes their trait values move
together. The output a biologist wants is the loading matrix
$\boldsymbol{\Lambda}_B$, the proportion of each trait's
between-individual variance that the shared axes explain (the
*communality*), and the residual trait-trait correlation matrix.

**Takeaway.** The biological question is *what stable axes of
among-individual variation organise these traits*, and the answer is a
loading matrix plus a communality vector.

## 2. The data

We simulate from a known truth so the biological reading at the end can
be checked. The truth puts traits 1-3 on a first axis (the
*boldness-exploration-aggression* syndrome) and traits 4-5 on a second
axis (the *activity-shelter-use* syndrome), with extra
trait-specific between-individual variance on top.

```{r data}
library(symbolizer)
library(gllvmTMB)

set.seed(20260523)
n_ind  <- 40   # individuals
n_tr   <- 5    # traits
n_sess <- 3    # repeated sessions per individual

# True between-individual loading matrix Lambda_B (5 traits x 2 latents)
Lam_B <- matrix(c(
  1.0, 0.0,   # trait 1 loads on axis 1 only
  0.7, 0.0,   # trait 2 loads on axis 1
  0.6, 0.0,   # trait 3 loads on axis 1
  0.0, 1.0,   # trait 4 loads on axis 2 only
  0.0, 0.7    # trait 5 loads on axis 2
), nrow = n_tr, byrow = TRUE)

sim <- simulate_site_trait(
  n_sites               = n_ind,
  n_species             = n_sess,
  n_traits              = n_tr,
  n_predictors          = 1,
  mean_species_per_site = n_sess,
  sigma2_eps            = 0.3,
  Lambda_B              = Lam_B,
  psi_B                 = rep(0.4, n_tr),
  seed                  = 1
)

# Relabel for syndromes vocabulary
dat <- sim$data
dat$individual <- dat$site          # unit
dat$session    <- dat$species       # repeated session
dat$obs        <- dat$site_species  # one row per (individual, session)
head(dat, 6)
```

`simulate_site_trait()` is the gllvmTMB recovery-test simulator. It uses
site/species names by default; we rename them to the syndromes
vocabulary. Each row is one (individual, session, trait) cell.

**Takeaway.** Forty individuals, three sessions each, five traits — long
format, one row per (individual, session, trait).

## 3. The model in symbols

The same model can be written two ways. The **long form** matches the
way `gllvmTMB()` actually fits the data — one row per (individual,
occasion, trait) cell, with a scalar response. The **wide form**
matches the factor-analytic textbooks — one row per observation, with a
trait-vector response. The two panels below place them side by side.

```{r model-long-wide, results = 'asis', echo = FALSE, eval = TRUE}
cat('<div style="display: flex; gap: 1rem; flex-wrap: wrap; margin: 1em 0;">

<div style="flex: 1; min-width: 320px;">

#### Long form

Each row of the data is one (individual, occasion, trait) cell. The
scalar response is $y_{ijt}$.

$$
y_{ijt} \\mid \\boldsymbol{\\mu},\\, \\boldsymbol{\\Lambda},\\, \\mathbf{z}_i,\\, \\boldsymbol{\\Psi}
  \\sim \\mathcal{N}\\!\\left(\\mu_t + \\sum_{k=1}^{d_B} \\lambda_{tk}\\, z_{ik},\\; \\sigma^2_{e}\\right)
$$

Native to `gllvmTMB`\'s formula interface. Each row\'s contribution is a
scalar Gaussian.

</div>

<div style="flex: 1; min-width: 320px;">

#### Wide form

Each row is one observation; the response is a $T$-vector
$\\mathbf{Y}_{ij}$.

$$
\\mathbf{Y}_{ij} \\mid \\boldsymbol{\\mu},\\, \\boldsymbol{\\Lambda},\\, \\mathbf{z}_i,\\, \\boldsymbol{\\Psi}
  \\sim \\mathcal{MN}\\!\\left(\\boldsymbol{\\mu} + \\boldsymbol{\\Lambda}\\,\\mathbf{z}_i,\\; \\boldsymbol{\\Sigma}\\right),
\\quad \\boldsymbol{\\Sigma} = \\boldsymbol{\\Lambda}\\boldsymbol{\\Lambda}^{\\!\\top} + \\boldsymbol{\\Psi}
$$

Native to factor-analytic textbooks. The covariance decomposition
$\\boldsymbol{\\Sigma} = \\boldsymbol{\\Lambda}\\boldsymbol{\\Lambda}^{\\!\\top} + \\boldsymbol{\\Psi}$
is the headline result.

</div>

</div>
')
```

*Bridge.* The two forms describe the same model. The wide form makes
the covariance decomposition explicit; the long form is what
`gllvmTMB()` actually fits.

Index $i$ ranges over individuals, $j$ over occasions, $t$ (or $j$ in
the wide form) over traits, $k$ over latent axes. $\mu_t$ is the trait
$t$ mean; $\lambda_{tk}$ is the loading of trait $t$ on axis $k$;
$z_{ik}$ is individual $i$'s position on axis $k$;
$\boldsymbol{\Psi} = \mathrm{diag}(\psi_1^2, \ldots, \psi_T^2)$ collects
the trait-specific between-individual uniquenesses; $\sigma^2_e$ is
the residual scalar variance in the long form.

**Takeaway.** Long form is one scalar equation per (individual,
occasion, trait); wide form is one $T$-vector equation per observation,
with the covariance decomposition written explicitly. Same model, two
notations.

## 4. The fit

`gllvmTMB()` takes a long-format frame and a glmmTMB-style formula.
`latent(0 + trait | individual, d = 2)` is the between-individual
reduced-rank decomposition with two latent axes. `unique(0 + trait |
individual)` adds the trait-specific between-individual uniqueness
$\boldsymbol{\Psi}_B$. The within-individual (across-session) residual
is absorbed into the row-level $\sigma^2_e$ that `gllvmTMB` carries
for the Gaussian family.

```{r fit}
fit <- gllvmTMB(
  value ~ 0 + trait +
          latent(0 + trait | individual, d = 2) +
          unique(0 + trait | individual),
  data     = dat,
  family   = gaussian(),
  trait    = "trait",
  unit     = "individual",
  unit_obs = "obs",
  cluster  = "session",
  silent   = TRUE
)
class(fit)
```

The fit returns a `gllvmTMB_multi` object — the multi-trait variant.
Convergence takes about a second on this dataset; the recovery-test
simulator is intentionally well-posed.

**Takeaway.** One `gllvmTMB()` call. The formula carries three things:
trait intercepts, a 2-axis between-individual latent decomposition,
and trait-specific between-individual uniquenesses. The row-level
$\sigma^2_e$ collects the within-individual residual; a separate
trait-specific within-individual uniqueness ($\boldsymbol{\Psi}_W$) is
on the roadmap but not in this first slice.

## 5. Symbolize it

`symbolize.gllvmTMB` covers the Gaussian and binomial latent-variable
families (First slice since v0.4 – v0.5):

```{r symbolize}
sym <- symbolize(
  fit,
  symbols = c(value = "y_{ij}", trait = "j", individual = "i"),
  context = "behavioural-syndromes GLLVM"
)
```

The returned `symbolized_model` carries the same surfaces as the
`drmTMB` examples in the other vignettes:

- `equations(sym)` returns one row for the conditional distribution,
  one for the linear predictor, and one for the latent variable
  distribution $\mathbf{z}_i \sim \mathcal{N}(\mathbf{0},
  \mathbf{I}_{d_B})$.
- `symbol_table(sym)` lists each symbol — $\mathbf{Y}$,
  $\boldsymbol{\Lambda}$, $\mathbf{Z}$, $\boldsymbol{\mu}$, $\mathbf{S}$
  — with its abstract dimension and its concrete dimension for this fit.
- `assumption_table(sym)` states the standard-Gaussian prior on the
  latent variables, the trait-specific Gaussian residuals, and the
  lower-triangular identification convention on
  $\boldsymbol{\Lambda}$.
- `parameter_interpretation(sym)` reads each loading as the
  contribution of a one-SD shift on axis $k$ to trait $j$.

```{r symbolize-outputs}
equations(sym)
```

The structural-symbolic surface is the same `symbolize()` call the
package uses for every other model class. Section 6 below reads the
loadings directly from `gllvmTMB`'s native accessors for comparison.

## 6. Reading the latent axes biologically

The raw loading matrix is identified only up to rotation when $d_B >
1$, which is the usual factor-analysis story. A varimax rotation makes
the axes interpretable: each axis tries to load strongly on a small set
of traits and near zero on the rest.

```{r loadings}
getLoadings(fit, rotate = "varimax")
```

Reading down the columns of the rotated loading matrix, the first axis
loads on traits 4 and 5 (the *activity-shelter-use* syndrome in the
simulated truth); the second axis loads on traits 1, 2, and 3 (the
*boldness-exploration-aggression* syndrome). The numerical labels swap
because varimax does not preserve column order — but the *block
structure* matches the truth.

The communality is the proportion of each trait's between-individual
variance that the shared axes capture:

```{r communality}
extract_communality(fit)
```

Traits with communality near 1 are well-explained by the shared axes;
traits with low communality have most of their between-individual
variability in the trait-specific uniqueness $s_j^2$.

The residual between-individual correlation matrix on the latent scale
falls out of `extract_correlations()`:

```{r correlations}
co <- extract_correlations(fit)
co_B <- co[co$tier == "B", c("trait_i", "trait_j", "correlation", "lower", "upper")]
co_B
```

The two within-syndrome correlations (traits 1-2-3 among themselves and
traits 4-5 among themselves) are clearly positive; the
across-syndrome pairs are smaller.

**Takeaway.** Two rotated axes pull apart the two simulated syndromes;
communalities tell you which traits live in the shared structure;
trait-trait correlations show the syndrome blocks directly.

## 7. Identifiability gotchas: rotation and sign of $\boldsymbol{\Lambda}_B$

The loading matrix $\boldsymbol{\Lambda}_B$ is identified only up to
**rotation** and **sign** when $d_B > 1$. Two fits that produce
indistinguishable likelihoods can have entirely different-looking
$\boldsymbol{\Lambda}_B$. This is the classical factor-analysis
identifiability problem. The implied between-unit covariance
$\boldsymbol{\Sigma}_B = \boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^\top
+ \boldsymbol{\Psi}_B$ is invariant under rotation;
$\boldsymbol{\Lambda}_B$ itself is not. The biologist's task is to
read the *block structure* of $\boldsymbol{\Lambda}_B$, not the
specific numbers in any one cell.

A symbolizer surface is explicit about this. Two practical
conventions exist, both surfaced in `assumption_table(sym)`:

- **Lower-triangular $\boldsymbol{\Lambda}_B$** for *confirmatory* use
  — pin the upper triangle to zero so each column has a designated
  leading trait. This is the default identification gllvmTMB uses on
  fitting. It produces a unique solution but its columns are not
  maximally interpretable.
- **Varimax-rotated $\boldsymbol{\Lambda}_B$** for *interpretation* —
  rotate the fitted loadings so each axis loads strongly on a small
  set of traits and near zero on the rest. The rotation is applied
  *after* fitting and preserves $\boldsymbol{\Sigma}_B$. The column
  labels may reorder under rotation; the block structure is what the
  reader interprets.

`symbolize.gllvmTMB()` returns the as-fit (lower-triangular)
$\boldsymbol{\Lambda}_B$ in `sym$loadings`. The assumption rows for
the fit carry a sentence each on the rotation and sign conventions
so a reader does not mistake non-uniqueness for non-replicability.
The varimax-rotated matrix shown earlier in Section 6 is the
reader-facing presentation; the lower-triangular form is what the
likelihood actually identifies.

**Takeaway.** Rotation and sign ambiguity are *stated assumptions*,
not bugs. The implied covariance is invariant; the loadings are not.
Read the block structure, not the cell values.

## 8. Where the symbolized story helps

With five traits and two latents, the per-element index form
$y_{ij} = \mu_j + \sum_{k} \lambda_{jk} z_{ik} + \epsilon_{ij}$ requires
the reader to mentally instantiate fifty equations (forty individuals
times five traits, with two latents apiece). The matrix block

$$
\mathbf{Y} = \mathbf{1}\boldsymbol{\mu}^{\!\top} + \mathbf{Z}\,\boldsymbol{\Lambda}^{\!\top} + \mathbf{E}
$$

says exactly the same thing in one line. A biologist can *see* that
$\mathbf{Z}$ stores one row of latent scores per individual,
$\boldsymbol{\Lambda}$ stores one row of loadings per trait, and the
product is the deterministic part of the trait matrix.

This is the educator-first surface symbolizer is built around: the
structural object always carries both notations, and the
`symbol_table()` always carries both abstract and concrete dimensions.
For latent-variable models, where the matrix form is genuinely the more
readable one, that bridge is where the package earns its keep.

**Takeaway.** Index form scales like $n \times T$; matrix form scales
like one block. For GLLVMs the matrix form is the reader-friendly one.

## 9. What's available now, what's next

The accessors in Section 6 — `getLoadings()`, `extract_communality()`,
`extract_correlations()` — are gllvmTMB-side. The symbolizer surface
wraps the Gaussian latent-variable case as a **First slice**: every
piece walked above is shipped today, and the matrix-form story is
fully populated.

**Shipped today:**

- `symbolize.gllvmTMB()` for Gaussian and binomial latent-variable
  families: mu, $\boldsymbol{\Lambda}_B$, $\boldsymbol{\Sigma}_B$,
  $\boldsymbol{\Psi}_B$, and the within-unit residual SD
  $\sigma_\varepsilon$ (Gaussian only). First slice since
  v0.4 – v0.5.
- `compare_symbolic(fit_d1, fit_d2)` — diffs the structural
  specifications of $d_B = 1$ versus $d_B = 2$ side by side
  (which lines change, which assumption rows change), and with
  `metrics = TRUE` also returns AIC / BIC / log-likelihood / df
  with the delta. See `vignette("symbolizer-compare")` for the
  worked example.

**Still planned:**

1. **Other non-Gaussian gllvmTMB families** (count, Tweedie,
   ordinal) — each adds a link to the linear predictor and reshapes
   the loadings interpretation.
2. **Automated syndromes reading** —
   `parameter_interpretation(sym, scale = "biological")` produces
   one row per (trait, axis) pair today; a later slice will add the
   natural-scale reading for the remaining families.
3. **Uncertainty on $\boldsymbol{\Lambda}$.** `bootstrap_Sigma()`
   from gllvmTMB returns parametric-bootstrap draws of the loading
   matrix; a future slice will surface these through `symbolize()`
   so the `symbol_table()` rows for $\boldsymbol{\Lambda}$ and
   $\boldsymbol{\Sigma}_B$ carry confidence regions, not just
   point estimates.
4. **Within-unit decomposition** ($\boldsymbol{\Sigma}_W$,
   $\boldsymbol{\Lambda}_W$) for repeated-measures behaviour /
   trait data — `gllvmTMB_multi`.

See `vignette("symbolizer-roadmap")` for the full capability
matrix and what's planned beyond gllvmTMB.
