---
title: "Get started with symbolizer"
description: "A five-minute first contact: fit a model, then read its equation, assumptions, and plain-English interpretation."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with symbolizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

`symbolizer` turns a fitted model into a *structured symbolic model* — an object
your renderers read to produce publication-ready equations, assumption tables,
and per-coefficient interpretations. This is the five-minute tour: one model, one
`symbolize()` call, and the three things you get back. It runs on **base R alone**
— nothing to install.

## Why structured symbolic models

[`equatiomatic`](https://datalorax.github.io/equatiomatic/) renders a fitted
model as a LaTeX string — enough when the only goal is a publishable equation.
`symbolizer` returns a richer object: from the same `symbolized_model` you can
pull the equation (in both index and matrix notation), the symbol dictionary, the
assumption table (each assumption labelled *stated* / *implied* / *not checked*),
the formula-to-math bridge, and the per-coefficient interpretation on link,
natural, variance, and biological scales.

**Takeaway.** The product is the `symbolized_model` object; everything else is a
renderer of that object.

## A short glossary

If formula-grammar terminology isn't second nature, here's the quickest map of
what each word means in a biology context.

```{r glossary, echo = FALSE}
glossary <- data.frame(
  word = c(
    "response", "predictor", "factor", "submodel", "linear predictor",
    "design matrix", "coefficient", "link function"
  ),
  meaning = c(
    "the outcome you measured (e.g., body mass, abundance).",
    "a variable you think influences the response.",
    "a categorical predictor with named levels (e.g., sex with \"female\" and \"male\").",
    "one piece of the formula. A location-scale model has two submodels: one for the mean (mu), one for the residual SD (sigma).",
    "the sum that determines a parameter for an observation: intercept + slope x predictor + ... Sometimes a link function (e.g., log) is applied first.",
    "the table the computer multiplies coefficients by to get fitted values. Each row is one observation; each column is one term.",
    "a single number the model estimates: an intercept, a slope, or a factor contrast.",
    "the transformation between a parameter's natural scale (e.g., a count) and the scale the linear predictor works on (e.g., log-count)."
  ),
  stringsAsFactors = FALSE
)
knitr::kable(glossary, col.names = c("word", "meaning"))
```

(For how a factor becomes 0/1 dummy columns, and contrasts in depth, see
`vignette("symbolizer-factors")`.)

## Your first fit

We use a model every R user already has: a base-R Poisson GLM of a count against
a predictor. No non-CRAN packages required.

```{r fit}
library(symbolizer)

set.seed(1)
n <- 80
temperature <- runif(n, 10, 25)
dat <- data.frame(
  count       = rpois(n, exp(0.3 + 0.08 * temperature)),
  temperature = temperature
)

fit <- glm(count ~ temperature, family = poisson, data = dat)
```

`symbolize()` builds the structured object that every other function reads. (In
a hurry? `explain(fit)` prints the equation, assumptions, and a coefficient
reading in one call — the recommended first look for newcomers; here we build the
object so we can explore each piece.) Pass user-facing symbols and units so
the equations carry biological meaning rather than R variable names:

```{r symbolize}
sym <- symbolize(
  fit,
  symbols = c(count = "N_i", temperature = "T_i"),
  units   = c(temperature = "C"),
  context = "abundance ~ temperature, Poisson GLM"
)
```

That's it — `sym` is the structured object. The rest of this page shows the three
things it gives you.

## The three things symbolizer gives you

### 1. The equation, in real math

`equations()` returns one row per renderable block; `as_latex()` produces the
string you splice into a manuscript. The Poisson log link shows up explicitly:

```{r equation, results = 'asis'}
cat("$$", as_latex(sym, notation = "both"), "$$", sep = "\n")
```

Bold lowercase is a vector ($\boldsymbol{\mu}$, $\boldsymbol{\beta}$), bold
uppercase a matrix ($\mathbf{X}$); the index form drops to per-observation
symbols ($\mu_i$, $\beta_0$, $\beta_1$, $T_i$).

### 2. Three views of the fit

`as_html_three_views()` is a self-contained widget with three tabs over the same
fit, in the order they appear in the widget: the per-observation **index** form,
the matrix-form **equation**, and **equations with data** — the response column,
the design-matrix rows, the coefficient vector, and the fitted mean
$\hat{\boldsymbol{\mu}}$.

```{r three-views, results = 'asis'}
as_html_three_views(sym, head = 5, tail = 2)
```

### 3. What each coefficient means

`parameter_interpretation()` reads every estimate on the scales that make sense
for the model. For a Poisson GLM the natural-scale reading is the **rate ratio**:
$\exp(\hat\beta)$ multiplies the expected count per unit of the predictor.

```{r interpretation}
parameter_interpretation(sym)
```

**Takeaway.** One `symbolize()` call → the equation, the three-views widget, and a
per-coefficient reading you can paste straight into a Methods section.

## Where to go next

- **Build up from `lm` to location-scale**, one rung at a time on a shared
  dataset — `vignette("symbolizer-ladder")`.
- **A tour of non-Gaussian families** (Gamma, beta, binomial, negative binomial,
  …) — `vignette("symbolizer-families")`.
- **Factors, dummies, and interactions** in depth —
  `vignette("symbolizer-factors")`.
- **Random effects, ICC, and repeatability** —
  `vignette("symbolizer-variance-components")`.
- **The full function reference** — `symbol_table()`, `assumption_table()`,
  `formula_bridge()`, `notation_bridge()`, `model_card()`, `expand()`,
  `as_dag()`, and the rest — in the package's *Reference* index.
- **What's supported and what's planned** across the eleven model families
  `symbolize()` reads (`drmTMB`, `gllvmTMB`, `glmmTMB`, `brms`, `lme4`,
  `MCMCglmm`, `sdmTMB`, `stats::lm`/`glm`, `metafor`, `mgcv`, `phylolm`) —
  `vignette("symbolizer-roadmap")`, or call `symbolizer_capabilities()`.
