---
title: "What a symbolized GLLVM should explain"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What a symbolized GLLVM should explain}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_gllvmTMB <- requireNamespace("gllvmTMB", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_gllvmTMB
)
```

## 1. Why a design target

A generalised linear latent variable model (GLLVM) is more than its
likelihood. It is a story about *shared between-trait variation*, *trait-
specific noise*, *latent axes*, and *identifiability constraints*. A
symbolizer surface for any GLLVM has to explain all four. This page is
the contract. It says what a complete answer looks like when
`symbolize.gllvmTMB()` returns its `symbolized_model`, and it pins each
item of the contract to the field of the object where you would find
that answer.

The companion vignette
[`symbolizer-gllvm`](symbolizer-gllvm.html) is the worked example: one
fit, one set of accessors, one biological reading. This vignette is the
target the worked example is aimed at. Every promise made here either
lands as a populated row of `symbolized_model` in the v0.4 First slice,
or is flagged "v0.5" or "planned" when it is on the roadmap but not yet
in.

**Takeaway.** This page is the contract — what `symbolize.gllvmTMB()`
must surface, and where in the returned `symbolized_model` a reader
should look for each piece.

## 2. The structural equation, both notations

The GLLVM has the same shape in two notations. **Index form** (one
equation per observation, per Eq. 11 of the Nakagawa et al. trait-
covariance framework):

$$
y_{ijt} = \mu_t + \sum_{k=1}^{d_B} \lambda_{tk}\, z_{ik} + s_{it} + e_{ijt}
$$

Index $i$ ranges over the between-unit replicate (individual, site, or
species), $j$ over the row, $t$ over the trait, $k$ over the latent
axis. The four pieces are: $\mu_t$ the trait grand mean; $\sum_k
\lambda_{tk} z_{ik}$ the shared low-rank between-unit signal;
$s_{it}$ the trait-specific between-unit deviation; $e_{ijt}$ the
within-unit (across-row) residual.

**Matrix form** (one block for the whole dataset):

$$
\mathbf{Y} = \mathbf{1}\boldsymbol{\mu}^{\!\top}
            + \mathbf{Z}\,\boldsymbol{\Lambda}^{\!\top}
            + \mathbf{S}
            + \mathbf{E}
$$

$\mathbf{Y}$ is the $n \times T$ matrix of trait values,
$\boldsymbol{\mu}$ the length-$T$ vector of trait means, $\mathbf{Z}$
the $n \times d_B$ matrix of latent scores, $\boldsymbol{\Lambda}$ the
$T \times d_B$ loading matrix, $\mathbf{S}$ the $n \times T$ matrix of
trait-specific between-unit deviations (rows distributed as
$\mathcal{N}(\mathbf{0}, \boldsymbol{\Psi}_B)$ with
$\boldsymbol{\Psi}_B = \mathrm{diag}(\psi_{B,1}, \ldots, \psi_{B,T})$),
and $\mathbf{E}$ the within-unit residual.

A symbolizer surface must carry *both* forms. Index form is the one a
reader instantiates row by row; matrix form is the one a reader uses to
see the linear algebra of the fit.

**Takeaway.** Both notations live in `equations(sym)` and on the
`equation` / `equation_matrix` columns of `components(sym)`.

## 3. The decomposition that matters most: $\boldsymbol{\Sigma}_B = \boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^\top + \boldsymbol{\Psi}_B$

The between-unit trait covariance is the deliverable of a GLLVM in
ecology and evolution. It decomposes additively: the shared low-rank
piece $\boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^\top$ holds the
covariation that travels through the latent axes; the diagonal
$\boldsymbol{\Psi}_B$ holds the trait-specific between-unit variance
that is not captured by the axes.

```{r sigma-bridge, echo = FALSE}
sigma_bridge <- data.frame(
  Symbol = c(
    "$\\boldsymbol{\\Sigma}_B$",
    "$\\boldsymbol{\\Lambda}_B$",
    "$\\boldsymbol{\\Lambda}_B \\boldsymbol{\\Lambda}_B^\\top$",
    "$\\boldsymbol{\\Psi}_B$",
    "$\\mathbf{C}_B = \\mathbf{D}_B^{-1/2} \\boldsymbol{\\Sigma}_B \\mathbf{D}_B^{-1/2}$"
  ),
  Meaning = c(
    "Implied between-unit trait covariance",
    "Shared low-rank loadings (trait by axis)",
    "Shared communality (rank-$d_B$ covariance from the axes)",
    "Trait-specific unique between-unit variance",
    "Between-unit trait correlation"
  ),
  `gllvmTMB accessor` = c(
    "`extract_Omega(fit)`",
    "`getLoadings(fit)`",
    "`extract_communality(fit)`",
    "`getResidualCov(fit)$Psi`",
    "`extract_correlations(fit)`"
  ),
  `Biological reading` = c(
    "Total between-unit trait covariance the model implies.",
    "How strongly trait $t$ loads on latent axis $k$.",
    "Variance shared across traits through the latent axes.",
    "Variance left after the shared axes; trait-specific noise.",
    "Positive entries are behavioural syndromes; negative entries are antagonistic axes.",
    NULL
  )[1:5],
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(sigma_bridge)
```

The five rows are not five independent quantities — they are five
*views* of the same between-unit trait covariance, each one tuned to a
different audience. A reader who wants the headline number wants
$\mathbf{C}_B$; a reader who wants the mechanism wants
$\boldsymbol{\Lambda}_B$; a reader who wants to know how much was *not*
explained wants $\boldsymbol{\Psi}_B$.

**Takeaway.** A symbolized GLLVM must carry all five rows. The
$\boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^\top + \boldsymbol{\Psi}_B$
identity is the bridge; the other three rows are derived views.

## 4. What `symbolize()` must populate

The symbol table below lists the `symbolized_model` fields a fitted
GLLVM should populate and the accessor that reads each field.

```{r symbolized-model-fields, echo = FALSE}
fields <- data.frame(
  Field = c(
    "`model`",
    "`submodels`",
    "`terms`",
    "`loadings`",
    "`variance_components`",
    "`components`",
    "`assumptions`",
    "`symbol_dictionary`",
    "`formula_bridge`",
    "`interpretation`"
  ),
  `What it carries` = c(
    "Fitted-object class, family, response, and model dimensions ($n$, $T$, $d_B$).",
    "One row per coefficient family: $\\boldsymbol{\\mu}$ (trait intercepts) and $\\boldsymbol{\\Lambda}_B$ (between-unit loadings).",
    "Per-trait intercept rows from the `0 + trait` design, each tagged `role = \"trait_intercept\"`.",
    "One row per ($\\lambda_{B,tk}$) loading entry with estimate and SE.",
    "Diagonal of $\\boldsymbol{\\Psi}_B$ (per-trait unique SD) and the shared row residual $\\sigma_\\epsilon$.",
    "Distribution, linear predictor, latent score distribution $\\mathbf{z}_{B,i} \\sim \\mathcal{N}(\\mathbf{0}, \\mathbf{I}_{d_B})$, and the implied covariance row $\\boldsymbol{\\Sigma}_B = \\boldsymbol{\\Lambda}_B \\boldsymbol{\\Lambda}_B^\\top + \\boldsymbol{\\Psi}_B$.",
    "Conditional independence given $\\mathbf{z}_{B,i}$, standard-Gaussian latent prior, lower-triangular rotation convention, missing trait cells, biological reading of $\\boldsymbol{\\Sigma}_B$.",
    "$\\mathbf{Y}$, $\\boldsymbol{\\mu}$, $\\boldsymbol{\\Lambda}_B$, $\\mathbf{Z}_B$, $\\boldsymbol{\\Psi}_B$, $\\sigma_\\epsilon$, $\\boldsymbol{\\Sigma}_B$ — each with abstract and concrete dimensions.",
    "R syntax (`0 + trait`, `latent(0 + trait | unit, d = K)`) mapped to the corresponding math in both notations.",
    "Per-trait intercept and per-loading prose, on link / natural / variance / biological scales."
  ),
  Accessor = c(
    "`sym$model`",
    "`sym$submodels`",
    "`sym$terms`",
    "`sym$loadings`",
    "`sym$variance_components`",
    "`components(sym)`",
    "`assumption_table(sym)`",
    "`symbol_table(sym)`",
    "`formula_bridge(sym)`",
    "`parameter_interpretation(sym)`"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(fields)
```

Two fields deserve a closer look. **`components`** is where the four
structural lines of section 2 live, one row each: the distribution, the
linear predictor for $\boldsymbol{\mu}$, the latent score distribution,
and the implied-covariance row. **`assumptions`** is where the rotation
and sign ambiguity (section 6 below) and the standard-Gaussian latent
prior are stated explicitly so a reader knows what the loading matrix
does and does not pin down.

**Takeaway.** Ten populated fields. The contract is that a `symbolized_
model` from a Gaussian GLLVM First slice carries all ten of them; a
renderer never has to read the fit again.

## 5. The hard truth of capabilities

Not every GLLVM is in scope yet. The capability registry
(`symbolizer_capabilities()`) gates `symbolize()` against the family,
class, and component tuple of the incoming fit. The gllvmTMB rows look
like this:

```{r gllvm-capabilities, echo = FALSE}
caps <- data.frame(
  class = "gllvmTMB",
  family = c(
    "gaussian", "gaussian", "gaussian", "gaussian", "gaussian",
    "*", "*", "*", "*", "*",
    "binomial", "poisson", "nbinom2", "Gamma", "lognormal",
    "Beta", "betabinomial", "tweedie", "student", "ordinal_probit"
  ),
  component = c(
    "mu", "Lambda_B", "Sigma_B", "Psi_B", "sigma_eps",
    "Lambda_W", "Psi_W", "phylo", "spatial", "Sigma",
    "*", "*", "*", "*", "*",
    "*", "*", "*", "*", "*"
  ),
  status = c(
    rep("First slice", 5),
    rep("Planned or reserved", 15)
  ),
  stringsAsFactors = FALSE
)
knitr::kable(caps)
```

The five gllvmTMB Gaussian rows — $\boldsymbol{\mu}$,
$\boldsymbol{\Lambda}_B$, $\boldsymbol{\Sigma}_B$, $\boldsymbol{\Psi}_B$,
$\sigma_\epsilon$ — are First slice in v0.4. Everything else is planned.
In particular, the within-unit decomposition ($\boldsymbol{\Lambda}_W$,
$\boldsymbol{\Psi}_W$) is planned for v0.5. Non-Gaussian families
(Bernoulli, Poisson, negative binomial, beta, gamma, tweedie, ordinal)
are planned for later v0.5 / v0.6 work. Phylogenetic and spatial
covstructs are planned beyond that.

**`symbolize()` will refuse to read a fit whose family or structure is
not in the registry as Stable or First slice.** The error message lists
the currently-readable tuples so the user sees what they *can* do
today. This is intentional: a symbolic surface that silently dropped
unsupported terms would be more dangerous than no surface at all.

**Takeaway.** Gaussian, between-unit, reduced-rank-plus-diagonal is in.
Everything else is planned and is rejected by the registry until its
row turns First slice.

## 6. Identifiability gotchas a biologist must know

The loading matrix $\boldsymbol{\Lambda}_B$ is identified only up to
**rotation** and **sign** when $d_B > 1$. Two fits that produce
indistinguishable likelihoods can have entirely different-looking
$\boldsymbol{\Lambda}_B$. This is the classical factor-analysis
identifiability problem. The implied between-unit covariance
$\boldsymbol{\Sigma}_B = \boldsymbol{\Lambda}_B \boldsymbol{\Lambda}_B^\top
+ \boldsymbol{\Psi}_B$ is invariant under rotation; $\boldsymbol{\Lambda}_B$
itself is not.

A symbolizer surface has to be explicit about this. There are two
practical conventions, both surfaced in `assumption_table(sym)`:

- **Lower-triangular** $\boldsymbol{\Lambda}_B$ for *confirmatory*
  use — pin the upper triangle to zero so each column has a
  designated leading trait. This is the default identification
  gllvmTMB uses on fitting. It produces a unique solution but its
  columns are not maximally interpretable.
- **Varimax-rotated** $\boldsymbol{\Lambda}_B$ for *interpretation* —
  rotate the fitted loadings so each axis loads strongly on a small
  set of traits and near zero on the rest. The rotation is applied
  *after* fitting and preserves $\boldsymbol{\Sigma}_B$. The column
  *labels* may reorder under rotation; the *block structure* is what
  the reader interprets.

`symbolize.gllvmTMB()` returns the as-fit (lower-triangular)
$\boldsymbol{\Lambda}_B$ in `sym$loadings`. The assumption rows for the
fit carry a sentence each on the rotation and sign conventions so a
reader does not mistake non-uniqueness for non-replicability. The
worked-example vignette
([`symbolizer-gllvm`](symbolizer-gllvm.html)) shows the varimax-rotated
matrix on the reader-facing side.

**Takeaway.** Rotation and sign ambiguity are *stated assumptions*, not
bugs. The implied covariance is invariant; the loadings are not.

## 7. What comes next (v0.5+)

Three extensions are on the roadmap but **not in** the v0.4 First
slice:

- **Within-unit decomposition.** v0.5 will surface
  $\boldsymbol{\Lambda}_W$ and $\boldsymbol{\Psi}_W$ — the
  reduced-rank-plus-diagonal decomposition of the *within-unit*
  trait covariance — so a fit with two `latent(...)` terms (one for
  between-unit, one for within-unit) carries the same Σ-bridge table
  twice, once per tier.
- **Phylogenetic and spatial structured latents.** Planned for the
  release after v0.5: `phylo_rr`, `phylo_diag`, `spde`, and `spatial`
  covstructs. The structural surface gains a phylogenetic /
  spatial covariance row alongside $\boldsymbol{\Sigma}_B$.
- **Non-Gaussian families.** Bernoulli (joint species distribution
  models from presence-absence), Poisson (count communities),
  negative binomial, beta, gamma, tweedie, ordinal probit are
  planned. The Σ-bridge generalises: the response covariance is
  read on the link scale via $\boldsymbol{\Sigma}_B = \boldsymbol{\Lambda}_B
  \boldsymbol{\Lambda}_B^\top + \boldsymbol{\Psi}_B$, but
  $\boldsymbol{\Psi}_B$'s interpretation changes — for example, in
  the Bernoulli case the diagonal residual variance is fixed by the
  link, so $\boldsymbol{\Psi}_B$ is the *extra* over-dispersion above
  what the link supplies.

**Takeaway.** v0.4 covers Gaussian, between-unit, reduced-rank-plus-
diagonal. The Σ-bridge generalises one tier at a time as the registry
turns each planned row into a First slice.
