---
title: "Three faces of meta-analysis: metafor, glmmTMB, drmTMB"
description: >
  The same two-tier meta-analytic model, fit three ways. Symbolizer
  reads each one and reveals the shared structure underneath the
  package-specific syntax.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Three faces of meta-analysis: metafor, glmmTMB, drmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_metafor  <- requireNamespace("metafor",  quietly = TRUE)
has_glmmTMB  <- requireNamespace("glmmTMB",  quietly = TRUE)
has_MASS     <- requireNamespace("MASS",     quietly = TRUE)
has_metadat  <- requireNamespace("metadat",  quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_metafor && has_glmmTMB && has_MASS && has_metadat
)
```

## The shared model

A meta-analysis is, structurally, a two-tier Normal model:

$$
\begin{aligned}
y_i \mid \theta_i &\sim \mathcal{N}(\theta_i,\, v_i), \quad v_i \text{ known} \\
\theta_i &= \beta_0 + \sum_k \beta_k\, x_{ki} + u_i \\
u_i &\sim \mathcal{N}(0,\, \tau^2)
\end{aligned}
$$

The observed effect size $y_i$ from study $i$ is a noisy estimate of a
true effect $\theta_i$, with **known** sampling variance $v_i$ (an
input, not a parameter). The true effects vary around a meta-analytic
mean with between-study heterogeneity variance $\tau^2$. Moderators
$x_{ki}$ explain part of the location; a separate scale model can
explain part of $\tau^2$ as well (the **location-scale**
extension; see Viechtbauer & López-López, 2022; Nakagawa et al.,
2025).

What changes from package to package isn't the model — it's the
syntax used to construct it. `symbolizer` reads all three.

## Setup — a tiny simulated meta-analytic dataset

```{r sim}
library(symbolizer)
library(metafor)
library(glmmTMB)

set.seed(1)
k_studies <- 10
k_per <- 3
study <- rep(seq_len(k_studies), times = k_per)
k <- length(study)
id <- seq_len(k)
# Sampling variances (known per effect size)
vi <- rbeta(k, 2, 20)
# Block-diagonal V with within-study correlation rho = 0.5
V <- matrix(0, k, k)
for (s in unique(study)) {
  idx <- which(study == s)
  for (i in idx) for (j in idx) {
    V[i, j] <- if (i == j) vi[i] else 0.5 * sqrt(vi[i] * vi[j])
  }
}
rownames(V) <- colnames(V) <- as.character(id)
# Simulate y_i around a true between-study mean
true_y <- 0 + rnorm(k_studies, 0, 0.5)[study]
y <- true_y + as.numeric(MASS::mvrnorm(1, rep(0, k), V))
dat <- data.frame(y = y, vi = vi, study = factor(study),
                  obs = factor(id), g = factor(1))
head(dat, 4)
```

## Face 1 — metafor (the canonical interface)

```{r metafor}
fit_meta <- rma.mv(yi = y, V = V,
                   random = list(~ 1 | study, ~ 1 | obs),
                   data = dat)
sym_meta <- symbolize(fit_meta)
```

The symbolic surface:

```{r metafor-out}
cat(as_latex(sym_meta), "\n")
```

```{r metafor-vc}
sym_meta$variance_components
```

The model the package teaches: `meta_normal` family, two random-effect
tiers (study + observation), sampling variances `v_i` known.

## Face 2 — glmmTMB via `propto()` (the GLMM bridge)

The same model can be fit through `glmmTMB` by attaching the known
`V` matrix via `propto()`. Structurally identical; syntactically
different.

```{r glmm}
fit_glmm <- glmmTMB(
  y ~ 1 + (1 | study) + propto(0 + obs | g, V),
  data = dat, REML = TRUE
)
sym_glmm <- symbolize(fit_glmm)
```

`symbolizer` detects the `propto()` block and adds an info-level
warning row:

```{r glmm-warn}
warning_table(sym_glmm)
```

This is the bridge: `propto(0 + obs | g, V)` **fixes** the residual
covariance to the known `V`, which is exactly what `rma.mv(..., V = V)`
does in `metafor`. The `(1 | study)` term carries the between-study
heterogeneity $\tau^2$.

When this pattern is present:

| Conceptual quantity | metafor | glmmTMB |
| --- | --- | --- |
| Known sampling-variance matrix | `V` argument | `propto(0 + obs | g, V)` block |
| Between-study heterogeneity $\tau^2$ | `sigma2[["study"]]` | `VarCorr(fit)$cond$study` |
| Moderator coefficients $\beta_k$ | `fit$beta` | `fixef(fit)$cond` |

```{r glmm-out, results = "markup"}
cat(as_latex(sym_glmm), "\n")
```

## Face 3 — drmTMB location-scale (the distributional surface)

`drmTMB` has location-scale Gaussian as its core form. With known
sampling variances entered as $\sigma_i^2 = v_i$ (via an offset on
the log-scale formula), `drmTMB` fits the meta-analytic model as a
distributional Gaussian. Symbolic output uses $\log(\sigma_i)$ on the
**SD scale** (the drmTMB convention), with $\gamma_k$ coefficients.

(Sketch only — a full drmTMB construction with known sigma offsets
is shown in `vignette("symbolizer-drmtmb")` and the drmTMB docs.)

## The variance / SD parameterization gap

Each package writes the scale model differently:

| Package | Quantity modelled | Link | Coefficient |
| --- | --- | --- | --- |
| `metafor::rma(scale = ~ z)` | $\tau^2$ (variance) | $\log(\tau^2_i) = \alpha_0 + \alpha_1 z_i$ | $\alpha$ |
| `brms::bf(sigma ~ z)` | $\sigma$ (SD) | $\log(\sigma_i) = \gamma_0 + \gamma_1 z_i$ | $\gamma$ |
| `glmmTMB(dispformula = ~ z)` | $\sigma$ (SD) | $\log(\sigma_i) = \gamma_0 + \gamma_1 z_i$ | $\gamma$ |
| `drmTMB(sigma ~ z)` | $\sigma$ (SD) | $\log(\sigma_i) = \gamma_0 + \gamma_1 z_i$ | $\gamma$ |

The two parameterizations are mathematically equivalent. Because
$\log(\tau^2) = 2 \log(\tau) + \mathrm{const}$, the slope on the
variance scale is twice the slope on the SD scale:

$$
\alpha_k \;\approx\; 2 \cdot \gamma_k
$$

(For the intercept the relationship is offset by $\log 2$ once.) So
a `metafor` location-scale slope of $\alpha = 0.4$ is the same
biological signal as a `brms` / `glmmTMB` / `drmTMB` location-scale
slope of $\gamma \approx 0.2$ — variance doubles ≈ SD increases by
$\sqrt{2}$.

`symbolizer` renders each faithfully in its own parameterization:
look for the coefficient symbol ($\alpha$ vs $\gamma$) and the
$\tau^2$-vs-$\sigma$ distinction in the LaTeX block to know which
convention is in play.

## When to use which

- **`metafor`** is the canonical interface for traditional
  meta-analysis. Rich diagnostics, well-documented effect-size
  computation (`escalc()`), and a long literature on small-sample
  corrections.
- **`glmmTMB` with `propto()`** brings meta-analysis into the GLMM
  ecosystem — letting you reuse the same modelling pipeline
  (model checking, prediction, marginal effects) you use for
  GLMMs in general.
- **`drmTMB`** lets you fit location-scale meta-analysis with
  moderators on heterogeneity directly, with the same syntax as any
  other location-scale model. This is the surface Nakagawa et al.
  (2025) explore in depth for ecology and evolution.

`symbolizer` lets all three speak the same teachable language.

## References

- Viechtbauer, W., & López-López, J. A. (2022). Location-scale
  models for meta-analysis. *Research Synthesis Methods*, **13**(6),
  697–715.
- Nakagawa, S., Mizuno, A., Morrison, K., Ricolfi, L., Williams, C.,
  Drobniak, S. M., Lagisz, M., & Yang, Y. (2025). Location-scale
  meta-analysis and meta-regression as a tool to capture large-scale
  changes in biological and methodological heterogeneity: A
  spotlight on heteroscedasticity. *Global Change Biology*, **31**,
  e70204.
- Williams, C. R. (2023). *Fitting meta-analysis with the glmmTMB R
  package: a worked tutorial.*
  <https://coraliewilliams.github.io/equalto_sim_study/webpage.html>
