---
title: "Three faces of meta-analysis: metafor, glmmTMB, drmTMB"
description: >
  The same two-tier meta-analytic model, fit three ways. Symbolizer
  reads each one and reveals the shared structure underneath the
  package-specific syntax.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Three faces of meta-analysis: metafor, glmmTMB, drmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_metafor  <- requireNamespace("metafor",  quietly = TRUE)
has_glmmTMB  <- requireNamespace("glmmTMB",  quietly = TRUE)
has_MASS     <- requireNamespace("MASS",     quietly = TRUE)
has_metadat  <- requireNamespace("metadat",  quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_metafor && has_glmmTMB && has_MASS && has_metadat
)
```

## The shared model

A meta-analysis is, structurally, a two-tier Normal model:

$$
\begin{aligned}
y_i \mid \theta_i &\sim \mathcal{N}(\theta_i,\, v_i), \quad v_i \text{ known} \\
\theta_i &= \beta_0 + \sum_k \beta_k\, x_{ki} + u_i \\
u_i &\sim \mathcal{N}(0,\, \tau^2)
\end{aligned}
$$

The observed effect size $y_i$ from study $i$ is a noisy estimate of a
true effect $\theta_i$, with **known** sampling variance $v_i$ (an
input, not a parameter). The true effects vary around a meta-analytic
mean with between-study heterogeneity variance $\tau^2$. Moderators
$x_{ki}$ explain part of the location; a separate scale model can
explain part of $\tau^2$ as well (the **location-scale**
extension; see Viechtbauer & López-López, 2022; Nakagawa et al.,
2025).

What changes from package to package isn't the model — it's the
syntax used to construct it. `symbolizer` reads all three.

## Setup — a tiny simulated meta-analytic dataset

```{r sim}
library(symbolizer)
library(metafor)
library(glmmTMB)

set.seed(1)
k_studies <- 10
k_per <- 3
study <- rep(seq_len(k_studies), times = k_per)
k <- length(study)
id <- seq_len(k)
# Sampling variances (known per effect size)
vi <- rbeta(k, 2, 20)
# Block-diagonal V with within-study correlation rho = 0.5
V <- matrix(0, k, k)
for (s in unique(study)) {
  idx <- which(study == s)
  for (i in idx) for (j in idx) {
    V[i, j] <- if (i == j) vi[i] else 0.5 * sqrt(vi[i] * vi[j])
  }
}
rownames(V) <- colnames(V) <- as.character(id)
# Simulate y_i around a true between-study mean
true_y <- 0 + rnorm(k_studies, 0, 0.5)[study]
y <- true_y + as.numeric(MASS::mvrnorm(1, rep(0, k), V))
dat <- data.frame(y = y, vi = vi, study = factor(study),
                  obs = factor(id), g = factor(1))
head(dat, 4)
```

## Face 1 — metafor (the canonical interface)

```{r metafor}
fit_meta <- rma.mv(yi = y, V = V,
                   random = list(~ 1 | study, ~ 1 | obs),
                   data = dat)
sym_meta <- symbolize(fit_meta)
```

The symbolic surface:

```{r metafor-out}
equations(sym_meta, notation = "index")
```

```{r metafor-vc}
sym_meta$variance_components
```

The model the package teaches: `meta_normal` family, two random-effect
tiers (study + observation), sampling variances `v_i` known.

## Face 2 — glmmTMB via `propto()` (phylogenetic / structured-covariance, NOT meta-analysis)

> **v0.20 correction.** v0.16 of this article claimed `propto(0 + obs | g, V)`
> was "structurally identical" to `metafor::rma.mv(..., V = V)`.
> **That was wrong.** Empirical verification (k = 30, seed 1, see Fisher's
> v0.20.0 audit in `NEWS.md`): metafor estimates β̂₀ = 0.357438 and
> τ̂² = 0.068911; the glmmTMB-propto fit on the same data estimates
> β̂₀ = 0.357024 and τ̂² = 0.071700, **with an additional free scalar
> σ̂²_propto = 0.942742** that propto inserts between Σ and V. The
> models differ in parameter count, standard errors, and log-likelihood.
> See §"What `propto` actually does" below for the corrected story.

### What `propto` actually does

`glmmTMB::propto(X, V)` parameterises

$$
\boldsymbol{\Sigma}_{\text{propto}} \;=\; \sigma_{\text{propto}}^{2}\, V
$$

with **σ²_propto estimated** as a free scalar. Combined with glmmTMB's
default residual term (unless `dispformula = ~ 0` is passed), the full
conditional covariance is

$$
\operatorname{Cov}(\mathbf{y}\mid \mathbf{u})
\;=\; \sigma_{\text{propto}}^{2}\,V \;+\; \sigma_{\text{res}}^{2}\,\mathbf{I},
$$

i.e. **two** free scalars sitting on top of `V`, not zero. This is the
**phylogenetic / pedigree / known-correlation** pattern: you know the
*shape* of the covariance (the matrix `V`, or a phylogenetic correlation
matrix `C`), and the model estimates how big it is.

The corresponding metafor construction is

```r
metafor::rma.mv(yi, V = 0 * diag(k),
                random = list(~ 1 | study, ~ 1 | obs),
                R = list(obs = V))
```

— i.e. `propto` mirrors metafor's `R = list(...)` argument, **not** its
`V = V` argument. Fisher's audit fitted both on the same AR(1)-correlation
data and they agree to five decimals.

### Why the package detects `propto` anyway

The `glmmTMB` extractor still detects covariance code 11 (the propto
block) and surfaces an info-level row in `warning_table()`. The info-row
text in v0.20+ names propto as a **structured-covariance / phylogenetic**
pattern, not meta-analysis:

```{r glmm}
fit_glmm <- glmmTMB(
  y ~ 1 + (1 | study) + propto(0 + obs | g, V),
  data = dat, REML = TRUE
)
sym_glmm <- symbolize(fit_glmm)
```

```{r glmm-warn}
warning_table(sym_glmm)
```

```{r glmm-out}
equations(sym_glmm, notation = "index")
```

### What you'd need for an actual meta-analytic glmmTMB fit

The structural twin of `rma.mv(..., V = V)` would be

```r
glmmTMB(y ~ 1 + (1 | study) + equalto(0 + obs | g, V),
        data = dat, dispformula = ~ 0)
```

where `equalto(X, V)` parameterises Σ = V exactly (no scalar
multiplier), and `dispformula = ~ 0` suppresses the residual term.
**`equalto` is reserved but not yet implemented in `glmmTMB` ≤ 1.1.11**
(`.valid_covstruct` shows codes 0–13; no equalto entry). Until glmmTMB
ships it, the GLMM-side meta-analytic bridge is incomplete; the honest
two routes today are metafor (Face 1) and drmTMB (Face 3 below).

## Face 3 — drmTMB location-scale (the distributional surface)

`drmTMB` has location-scale Gaussian as its core form. With known
sampling variances entered as $\sigma_i^2 = v_i$ (via an offset on the
log-scale formula), `drmTMB` fits a meta-analytic-flavoured model as a
distributional Gaussian. Symbolic output uses $\log(\sigma_i)$ on the
**SD scale** (the drmTMB convention), with $\gamma_k$ coefficients.

> **Caveat (v0.20).** Writing `sigma ~ 1 + offset(0.5 * log(vi))` does
> not pin $\sigma_i^{2} = v_i$ unless the intercept $\gamma_0$ is *also*
> constrained to 0 (e.g. via `map`). Without that constraint, drmTMB
> estimates $\sigma_i = \exp(\hat{\gamma}_0) \sqrt{v_i}$ — **proportional**
> to the sampling SE, not equal to it. The construction is genuinely
> useful (the proportionality constant captures over- or under-dispersion
> beyond the known sampling variance), but it is *not* the exact metafor
> fixed-V identity.

(Sketch only — a full drmTMB construction with known sigma offsets is
shown in `vignette("symbolizer-drmtmb")` and the drmTMB docs.)

## The variance / SD parameterization gap

Each package writes the scale model differently:

| Package | Quantity modelled | Link | Coefficient |
| --- | --- | --- | --- |
| `metafor::rma(scale = ~ z)` | $\tau^2$ (variance) | $\log(\tau^2_i) = \alpha_0 + \alpha_1 z_i$ | $\alpha$ |
| `brms::bf(sigma ~ z)` | $\sigma$ (SD) | $\log(\sigma_i) = \gamma_0 + \gamma_1 z_i$ | $\gamma$ |
| `glmmTMB(dispformula = ~ z)` | $\sigma$ (SD) | $\log(\sigma_i) = \gamma_0 + \gamma_1 z_i$ | $\gamma$ |
| `drmTMB(sigma ~ z)` | $\sigma$ (SD) | $\log(\sigma_i) = \gamma_0 + \gamma_1 z_i$ | $\gamma$ |

The two parameterizations are mathematically equivalent. Because
$\log(\tau^2) = 2 \log(\tau) + \mathrm{const}$, the slope on the
variance scale is twice the slope on the SD scale:

$$
\alpha_k \;\approx\; 2 \cdot \gamma_k
$$

(For the intercept the relationship is offset by $\log 2$ once.) So
a `metafor` location-scale slope of $\alpha = 0.4$ is the same
biological signal as a `brms` / `glmmTMB` / `drmTMB` location-scale
slope of $\gamma \approx 0.2$ — variance doubles ≈ SD increases by
$\sqrt{2}$.

`symbolizer` renders each faithfully in its own parameterization:
look for the coefficient symbol ($\alpha$ vs $\gamma$) and the
$\tau^2$-vs-$\sigma$ distinction in the LaTeX block to know which
convention is in play.

## When to use which

- **`metafor`** is the canonical interface for traditional
  meta-analysis. Rich diagnostics, well-documented effect-size
  computation (`escalc()`), and a long literature on small-sample
  corrections.
- **`glmmTMB` with `propto()`** is **not** a meta-analytic surface in
  the strict sense (propto estimates a scalar multiplier on the known
  covariance; meta-analysis fixes the covariance to V exactly). It
  *is* the right glmmTMB surface for **phylogenetic / pedigree /
  structured-covariance** models, where you know the *shape* of the
  covariance but want to estimate its size. The meta-analytic
  identity Σ = V (without a scalar multiplier) waits on glmmTMB
  shipping its planned `equalto()` block.
- **`drmTMB`** lets you fit location-scale meta-analysis with
  moderators on heterogeneity directly, with the same syntax as any
  other location-scale model. This is the surface Nakagawa et al.
  (2025) explore in depth for ecology and evolution.

`symbolizer` lets all three speak the same teachable language.

## References

- Viechtbauer, W., & López-López, J. A. (2022). Location-scale
  models for meta-analysis. *Research Synthesis Methods*, **13**(6),
  697–715.
- Nakagawa, S., Mizuno, A., Morrison, K., Ricolfi, L., Williams, C.,
  Drobniak, S. M., Lagisz, M., & Yang, Y. (2025). Location-scale
  meta-analysis and meta-regression as a tool to capture large-scale
  changes in biological and methodological heterogeneity: A
  spotlight on heteroscedasticity. *Global Change Biology*, **31**,
  e70204.
- Williams, C. R. (2023). *Fitting meta-analysis with the glmmTMB R
  package: a worked tutorial.*
  <https://coraliewilliams.github.io/equalto_sim_study/webpage.html>
