---
title: "Three flavors of meta-analysis: a cross-package tour"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Three flavors of meta-analysis: a cross-package tour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_pkgs <- requireNamespace("metafor", quietly = TRUE) &&
            requireNamespace("symbolizer", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = has_pkgs
)
```

## 1. Three flavors of meta-analysis

Meta-analysis pools $K$ study-level effect sizes $y_k$, each with a
**known** sampling variance $v_k$, into a single summary. The same idea
can be tightened in three different directions; this article walks all
three side-by-side across three packages, so a reader can land on
whichever combination matches their toolkit.

**Flavor 1 — Traditional random-effects pooling.** Studies vary
beyond what sampling alone explains. The headline parameter is
$\tau^2$, the **between-study heterogeneity variance**:

$$
y_k \mid \theta_k \sim \mathcal{N}(\theta_k, v_k),
\qquad
\theta_k \sim \mathcal{N}(\mu, \tau^2),
\qquad k = 1, \ldots, K.
$$

A single $\tau^2$ describes how much "true" effect sizes scatter across
studies after sampling noise is accounted for. This is the textbook
random-effects model (Borenstein et al. 2009; Viechtbauer 2010).

**Flavor 2 — Phylogenetic meta-analysis.** When studies are on related
species, the between-study tier becomes structured by phylogeny. The
total between-species variance splits into a **phylogenetic** piece
(constrained by relatedness) and a **non-phylogenetic** piece:

$$
\theta_k = \mu + u_{p,k} + u_{s,k},
\quad u_p \sim \mathcal{N}(\mathbf{0}, \sigma_p^2 \mathbf{A}),
\quad u_s \sim \mathcal{N}(\mathbf{0}, \sigma_s^2 \mathbf{I}).
$$

$\mathbf{A}$ is the phylogenetic correlation matrix (Cinar et al. 2022;
Mizuno et al. 2026).

**Flavor 3 — Location-scale meta-analysis.** When moderators predict
not just the *mean* effect but also the *heterogeneity*, the scale
parameter itself becomes a regression:

$$
y_k \sim \mathcal{N}(\mu_k, v_k + \tau_k^2),
\quad
\mu_k = \beta_0 + \beta_1 x_k,
\quad
\log \tau_k = \gamma_0 + \gamma_1 x_k.
$$

The biological question shifts: *which moderators predict how variable
study effects are?* — separate from the mean.

**Three packages, one math.** The same model can be fit in `metafor`,
`glmmTMB`, `drmTMB`, `brms`, or `MCMCglmm`. Each section below picks
one **deep Face** (the package whose interface most cleanly expresses
that flavor) and two **light Faces** (showing the cross-package
equivalence so a reader's existing fluency carries over).

**Takeaway.** Three flavors, one statistical motivation: separate the
sampling variance $v_k$ (known per study) from the heterogeneity
$\tau^2$ (estimated). Phylogenetic and location-scale flavors *structure*
that heterogeneity further.

## 2. The data: BCG vaccine efficacy trials

We use `dat.bcg` from `metafor` (Colditz et al. 1994): thirteen
randomised trials of the BCG vaccine against tuberculosis, with
treatment / control case counts per trial. The summary effect is the
**log risk ratio**:

$$
y_k = \log\!\left(\frac{a_k / (a_k + b_k)}{c_k / (c_k + d_k)}\right),
\qquad
v_k = \frac{1}{a_k} - \frac{1}{a_k + b_k} + \frac{1}{c_k} - \frac{1}{c_k + d_k}.
$$

```{r data}
library(metafor)
library(symbolizer)

data(dat.bcg, package = "metafor")
dat <- escalc(measure = "RR",
              ai = tpos, bi = tneg,
              ci = cpos, di = cneg,
              data = dat.bcg)
head(dat[, c("author", "year", "yi", "vi")], 6)
```

Thirteen trials; effect sizes range from strong protection ($y_k < 0$)
to no effect ($y_k \approx 0$). The sampling variances $v_k$ are
**known** (computed from the trial counts) and differ across trials —
large trials have small $v_k$, small trials have large $v_k$. *Any*
meta-analysis must respect that.

**Takeaway.** Same effect-size metric ($y_k$ = log RR), same known
sampling variance metric ($v_k$). One dataset reused across the three
flavors below.

## 3. Traditional pooling

The simplest flavor. Each trial gives $y_k$ with known $v_k$; we pool
to estimate a single mean $\mu$ and the between-study heterogeneity
$\tau^2$.

### 3.1 The model in symbols

$$
y_k \mid \theta_k \sim \mathcal{N}(\theta_k, v_k),
\qquad
\theta_k \sim \mathcal{N}(\mu, \tau^2),
\qquad k = 1, \ldots, 13.
$$

The marginal form (integrating out $\theta_k$) is the more commonly
fit one:

$$
y_k \sim \mathcal{N}(\mu, v_k + \tau^2),
\qquad
\mathrm{Cov}(y_k, y_{k'}) = 0 \text{ for } k \ne k'.
$$

**Two parameters of interest**: $\mu$ (the average log RR) and $\tau^2$
(how much trials disagree beyond sampling noise).

### 3.2 Face 1 — `metafor::rma()` (deep dive)

`metafor` was built for meta-analysis; `rma()` is the canonical
random-effects fit.

```{r metafor-fit}
fit_metafor <- rma(yi, vi, data = dat)
fit_metafor
```

`tau^2 = 0.31` says the true effects scatter across trials with
standard deviation $\hat\tau \approx 0.56$ on the log RR scale —
substantial heterogeneity. $I^2 = 92\%$ confirms: of total
variation, 92% is heterogeneity not sampling noise.

```{r metafor-symbolize}
sym_metafor <- symbolize(fit_metafor)
equations(sym_metafor)
```

```{r metafor-assumptions}
assumption_table(sym_metafor)
```

**Coefficient reading on μ.** $\hat\mu = -0.71$ on the log RR scale is
$\exp(-0.71) \approx 0.49$ on the response scale — vaccinated children
have about half the TB rate of unvaccinated children, averaged across
trials, with substantial trial-to-trial variation.

### 3.3 Face 2 — `glmmTMB` (light)

`glmmTMB` fits the same math via a random-intercept model with a
per-observation weight that pins the residual variance to $v_k$:

```{r glmmtmb-fit, message = FALSE, warning = FALSE}
if (requireNamespace("glmmTMB", quietly = TRUE)) {
  dat$study <- factor(seq_len(nrow(dat)))
  fit_glmmTMB <- glmmTMB::glmmTMB(
    yi ~ 1 + (1 | study),
    weights = 1 / vi,
    dispformula = ~ 0,        # pin residual SD to 1; weights carry v_k
    data = dat
  )
  cat("glmmTMB tau^2:",
      round(as.numeric(glmmTMB::VarCorr(fit_glmmTMB)$cond$study), 3),
      "(vs metafor:", round(fit_metafor$tau2, 3), ")\n")
} else {
  cat("glmmTMB not installed; skipping Face 2.\n")
}
```

The bridge: `dispformula = ~0` fixes the Gaussian dispersion at 1, and
`weights = 1/vi` makes each observation's effective residual variance
equal to $v_k$ rather than 1 — which reproduces the meta-analytic
likelihood. The estimated random-intercept variance is $\hat\tau^2$.

### 3.4 Face 3 — `drmTMB` (light)

`drmTMB` is symbolizer's distributional-regression workhorse. For
traditional meta-analysis without moderators it reduces to a weighted
Gaussian fit:

```{r drmtmb-fit, message = FALSE}
if (requireNamespace("drmTMB", quietly = TRUE)) {
  fit_drm <- drmTMB::drmTMB(
    drmTMB::drm_formula(yi ~ 1),
    family  = stats::gaussian(),
    weights = 1 / vi,
    data    = dat
  )
  cat("drmTMB residual sigma^2:",
      round(fit_drm$report$sigma_eps^2, 3),
      "\n")
} else {
  cat("drmTMB not installed; skipping Face 3.\n")
}
```

drmTMB without `sigma ~ x` or a random intercept does **not** separate
sampling from heterogeneity — the residual variance absorbs both. The
useful drmTMB Face is in §5 (location-scale), where the
`sigma ~ moderator + offset(0.5 * log(vi))` syntax models heterogeneity
explicitly as a function of moderators while still respecting the
known sampling variances $v_k$.

### 3.5 Cross-package summary

| Package | Call | Captures $\tau^2$ as |
|---|---|---|
| `metafor` | `rma(yi, vi, data = dat)` | `fit$tau2` directly. |
| `glmmTMB` | `glmmTMB(yi ~ 1 + (1\|study), weights = 1/vi, dispformula = ~0)` | `VarCorr(fit)$cond$study[1]`. |
| `drmTMB` | `drmTMB(drm_formula(yi ~ 1), weights = 1/vi)` | Residual $\sigma_\varepsilon^2$ (no separate $\tau^2$ without a random intercept). |
| `brms` | `brm(yi \| se(sqrt(vi)) ~ 1 + (1\|study))` | Random-intercept SD posterior. |
| `MCMCglmm` | `MCMCglmm(yi ~ 1, random = ~ study, mev = vi)` | `VCV[,"study"]` posterior. |

The math is one model; the syntax differs by package. The bridge
columns in `formula_bridge(sym_metafor)` summarise the rest.

**Takeaway.** Traditional pooling separates known sampling variance
$v_k$ from estimated heterogeneity $\tau^2$. metafor's `rma()` is the
cleanest interface; glmmTMB and drmTMB reproduce the math via
`weights = 1/vi`; brms and MCMCglmm use their own meta-analytic
bridges (`se(sqrt(vi))`, `mev = vi`).

## 4. Phylogenetic multilevel meta-analysis

When studies are on related species, the between-study tier splits
into a phylogenetic piece (constrained by the phylogeny via
$\mathbf{A}$) and a non-phylogenetic study tier.

### 4.1 The model in symbols

For each effect $kt$ in study $k$ on species $s(k)$:

$$
y_{kt} = \beta_0 + \beta_1\, x_{kt} + u_{p,\,s(k)} + u_{\text{study},\,k} + \epsilon_{kt}
$$

with

$$
u_{p,\,s} \sim \mathcal{N}(\mathbf{0},\, \sigma_p^2\, \mathbf{A}),
\quad
u_{\text{study},\,k} \sim \mathcal{N}(0,\, \sigma_{\text{study}}^2),
\quad
\epsilon_{kt} \sim \mathcal{N}(0,\, v_{kt}).
$$

The known sampling variance $v_{kt}$ is fixed per effect (computed from
the original study's sample size). The marginal variance of one effect
is

$$
\mathrm{Var}(y_{kt}) = v_{kt} + \sigma_{\text{study}}^2 + \sigma_p^2\, \mathbf{A}_{s(k),\,s(k)}.
$$

This is the Nakagawa-paper-style "phylogenetic multilevel meta-analysis"
form, with the location-scale extension deferred to §5.

### 4.2 Data

A 35-species, 164-effect subset of the Pottier et al. (2022) thermal
acclimation dataset (effect-size = `dARR`; moderator = `habitat`,
aquatic vs terrestrial). Subsample is shipped with the package so the
vignette is fully reproducible without network access.

```{r meta-phylo-data}
library(symbolizer)
library(ape)
dat <- read.csv(system.file("extdata", "thermal_subset.csv",
                            package = "symbolizer"),
                stringsAsFactors = FALSE)
tree <- ape::read.tree(system.file("extdata", "thermal_subset_tree.tre",
                                   package = "symbolizer"))
cat(sprintf("%d effects across %d species across %d studies; habitats: %s\n",
            nrow(dat),
            length(unique(dat$phylogeny)),
            length(unique(dat$study_ID)),
            paste(sort(unique(dat$habitat)), collapse = ", ")))
```

### 4.3 Face 1 — `drmTMB` (deep dive)

drmTMB's native idiom uses `meta_V()` *inside* the formula (it
replaces brms's response-side `se(sqrt(vi))`) and `phylo()` *inside*
the formula too (it replaces brms's `gr(., cov = A)`). One important
gotcha: drmTMB's formula parser does NOT accept namespace-qualified
names inside the formula, so we bind the helpers to local names
first.

```{r drmtmb-fit-phylo, message = FALSE, warning = FALSE}
library(drmTMB)

# drmTMB requires unqualified names INSIDE the formula -- bind locally.
bf      <- drmTMB::drm_formula
meta_V  <- drmTMB::meta_V
phylo   <- drmTMB::phylo

fit_drm_phylo <- drmTMB::drmTMB(
  bf(
    dARR ~ 1 + habitat +
            meta_V(V = Var_dARR) +
            phylo(1 | phylogeny, tree = tree) +
            (1 | study_ID),
    sigma ~ 1),
  family = stats::gaussian(),
  data   = dat
)
sym_drm_phylo <- symbolize(
  fit_drm_phylo,
  context = "phylogenetic multilevel meta-analysis"
)
```

#### Three views — phylogenetic multilevel

```{r widget-phylo, results = 'asis', echo = FALSE, message = FALSE}
as_html_three_views(sym_drm_phylo, id = "phylo-multilevel")
```

The widget's **Tab 3** decomposes the marginal variance into the known
sampling tier $v_{kt}$ and the two estimated tiers
($\sigma_{\text{study}}^2$, $\sigma_p^2\,\mathbf{A}_{kk}$).

### 4.4 Face 2 — `brms` (light, with `se(sqrt(vi))` + `gr(., cov = A)`)

The same math in brms's idiom. Note that `se()` is parsed by brms
internally — you write it *unqualified* inside `bf()`, not as
`brms::se()`.

```{r brms-fit-phylo, eval = FALSE}
library(brms)
A_phylo <- ape::vcv.phylo(tree, corr = TRUE)
fit_brms_phylo <- brm(
  bf(dARR | se(sqrt(Var_dARR)) ~ 1 + habitat
       + (1 | study_ID)
       + (1 | gr(phylogeny, cov = mat))),
  data   = dat,
  data2  = list(mat = A_phylo),
  family = gaussian(),
  chains = 2, iter = 1000, warmup = 500,
  file = "cache/brms-phylo-multilevel"   # cache to disk
)
```

The cross-package bridge (cf. §7): brms `se(sqrt(vi))` ↔ drmTMB
`meta_V(V = vi)`; brms `gr(g, cov = A)` ↔ drmTMB
`phylo(1 | g, tree = ...)`.

### 4.5 Face 3 — `metafor::rma.mv` (light)

`metafor` was built for meta-analysis; `rma.mv()` accepts `V = vi` for
the known sampling variances and `R = list(... = A)` for the
phylogenetic correlation:

```{r metafor-fit-phylo}
library(metafor)
A_phylo <- ape::vcv.phylo(tree, corr = TRUE)
fit_metafor_phylo <- rma.mv(
  yi     = dARR,
  V      = Var_dARR,
  mods   = ~ 1 + habitat,
  random = list(~ 1 | phylogeny, ~ 1 | study_ID),
  R      = list(phylogeny = A_phylo),
  data   = dat
)
fit_metafor_phylo$sigma2
```

The two `$sigma2` rows correspond to the two random-effect tiers in
the math above.

**Takeaway.** Three packages, one math. drmTMB leads with the
cleanest meta-analytic idiom (`meta_V` + `phylo` inside the formula);
brms and metafor offer equivalent fits via package-specific syntax.
The next step (§5) adds the location-scale extension — modelling
$\tau^2$ itself as a function of the moderator.

## 5. Location-scale meta-analysis

*(Scaffold; full coverage lands in v0.22.2.)*

When moderators predict heterogeneity, fit the variance side:

$$
y_k \sim \mathcal{N}(\mu(x_k), v_k + \tau^2(x_k)),
\quad
\log \tau(x_k) = \gamma_0 + \gamma_1 x_k.
$$

The deep Face is
`drmTMB::drmTMB(drm_formula(yi ~ x, sigma ~ x + offset(0.5*log(vi))),
                family = gaussian())`.
The `offset(0.5*log(vi))` term makes the σ-submodel additive on the
known sampling variances — `glmmTMB::glmmTMB(..., dispformula = ~ x,
weights = 1/vi)` is an equivalent Face.

## 6. Reading biologically

*(Will land in v0.22.3, after all three flavors' deep widgets ship.)*

Three readings the article will tie together:

- **Heterogeneity τ² vs sampling v_k**: τ² is "real" disagreement
  among studies; v_k is just imprecision.
- **Phylogenetic structure σ_p²A**: when most heterogeneity sits on
  the phylogeny, related species' effects co-vary even after $v_k$ is
  accounted for. Read as: "the trait has a phylogenetic signal in its
  effect size."
- **Moderator-driven τ²(x)**: scale-regression coefficients γ are
  read on the log-SD scale, exponentiated to multiplicative scale.

## 7. References

- Borenstein, M., Hedges, L. V., Higgins, J. P. T., & Rothstein, H. R.
  (2009). *Introduction to Meta-Analysis*. Wiley.
- Cinar, O., Nakagawa, S., & Viechtbauer, W. (2022). Phylogenetic
  multilevel meta-analysis: a simulation study on the importance of
  modelling the phylogeny. *Methods in Ecology and Evolution* 13(2):
  383-395.
- Colditz, G. A., et al. (1994). Efficacy of BCG vaccine in the
  prevention of tuberculosis: meta-analysis of the published
  literature. *JAMA* 271: 698-702.
- Mizuno, A., et al. (2026). Unified phylogenetic + spatial multilevel
  meta-analysis. *Research Synthesis Methods*, in press.
- Nakagawa, S., & Santos, E. S. A. (2012). Methodological issues and
  advances in biological meta-analysis. *Evolutionary Ecology* 26:
  1253-1274.
- Viechtbauer, W. (2010). Conducting meta-analyses in R with the
  metafor package. *Journal of Statistical Software* 36(3): 1-48.
